---
title: "Making it Rain"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%",
                      fig.height = 4,
                      fig.width = 6)
```

## Summary

This page provides a quick tutorial on how to make a raincloud plot. The code is inspired by a blog post by [Cedric Scherer](https://www.cedricscherer.com/2021/06/06/visualizing-distributions-with-raincloud-plots-and-how-to-create-them-with-ggplot2/).

## Getting started

Okay, start by opening packages:

```{r}
library(tidyverse)
library(coolorrr)
set_theme()
set_palette()
```

Next, we'll create a function to produce a "raincloud plot":

```{r}
raincloud <- function(df, x, y, 
                      fill = NULL, 
                      alpha = 1,
                      flip = FALSE) {
  x <- enquo(x)
  y <- enquo(y)
  fill <- enquo(fill)
  
  ggplot(df) +
    aes(x = !! x, y = !! y, fill = !! fill) +
  ## add half-violin from {ggdist} package
  ggdist::stat_halfeye(
    ## customize thickness
    aes(thickness = stat(pdf*n)),
    ## custom bandwidth
    adjust = .5, 
    ## adjust height
    width = .6, 
    ## move geom to the right
    justification = -.2, 
    ## set alpha
    alpha = alpha,
    ## remove slab interval
    .width = 0, 
    point_colour = NA
  ) + 
  geom_boxplot(
    width = .12, 
    ## remove outliers
    outlier.color = NA ## `outlier.shape = NA` works as well
  ) +
  ## add dot plots from {ggdist} package
  ggdist::stat_dots(
    ## orientation to the left
    side = "left", 
    ## move geom to the left
    justification = 1.1, 
    ## adjust grouping (binning) of observations 
    binwidth = .25,
    linewidth = 0
  ) -> plot ## save new plot object
  
  ## option to flip the coordinates
  if(flip) { ## if flip = TRUE
    plot + 
      coord_flip()
  } else {   ## if flip = FALSE
    plot
  }
}
```


## How it works

The function can be used a few different ways. I'll illustrate using the `mtcars` dataset available in R.

First, I want car cylinders to be a factor:

```{r}
mtcars <- mtcars |>
  mutate(cyl = as.factor(cyl))
```

Next, I'll create a raincloud plot using cylinders as my grouping factor and miles per gallon or mpg as the variable to be summarized. 
```{r}
raincloud(mtcars, x = cyl, y = mpg)
```

If you want to flip the coordinates so the output looks like true rainclouds, just set `flip = TRUE`:

```{r}
raincloud(mtcars, cyl, mpg, flip = T)
```

If you want to update labels or other settings, you can do this like with any other ggplot() object. Just remember that when you set `flip = T` that the x and y axis labels will be reversed.

```{r}
raincloud(mtcars, cyl, mpg, flip = T) +
  labs(
    y = "Miles per Gallong",
    x = NULL,
    title = "MPG by number of cylinders"
  )
```

## Adding color

Sometimes adding color can be a nice touch, even if it's a tad redundant. It's easy to do this and customize using `ggpal()` from `{coolorrr}`. All you need to do is set `fill` in the raincloud function.

```{r}
raincloud(mtcars, cyl, mpg, fill = cyl, flip = T) +
  labs(
    x = NULL,
    y = "Miles per Gallon",
    fill = "Cylinders"
  ) +
  ggpal(aes = "fill")
```

If you want to use color more substantively, you can do that too. Just map the fill aesthetic to a different categorical variable. When you do this, it helps to increase the transparency of the density "cloud" since these will overlap in the figure. Just use the `alpha` command in the raincloud function.

```{r}
## make the 'am' column categorical
mtcars |>
  mutate(
    am = ifelse(am == 0, "Auto", "Manual")
  ) -> mtcars

## plot
raincloud(mtcars, cyl, mpg, am, 0.5, flip = T) +
  ggpal(type = "binary", aes = "fill")
```

