---
output: 
  stevetemplates::beamer:
    keep_tex: TRUE
    latex_engine: pdflatex # use xelatex here instead! I recommend it, but this is minimal reprex
    dev: cairo_pdf # I typically comment this out  if latex_engine: pdflatex
    slide_level: 3 # I prefer this, but I won't force it on you.
title: Cross-sectional Data
subtitle: DPR 190
author: Miles D. Williams
institute: Data for Political Research
# titlegraphic: "`r paste0(Sys.getenv('HOME'), '/Dropbox/clemson/watermarks/clemson-university-wordmark-orange-purple.png')`"
# scaletg: .9 # Optional, proportional (0, 1) to paper width. Defaults to .3
fontsize: 12pt
make149: TRUE
# mainfont: "Open Sans" # Try out some font options if xelatex
# titlefont: "Titillium Web" # Try out some font options if xelatex
# specify color as six-digit RGB (no pound sign)
# primarycolor: "990000" 
# secondarycolor: "000099"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(
  echo = F,
  warning = F,
  message = F
)
library(tidyverse)
library(stevemisc)
library(kableExtra)

## some data
path <- "https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20101/Data/MIE_2000-2014.csv"
mie_data <- read_csv(path)
```


# Cross-sectional data

### Cross-sectional data

\begin{centering}

What is it?

\end{centering}

### Cross-sectional data

**A dataset:**

```{r}
mie_data |>
  filter(fatalmin > 0) |>
  sample_n(5) 
```

### Cross-sectional data

*Tidy* cross-sectional data:

- Each *row* is a unique unit of observation
- Each *column* is a variable for a unique unit
- Each *cell* contains only one value

### Cross-sectional data

**Not a cross-sectional dataset**:

```{r}
gapminder::gapminder |>
  slice_head(n = 5) 
```

### Cross-sectional data

\begin{centering}
If a unit makes \textbf{makes more than one appearance} in the data, the data is either not tidy or not cross-sectional.
\end{centering}

# Uses of cross-sectional data

### Uses of cross-sectional data

- What's a benefit of cross-sectional data?
- What's a limitation?


### Uses of cross-sectional data

- \textcolor{purple}{Efficient, snap-shot, case comparisons, generate hypotheses}
- What's a limitation?


### Uses of cross-sectional data

- \textcolor{purple}{Efficient, snap-shot, case comparisons, generate hypotheses}
- \textcolor{purple}{Apples-to-oranges, selection bias}
