---
title: "Garlic Mustard I"
author: "Miles D. Williams"
date: '2023-01-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Importing and manipulating data

```{r}
## Install dplyr and ggplot2 if needed
# install.packages(c("dplyr", "ggplot2"))

## Load them into the current session:
library(dplyr) 
library(ggplot2)

```


```{r}
gmdata <- read.csv("GarlicMustardData.csv")
```

We can look at some details with the summary function:

```{r}
summary(gmdata)
```

We can manipulate the data. Let's create a vector of just 3 variables:

```{r}
myvars <- c("Latitude", "Longitude", "Altitude")
```

And now let's ask R to make a new dataset that only includes these variables.

```{r}
gmdata_geo <- gmdata[myvars]
summary(gmdata_geo) # check it out
```

We could have done the above using the select function:

```{r}
gmdata_geo <- gmdata %>%
  select(Latitude, Longitude, Altitude)
summary(gmdata_geo)
```

We can use select to get a subset of both adjoining and non-adjoining columns.

```{r}
gmdata_subset <- gmdata %>%
  select(1, 5:8)
summary(gmdata_subset)
```

We can also subset the data by filtering. Let's filter to observations with a total density of 4 or greater.

```{r}
gmdata_filter <- gmdata %>%
  filter(TotalDens >= 4 & Altitude >= 100)
summary(gmdata_filter)
```


## 2. Making plots with ggplot



```{r}

```

