---
title: "Playing with DU fall 2023 survey data"
author: "Miles D. Williams"
date: "2023-10-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F)
library(tidyverse)
library(socsci)
library(coolorrr)
set_theme() 
set_palette(
  # qualitative = "https://coolors.co/palette/588b8b-ffffff-ffd5c2-f28f3b-c8553d-2d3047-93b7be",
  binary = "https://coolors.co/palette/0a2463-fb3640"
)

## the survey data:
library(googlesheets4)
gs4_deauth()
o23 <- range_read(
  "https://docs.google.com/spreadsheets/d/19Et52eKP1Yte9tainpfB7f_9sCB8UAVM36DbR2ySyaI/edit#gid=1526346509"
)

## recodes from Paul

o23 <- o23 %>% rename_all(tolower)

#Some Recodes####
#Race
o23 <- o23 %>% mutate(white=car::recode(q57_1, "NA=0"),
                      hispanic=car::recode(q57_2, "NA=0"),
                      black=car::recode(q57_3, "NA=0"),
                      asian=car::recode(q57_4, "NA=0"),
                      other=car::recode(q57_5, "NA=0"))


o23 <- o23 %>% mutate(race=case_when(white==1 & (black!=1 & hispanic!=1 & asian!=1 & other!=1) ~ 1,
                                     black==1 ~ 2,
                                     asian==1 & (black!=1 & hispanic!=1) ~ 4,
                                     hispanic==1 ~ 3,
                                     other==1 & (black!=1 & hispanic!=1 & asian!=1)~ 5))
o23 <- o23 %>% mutate(race=frcode(race=="1" ~ "White",
                                  race=="2" ~ "Black",
                                  race=="3" ~ "Hispanic",
                                  race=="4" ~ "Asian",
                                  race=="5" ~ "Other"))

o23 <- o23 %>% mutate(female=frcode(q53==1 ~ "Men",
                                    q53==2 ~ "Women"))
o23 <- o23 %>% mutate(femtran=frcode(q54_1==1 ~ "Men",
                                     q54_2==1 ~ "Women",
                                     q54_3==1 | q54_4==1 ~ "Trans/N-B"))

o23 <- o23 %>% mutate(greek=car::recode(q44_6, "NA=0"))
o23 <- o23 %>% mutate(varsity=car::recode(q44_1, "NA=0"))

o23 <- o23 %>% mutate_at(., vars(starts_with('q44_')), funs(replace_na(., 0)))
o23 <- o23 %>% mutate(groups=rowSums(dplyr::select(., starts_with("q44_")), na.rm = TRUE))


o23 <- o23 %>% mutate(pid7=car::recode(q3, "8=4"))
o23 <- o23 %>% mutate(pidst=car::recode(pid7, "1=4; 2=3;3=2;4=1;5=2;6=3;7=4"))
o23 <- o23 %>% mutate(pidstf=frcode(pidst==1 ~ "Independents",
                                    pidst==2 ~ "Leaning Partisan",
                                    pidst==3 ~ "Weak Partisans",
                                    pidst==4 ~ "Strong Partisans"))
o23 <- o23 %>% mutate(pid3=car::recode(pid7, "1:3=1; 4=2;5:7=3"),
                      pid3f=frcode(pid3==1 ~ "Democrat",
                                   pid3==2 ~ "Independent",
                                   pid3==3 ~ "Republican"))
o23 <- o23 %>% mutate(pid7f=frcode(pid7==1 ~ "Strong\nDemocrat", 
                                   pid7==2 ~ "Democrat", 
                                   pid7==3 ~ "Leaning\nDemocrat",
                                   pid7==4 ~ "Independent", 
                                   pid7==5 ~ "Leaning\nRepublican", 
                                   pid7==6 ~ "Republican",
                                   pid7==7 ~ "Strong\nRepublican"))
#Partisan polarization
o23 <- o23 %>% mutate(polar=abs(q11_6-q11_8))  
o23 <- o23 %>% mutate(polint=6-q2)

o23 <- o23 %>% mutate(classyr=frcode(q63==4 ~ "Freshman",
                                     q63==3 ~ "Sophomore",
                                     q63==2 ~ "Junior",
                                     q63==1 ~ "Senior"))

o23 <- o23 %>% mutate(hetero=car::recode(q55, "1=1; 2:6=0"))


#Intense discussion experiment
o23 <- o23 %>% gather(key="disctype", value="discamt", q15, q18, na.rm=T)
o23 <- o23 %>% gather(key="disctype", value="discmore", q16, q19, na.rm=T)
o23 <- o23 %>% gather(key="disctype", value="lac_disc", q17_1, q20_1, na.rm=T)
o23 <- o23 %>% gather(key="disctype", value="lac_lead", q17_2, q20_2, na.rm=T)
o23 <- o23 %>% gather(key="disctype", value="lac_div", q17_3, q20_3, na.rm=T)
o23 <- o23 %>% gather(key="disctype", value="lac_zone", q17_4, q20_4, na.rm=T)

o23 <- o23 %>% mutate(disctype=frcode(disctype=="q17_4" ~ "Non-political",
                                      disctype=="q20_4" ~ "Political."))


## Adding mine about foreign policy Qs:

# pick 3 FP issues that are most important (Q8)
o23 <- o23 |>
  mutate(
    across(q8_1:q8_12, ~ replace_na(.x, 0))
  ) |>
  rename(
    cyber_security = q8_1,
    ethnic_conflict = q8_2,
    climate_change = q8_3,
    human_rights = q8_4,
    migration = q8_5,
    terrorism = q8_6,
    trade = q8_7,
    regional_disintegration = q8_8,
    russian_aggression = q8_9,
    rising_china = q8_10,
    domestic_political_instability = q8_11,
    weapons_of_mass_destruction = q8_12,
    other_fp_concerns = q8_13
  ) 

# how like are the following conflicts?
o23 <- o23 |>
  mutate(
    war_us_china = q9_1,   # US vs China
    war_us_russia = q9_2,  # US vs Russia
    war_china_taiwan = q9_3# China invades Taiwan
  )

# Biden's FP performance
o23 <- o23 |>
  mutate(
    biden_fp = frcode(
      q10_1 == 5 ~ "Very poorly",
      q10_1 == 4 ~ "Poorly",
      q10_1 == 3 ~ "Neither",
      q10_1 == 2 ~ "Well",
      q10_1 == 1 ~ "Very well"
    )
  )
```



Okay let's do some analysis.

What FP issues do DU students think are important?

```{r}
fp_experts <- tibble(
  name = str_replace_all(
    names(
      o23 |> select(cyber_security:weapons_of_mass_destruction)
    ), 
    "_",
    " "
  ) |> str_to_title(),
  fpvalue = c(
    19.03, 2.14, 67.83, 6.03,
    11.93, 2.28, 7.91, 3.89, 46.78,
    63, 52.14, 9.79
  ) / 100
)
o23 |>
  summarize(
    across(cyber_security:weapons_of_mass_destruction, ~ mean(.x, na.rm = T))
  ) |>
  pivot_longer(
    everything()
  ) |>
  mutate(
    name = str_replace_all(name, "_", " ") |>
      str_to_title()
  ) |>
  left_join(fp_experts) -> plt_dt
ggplot(plt_dt) +
  aes(
    x = value,
    y = reorder(name, value)
  ) +
  geom_pointrange(
    aes(xmin = fpvalue, xmax = value),
    show.legend = F,
    color = "steelblue"
  ) +
  geom_point(
    aes(x = fpvalue),
    color = "darkred",
    size = 2,
    shape = 15
  ) +
  labs(
    x = "Share of Students",
    y = NULL,
    title = "What foreign policy issues are Denison\nstudents concerned about?",
    subtitle = "Student concerns vs. IR faculty concerns from TRIP Survey"
  ) +
  scale_x_continuous(
    labels = scales::percent
  ) +
  ggpal(
    type = "sequential"
  ) +
  theme(
    panel.grid.major.y = element_blank()
  ) +
  geom_text(
    data = plt_dt |>
      filter(name == "Human Rights"),
    aes(label = "Students"),
    color = "steelblue",
    vjust = -0.75
  ) +
  geom_text(
    data = plt_dt |>
      filter(name == "Human Rights"),
    aes(label = "IR Faculty", x = fpvalue),
    color = "darkred",
    vjust = -0.75
  )
```

Denison students and IR faculty evaluate the importance of certain issues at nearly similar rates, such as climate change, trade, and cybersecurity. But there are chasmic divides on issues such as human rights, a rising China, and a revanchist Russia. IR faculty are deeply concerned about China and Russia while the average Denison student is not. Conversely, the average Denison student is highly concerned about human rights in US foreign policy, but the average IR faculty thinks human rights is of lesser concern.

How do students evaluate the likelihood of different conflicts in the next 10 years?

I'll make a rain cloud plot:

```{r}
raincloud <- function(df, x, y, 
                      fill = NULL, 
                      alpha = 1,
                      flip = FALSE) {
  x <- enquo(x)
  y <- enquo(y)
  fill <- enquo(fill)
  
  ggplot(df) +
    aes(x = !! x, y = !! y, fill = !! fill) +
  ## add half-violin from {ggdist} package
  ggdist::stat_halfeye(
    ## customize thickness
    aes(thickness = stat(pdf*n)),
    ## custom bandwidth
    adjust = .5, 
    ## adjust height
    width = .6, 
    ## move geom to the right
    justification = -.2, 
    ## set alpha
    alpha = alpha,
    ## remove slab interval
    .width = 0, 
    point_colour = NA
  ) + 
  geom_boxplot(
    width = .12, 
    ## remove outliers
    outlier.color = NA ## `outlier.shape = NA` works as well
  ) +
  ## add dot plots from {ggdist} package
  ggdist::stat_dots(
    ## orientation to the left
    side = "left", 
    ## move geom to the left
    justification = 1.1, 
    ## adjust grouping (binning) of observations 
    binwidth = .25,
    linewidth = 0
  ) -> plot ## save new plot object
  
  ## option to flip the coordinates
  if(flip) { ## if flip = TRUE
    plot + 
      coord_flip()
  } else {   ## if flip = FALSE
    plot
  }
}
```


```{r}
library(ggridges)
o23 |>
  pivot_longer(
    starts_with("war")
  ) |>
  mutate(
    name = frcode(
      name == "war_us_china" ~ "War between \nUS vs. China",
      name == "war_us_russia" ~ "War between\nUS vs. Russia",
      name == "war_china_taiwan" ~ "China invades\nTaiwan"
    )
  ) |>
  raincloud(
    x = name,
    y = value,
    flip = T,
    fill = name
  ) +
  ggpal(aes = "fill") +
  labs(
    y = "% Likelihood",
    x = NULL,
    title = "How likely do Denison students think the\nfollowing conflicts are in the next 10 years?"
  ) +
  theme(
    legend.position = "none"
  )
```

Do Denison students think Biden is doing well in foreign policy?

```{r}
fp_experts <- tibble(
  biden_fp = c("Very well", "Well", "Neither", "Poorly", "Very poorly"),
  pct = c(23.74, 49.93, 15.83, 6.14, 4.37) / 100
)
library(ggtext)
sbtext <- paste0(
  "Denison ", "<span style = 'color:steelblue'>***students***</span>",
  " compared to ", "<span style = 'color:darkred'>***IR faculty***</span>",
  " responses from the TRIP Survey"
)
o23 |>
  ct(biden_fp, show_na = F) |>
  ggplot() +
  aes(
    x = biden_fp,
    y = pct
  ) +
  geom_col(
    width = 0.5,
    fill = "steelblue"
  ) +
  geom_pointrange(
    data = fp_experts,
    aes(ymin = 0, ymax = pct),
    shape = 15,
    size = 3,
    color = "darkred"
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = NULL,
    y = "Share of Students",
    title = "How do Denison students think Biden has\nperformed in foreign policy the last 3 years?",
    subtitle = sbtext
  ) +
  theme(
    panel.grid.major.x = element_blank(),
    plot.subtitle = element_markdown()
  )
```

Who thinks Biden has done better or worse?

```{r}
o23 |>
  group_by(pid3f) |>
  ct(biden_fp, show_na = F) |>
  drop_na() |>
  mutate(
    value = as.numeric(biden_fp)
  ) |>
  ggplot() +
  aes(
    x = value,
    y = pct,
    color = pid3f
  ) +
  geom_point(size = 3) +
  geom_line(size = 1) +
  scale_x_continuous(
    labels = unique(o23$biden_fp) |> na.omit() |> sort()
  ) +
  ggpal() +
  theme(
    panel.grid.major.x = element_blank()
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = NULL,
    y = "Share of Students",
    title = "How do student evaluations of Biden\nbreak down by partisanship?",
    color = NULL
  )
```

