---
title: "Untitled"
author: "Miles D. Williams"
date: "2023-11-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = F,
                      dpi = 800,
                      out.width = "80%")

## packages
library(tidyverse)
library(coolorrr)
library(socsci)
library(googlesheets4)
library(showtext)

## custom data viz
set_theme()   # you can get rid of this if you don't like it
set_palette(
  qualitative = c("steelblue", "gray20", "red3", "purple4", "forestgreen"),
  from_coolors = F
)
# font_add_google("EB Garamond")
# showtext_auto()

## survey data
gs4_deauth()
o23 <- range_read(
  "https://docs.google.com/spreadsheets/d/19Et52eKP1Yte9tainpfB7f_9sCB8UAVM36DbR2ySyaI/edit#gid=1526346509"
)

## link to the code book:
# https://drive.google.com/file/d/13HQ7HgKD4JS6CMYGUIUwgUykGqXttSV9/view?usp=sharing

## Some  Gifted Recodes 
o23 |>
  
    # make all column names lowercase
  rename_with(tolower, everything()) |>
    
    # start the recodes
  mutate(
    
    # race
    white = replace_na(q57_1, 0),
    hispanic = replace_na(q57_2, 0),
    black = replace_na(q57_3, 0),
    asian = replace_na(q57_4, 0),
    other = replace_na(q57_5, 0),
    race = case_when(
      white==1 & (black!=1 & hispanic!=1 & asian!=1 & other!=1) ~ "White",
      black==1 ~ "Black",
      asian==1 & (black!=1 & hispanic!=1) ~ "Asian",
      hispanic==1 ~ "Hispanic",
      other==1 & (black!=1 & hispanic!=1 & asian!=1)~ "Other"
    ),
    
    # gender
    gender = frcode(
      q53 == 1 ~ "Men",
      q53 == 2 ~ "Women"
    ),
    femtran = frcode(
      q54_1 == 1 ~ "Men",
      q54_2 == 1 ~ "Women",
      q54_3 == 1 | q54_1 == 1 ~ "Trans/N-B"
    ),
    
    # group participation
    greek = replace_na(q44_6, 0),
    varsity = replace_na(q44_1, 0),
    across(
      starts_with("q44_"),
      ~ replace_na(.x, 0)
    )
  ) |>
  rowwise() |>
  mutate(
    num_groups = 
      sum(c_across(starts_with("q44_")), na.rm  = T)
  ) |>
  ungroup() |>
  mutate(
    
    # partisanship
    pid7 = frcode(
      q3 == 1 ~ "Strong Democrat",
      q3 == 2 ~ "Democrat",
      q3 == 3 ~ "Lean Democrat",
      q3 == 4 ~ "Independent",
      q3 == 5 ~ "Lean Republican",
      q3 == 6 ~ "Republican",
      q3 == 7 ~ "Strong Republican"
    ),
    pid3 = frcode(
      q3 %in% 1:3 ~ "Democrat",
      q3 == 4 ~ "Independent",
      q3 %in% 5:7 ~ "Republican"
    ),
    
    # polarization
    polarized = abs(q11_6 - q11_8)
  ) |>
  
    # importance of foreign policy issues
  mutate(
    across(q8_1:q8_12, ~ replace_na(.x, 0))
  ) |>
  rename(
    cyber_security = q8_1,
    ethnic_conflict = q8_2,
    climate_change = q8_3,
    human_rights = q8_4,
    migration = q8_5,
    terrorism = q8_6,
    trade = q8_7,
    regional_disintegration = q8_8,
    russian_aggression = q8_9,
    rising_china = q8_10,
    domestic_political_instability = q8_11,
    weapons_of_mass_destruction = q8_12,
    other_fp_concerns = q8_13
  ) |>
  mutate(
    # likelihood of war...
    war_us_china = q9_1,    # US vs China
    war_us_russia = q9_2,   # US vs Russia
    war_china_taiwan = q9_3,# China invades Taiwan
    
    # biden's foreign policy performance
    biden_fp = frcode(
      q10_1 == 5 ~ "Very poorly",
      q10_1 == 4 ~ "Poorly",
      q10_1 == 3 ~ "Neither",
      q10_1 == 2 ~ "Well",
      q10_1 == 1 ~ "Very well"
    ),
    
    # for reference, biden's domestic policy performance
    biden_dp = frcode(
      q10_2 == 5 ~ "Very poorly",
      q10_2 == 4 ~ "Poorly",
      q10_2 == 3 ~ "Neither",
      q10_2 == 2 ~ "Well",
      q10_2 == 1 ~ "Very well"
    )
  ) -> o23

```

## Analysis

### Concern about foreign policy issues

```{r}
fp_experts <- tibble(
  name = str_replace_all(
    names(
      o23 |> select(cyber_security:weapons_of_mass_destruction)
    ), 
    "_",
    " "
  ) |> str_to_title(),
  fpvalue = c(
    19.03, 2.14, 67.83, 6.03,
    11.93, 2.28, 7.91, 3.89, 46.78,
    63, 52.14, 9.79
  ) / 100
)
library(ggtext)
sbtext <- paste0(
  "<span style = 'color:steelblue'>***Students***</span>",
  " versus ", "<span style = 'color:darkred'>***IR professors***</span>"
)
o23 |>
  summarize(
    across(cyber_security:weapons_of_mass_destruction, ~ mean(.x, na.rm = T))
  ) |>
  pivot_longer(
    everything()
  ) |>
  mutate(
    name = str_replace_all(name, "_", " ") |>
      str_to_title()
  ) |>
  left_join(fp_experts) -> plt_dt
ggplot(plt_dt) +
  aes(
    x = value,
    y = reorder(name, value)
  ) +
  geom_pointrange(
    aes(xmin = fpvalue, xmax = value),
    show.legend = F,
    color = "gray"
  ) +
  geom_point(
    color = "steelblue",
    size = 2.5
  ) +
  geom_point(
    aes(x = fpvalue),
    color = "darkred",
    size = 2,
    shape = 15
  ) +
  labs(
    x = "Share of Responses",
    y = NULL,
    title = "What foreign policy issues are Denison\nstudents concerned about?",
    subtitle = sbtext
  ) +
  scale_x_continuous(
    labels = scales::percent
  ) +
  ggpal(
    type = "sequential"
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.subtitle = element_markdown()
  ) 
ggsave(
  here::here(
    "Misc",
    "DU_survey_f2023",
    "figs",
    "01_fig.png"
  )
)
```

Denison students and IR faculty evaluate the importance of certain issues at nearly similar rates, such as climate change, trade, and cybersecurity. But there are chasmic divides on issues such as human rights, a rising China, and a revanchist Russia. IR faculty are deeply concerned about China and Russia while the average Denison student is not. Conversely, the average Denison student is highly concerned about human rights in US foreign policy, but the average IR faculty thinks human rights is of lesser concern.

How do students evaluate the likelihood of different conflicts in the next 10 years?

I'll make a rain cloud plot:

```{r}
raincloud <- function(df, x, y, 
                      fill = NULL, 
                      alpha = 1,
                      flip = FALSE) {
  x <- enquo(x)
  y <- enquo(y)
  fill <- enquo(fill)
  
  ggplot(df) +
    aes(x = !! x, y = !! y, fill = !! fill) +
  ## add half-violin from {ggdist} package
  ggdist::stat_halfeye(
    ## customize thickness
    aes(thickness = stat(pdf*n)),
    ## custom bandwidth
    adjust = .5, 
    ## adjust height
    width = .6, 
    ## move geom to the right
    justification = -.2, 
    ## set alpha
    alpha = alpha,
    ## remove slab interval
    .width = 0, 
    point_colour = NA
  ) + 
  geom_boxplot(
    width = .12, 
    ## remove outliers,
    color = "gray90",
    outlier.color = NA ## `outlier.shape = NA` works as well
  ) +
  ## add dot plots from {ggdist} package
  ggdist::stat_dots(
    ## orientation to the left
    side = "left", 
    ## move geom to the left
    justification = 1.1, 
    ## adjust grouping (binning) of observations 
    binwidth = .25,
    linewidth = 0
  ) -> plot ## save new plot object
  
  ## option to flip the coordinates
  if(flip) { ## if flip = TRUE
    plot + 
      coord_flip()
  } else {   ## if flip = FALSE
    plot
  }
}
```


```{r}
library(ggridges)
fp_tibble <- tibble(
  name = c("War between \nUS & China",
           "War between\nUS & Russia",
           "China invades\nTaiwan"),
  value = c(24.5, 34.5, 14.5),
  label = c("20-29", "30-39", "10-19")
)
o23 |>
  pivot_longer(
    starts_with("war")
  ) |>
  mutate(
    name = frcode(
      name == "war_us_china" ~ "War between \nUS & China",
      name == "war_us_russia" ~ "War between\nUS & Russia",
      name == "war_china_taiwan" ~ "China invades\nTaiwan"
    )
  ) |>
  raincloud(
    x = name,
    y = value,
    flip = T,
    fill = name
  ) +
  geom_point(
    data = fp_tibble,
    aes(x = name, y = value),
    shape = 23,
    fill = "gold",
    size = 3
  ) +
  ggpal(aes = "fill") +
  labs(
    y = "% Likelihood",
    x = NULL,
    title = "How likely do Denison students think the\nfollowing conflicts are in the next 10 years?"
  ) +
  theme(
    legend.position = "none"
  ) +
  annotate(
    "point",
    y = 75,
    x = 0.5,
    pch = 23,
    fill = "gold",
    size = 3
  ) +
  annotate(
    "text",
    y = 75,
    x = 0.5,
    label = " = Median IR Faculty",
    hjust = -0.05
  )
ggsave(
  here::here(
    "Misc",
    "DU_survey_f2023",
    "figs",
    "02_fig.png"
  )
)
```

Do Denison students think Biden is doing well in foreign policy?

```{r}
fp_experts <- tibble(
  biden_fp = c("Very well", "Well", "Neither", "Poorly", "Very poorly"),
  pct = c(23.74, 49.93, 15.83, 6.14, 4.37) / 100
)
library(ggtext)
sbtext <- paste0(
  "<span style = 'color:steelblue'>***Students***</span>",
  " versus ", "<span style = 'color:darkred'>***IR professors***</span>"
)
o23 |>
  ct(biden_fp, show_na = F) |>
  ggplot() +
  aes(
    x = biden_fp,
    y = pct
  ) +
  geom_col(
    width = 0.5,
    fill = "steelblue"
  ) +
  geom_pointrange(
    data = fp_experts,
    aes(ymin = 0, ymax = pct),
    shape = "*",
    size = 3,
    linewidth = 1,
    color = "darkred"
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = NULL,
    y = "Share of Responses",
    title = "How do Denison students think Biden has\nperformed in foreign policy the last 3 years?",
    subtitle = sbtext
  ) +
  theme(
    panel.grid.major.x = element_blank(),
    plot.subtitle = element_markdown()
  )
ggsave(
  here::here(
    "Misc",
    "DU_survey_f2023",
    "figs",
    "03_fig.png"
  )
)
```

```{r}
plot(o23$biden_dp)
```

