---
title: "Selection: Why We Need Variation for Correlation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Learning Goals

- Understand why variation is essential for learning about the correlation between two features of the world.
- Understand what **selecting on the dependent variable** is and why it is not recommended for quantitative analysis.
- Understand why the temptation/pressure to select on the DV persists.


## Where we've been

We've been talking about correlation. Last time we introduced some important concepts related to correlation (mean, variance, covariance, linear regression, etc.) and discussed the utility of correlations (description, prediction, and causal inference). [[Check out the correlation notes here](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20201/Notes/01_correlation.md)]


## Where we're going

At present, I want us to sit with a particular element of correlation: **variation**.

To estimate a correlation between two features of the world, we need variation in each of those features. This may seem obvious, but you'd be surprised by the number of real-world cases (in the media or even scholarly articles) that are missing variation in at least one variable.

In Chapter 2 of our *Thinking Clearly with Data* text, we saw a few examples of facts, some of which are correlations and others of which are not. Can you remember which are which?

1. People who live to 100 years of age typically take vitamins.
2. Cities with more crime tend to higher more police officers.
3. Successful people have spent at least ten thousand hours honing their craft.
4. Most politicians facing a scandal win reelection.
5. Older people vote more than younger people.

2 and 5 are the only two statements above that tell us about a correlation. 1, 3, and 4 sound like they tell us something about the relationship between two features of the world, but in reality they are missing variation in one of those features.

Take 1. The below R code simulates some data based on the kind of variation (or lack thereof) that statement 1 gives us. Can you catch the problem?

```{r}
library(tidyverse) 

# simulate some data
vitamin_data <- tibble(
  live_to_100 = "Yes",
  take_vitamins = rbinom(n = 200, size = 1, prob = 0.75)
)          # n = number of obs; size means 0 or 1; prob means expected proportion

# plot the relationship
vitamin_data %>%
  group_by(live_to_100) %>%
  summarize(
    mean = mean(take_vitamins)
  ) %>%
  ggplot() +
  aes(x = live_to_100, y = mean) +
  geom_col(width = 0.5) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = "Live to 100?",
    y = "% who take vitamins"
  )
```

The problem is obvious. We don't have variation in age!

How about 5? Or something close to it anyway. In the below code, I'm attaching the `{socviz}` package which lets me access a dataset called `gss_sm`. This is a selection of individual responses to the 2016 General Social Survey. It has a variable for age and a variable for whether a person voted for Barak Obama in the 2012 election. Let's check out their relationship.

```{r}
library(socviz)
ggplot(gss_sm) +
  aes(x = age,
      y = obama) +
  geom_smooth()
```

Can you identify a relationship? Of course you can! That's because we have variation in both of these features of the world. We have some people who are older, some younger. And among those groups we have people who voted for Obama in 2012 and others that didn't.

This analysis tells us something about a relationship or regularity in the world. The first dealing with longevity and vitamins does not. The fact that people who live to 100 years of age regularly take vitamins does not tell us whether taking vitamins actually promotes longevity. It's just a bit of trivia about a particular age group.


## Selecting on the DV

The mistake that was made in statement 1 is an example of **selecting on the dependent variable**. Why do many people do this?

It's actually easier than you many think to select on the DV, even without noticing. After all, the statement that "people who live to be 100 typically take vitamins" sounds scientific. But when we make things a little more explicit (like I did with the R code example above), it's plain to see that this statement doesn't actually tell us anything useful (certainly not actionable).

A lot of people slip into selecting on the DV when trying to make correlational (even causal) claims. This is partially a handicap of talking about these things in plain English. I have to admit, I had to think carefully about claims 1-5 above before I knew with certainty which dealt with correlations and which didn't.

The dead give-away that a claim or analysis is predicated on selecting on the DV is that it starts with some phenomenon or outcome and then tries to look backward to identify causes. This is the case with the claim about vitamins and longevity. It starts with an outcome (living to over 100 years of age) and looks back at habits or lifestyle factors that are common to people who live to be this old. 

We see a few different examples of this error in practice in *Thinking Clearly with Data*.

1. Malcom Gladwell's claim that 10,000 hours of practice leads to becoming a great acheiver.
2. Rock'n Roll is currupting the youth.
3. High schoold dropouts.
4. Suicide attacks.

Let's take a look at an example using some data that were collected in a 1998 field experiment involving a get out the vote (GOTV) campaign in New Haven, CT.

```{r}
url <- "https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20201/Data/GOTV_Experiment.csv"
Data <- read_csv(url)
```

This dataset is in tidy format. It has 9 columns each corresponding to a different variable measured on individuals (the unit of observation). These variables are:

- female: 0 = no, 1 = yes
- age: in years
- white: 0 = no, 1 = yes
- black: 0 = no, 1 = yes
- employed: 0 = no, 1 = yes
- urban: 0 = no, 1 = yes
- treatmentattempt: 0 = did not receive a GOTV phone call, 1 = did
- successfultreatment: 0 = was not treated or didn't answer the phone, 1 = answered the phone if treated
- turnout: 0 = didn't vote, 1 = did vote

For the sake of argument, let's just look at folks who turned out to vote:

```{r}
Data_turnout <- Data %>%
  filter(turnout == 1)
```

With this data, can we answer the question: did getting a call increase turnout? Let's see:

```{r}
Data_turnout %>%
  summarize(
    phone_call = mean(treatmentattempt),
    answered_phone = mean(successfultreatment)
  )
```

About 53-54 percent of people who turned out to vote received a GOTV phone call. 44 percent answered the phone. Did getting a call lead to an increase in the likelihood of turnout?

Maybe it would help if we knew the likelihood that anyone got a call:

```{r}
# What share people got a GOTV call in total?
Data %>%
  summarize(
    phone_call = mean(treatmentattempt),
    answered_phone = mean(successfultreatment)
  )
```

About 50 percent of all people got a call, and among those a little over 39 percent answered the phone. Did the GOTV campaign work?

It's tempting to think that this is enough information. But we still don't have variation in the outcome of interest: turnout. Let's see what the data look like with this little detail in place:

```{r}
Data %>%
  group_by(turnout) %>%
  summarize(
    phone_call = mean(treatmentattempt),
    answered_phone = mean(successfultreatment)
  )
```

Now we're cooking! Those who came out to vote disproportionately received phone calls. They were also more likely to answer the phone.

In reality, there are some issues with this field experiment, one of which is obvious from the output above. We're going to bracket that for now. It is most important to note that drawing inferences by selecting on the DV is not only irresponsible; it can be dangerous, or at the very least expensive. 

If you only looked at the phone call data among individuals who voted in an election, you might conclude that the phone calls didn't actually help. But it turns out, with the bigger picture in view, people who turned out to vote were more likely to get a call and more likely to answer.


## Try out some examples yourself

Here are links to four different datasets. For each one, there's a research question. Get the data into R and see if it gives you the variation you need to answer the research question. If it can, answer the research question. If it can't, can you leverage the data to draw inferences about other correlations?

#### Terrorism and Foreign Occupation

- **link:** https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20201/Data/terror_data.csv

- **Research Question:** Does the presence of a foreign occupying force in a country lead to an increase in the number of suicide terrorist attacks?

#### Vote on Immigration Proposition

- **link:** https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20201/Data/immigration_proposition.csv

- **Research Question:** Does political ideology predict support for a pro-immigration proposition on the ballot?

#### Gender in Congress

- **link:** https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20201/Data/gender_in_congress.csv

- **Research Question:** Are women more likely to hold a seat in the U.S. Congress if they are Republicans or Democrats?

#### Economic Growth and Immigration

- **link:** https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20201/Data/gdp_and_immigration.csv

- **Research Question:** Do countries with higher GDP (gross domestic product) attract more immigrants?