---
title: "Estimating the ATE in R"
output: 
  github_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")
```

## Goals

- The average treatment effect or ATE is usually our target estimand in causal inference.
- R provides an infinite variety of tools for estimating the ATE, but there are advantages to using the `lm_robust()` function from `{estimatr}`.
- There are also many tools for reporting your results like regression tables and coefficient plots.


## The Average Treatment Effect

The average treatment effect is defined as the *average difference in some outcome due to a causal variable or treatment*.

In a recent assigment, we talked about the effects of letters to households on household applications for fire-proofing inspections.

Let's use $T \in \{0, 1\}$ to denote treatment where 1 means a household got a letter and 0 means they didn't. Let $Y \in \{0, 1\}$ denote whether a household applied for an inspection:

$$
\begin{aligned}
Y_{1i} & = \text{outcome for unit } i \text{ if } T = 1\\
Y_{0i} & = \text{outcome for unit } i \text{ if } T = 0
\end{aligned}
$$

Using this notation, the effect of the letters on applications for household $i$ is:

$$\delta_i = Y_{1i} - Y_{0i}$$

This difference or "delta" tells us the causal effect of the letters. 
Often times, we're more interested in the average causal effect rather than the effect for an individual observation. We would represent this as follows:

$$
\begin{aligned}
\overline Y_{1} & = \text{average outcome for all units if } T = 1\\
\overline Y_{0} & = \text{average outcome for all units if } T = 0
\end{aligned}
$$

With this notation, we can specify the *average treatment effect* or ATE as:

$$\text{ATE} = \overline \delta = \overline Y_1 - \overline Y_0$$

As you'll recall, we face a practical challenge when it comes to calculating the ATE. Namely, we can't actually calculate it---not directly anyway---because we cannot observe all potential outcomes in the study. The randomization process leads the the following revealed outcomes:

$$Y_i = Y_{1i} \times T_i + Y_{0i} \times (1 - T_i)$$

We can only estimate the ATE using the revealed outcomes. The simplest way to do this is to take the average of the outcome among the treated and compare it to the average of the outcome among the untreated:

$$
\begin{aligned}
\overline Y(T_i = 1) & = \text{average outcome for all units where } T_i = 1\\
\overline Y(T_i = 0) & = \text{average outcome for all units where } T_i = 0
\end{aligned}
$$

Using this notation we have:

$$\hat \delta = \overline Y(T_i = 1) - \overline Y(T_i = 0)$$

This "delta hat" is the best we can do, and under the right conditions, it actually is a pretty good approximation for the true ATE.


## How to estimate the ATE in R

In the most recent assignment, you all estimated the ATE for letters on inspections using a linear regression model. In political science, this is our preferred way to do this. There are other approaches, but linear regression is the most flexible and it provides a nice framework for adjusting for confounding covariates.

At places like the Office of Evaluation Sciences the method we always teach newcomers is the `lm_robust()` function from the `{estimatr}` package. We prefer this to `lm()` for a few reasons:

1. `lm_robust()` makes doing statistical inference much easier.
2. `lm_robust()` can be quickly converted to a form useful for data viz.
3. `lm_robust()` makes dealing with fixed effects easier (we'll talk about these later).

Let's look at an example involving conflict and the Cold War. We're going to use some political science data from the `{peacesciencer}` package:
```{r}
library(tidyverse)
library(peacesciencer)
create_stateyears(subset_years = 1945:2007) %>%
  add_cow_wars(type = "intra") %>%
  add_sdp_gdp() %>%
  add_democracy() -> Data
```

The period during the collapse of the Soviet Union was a tumultuous time. Did instability in the USSR have an effect on political instability and conflict globally? Let's add a variable to our data that indicates the period of time during which the USSR began to and ultimately did meet its end:
```{r}
Data %>%
  mutate(
    ussr_collapse = ifelse(year %in% 1989:1992, 1, 0)
  ) -> Data
```

Then we'll fit the model:
```{r}
library(estimatr)
lm_fit <- lm_robust(
  cowintraonset ~ ussr_collapse, 
  data = Data,
  se_type = "HC1"
)
```

If we use summary to check the results, we can see whether this period really was more conflict-prone:
```{r}
summary(lm_fit)
```

The estimate is indeed statistically significant and positive. Let's interpret the coefficients really quickly.

- Intercept: This tells us the average rate of civil war onset durring all years in the data other than 1989-1992. This value equals 0.01387. Multiply that by 100 and we have a conflict initiation rate of 1.4%.
- Slope: This tels us the difference in the average rate of civil war onset outside of 1989-1992 and during this period. This value equals 0.01829. Multiply that by 100 and we have an average 1.8 percentage point increase in conflict initiation during this period. That's more than double the conflict.

The results suggest that the collapse of the Soviet Union may have had an effect on conflict globally. Does that make sense to you? Can you think of reasons why this correlation might be spurious?

Our data includes a number of other covariates that we can try to adjust for. We'll talk more indepth about controlling for covariates later. For now, let's just add a few under the assumption that they help us rule out whether these other variables might explain the correlation between the collapse of the USSR and global conflict:
```{r}
lm_fit_controls <- lm_robust(
  cowintraonset ~ ussr_collapse + 
    v2x_polyarchy + wbgdppc2011est + wbpopest,
  data = Data,
  se_type = "HC1"
)
```

The above allows us to control for the following covariates:

- `v2x_polyarchy`: This is the Varieties of Democracy (V-Dem) project's measure of the quality of democracy. Values range from 0-1 where 1 means most democratic and 0 means least.
- `wbgdppc2011est`: This is the natural log of GDP per capita in countries over time.
- `wbpopest`: This is the natural log of country population over time.

Does accounting for these variables change our estimate of the effect of the Soviet Union's collapse?
```{r}
summary(lm_fit_controls)
```

Wow! Rather than reducing the estimate, it's greater in magnitude and has an even smaller p-value. 

Since we're now controlling for other variables, we would say that, *all else equal*, the collapse of the USSR increased the rate of conflict intitiation by 2.3 percentage points.


## Summarizing the ATE with Regression Tables

When we report results for the ATE, we either will make a regression table or a coefficient plot. We talked about these things before, but we didn't discuss how to make them.

Because people use R so much for their research, there are lots of available tools for producing professional looking regression tables in R. One of my favorites is `{texreg}`. If you don't have it, you can install it by writing the following in the console:
```
install.packages("texreg")
```

We can easily plot the results from one or more regression models side-by-side. Here's a table with just our original model:
```{r}
library(texreg)
screenreg(lm_fit)
```

Here's one that compares both of our models:
```{r}
screenreg(
  list(lm_fit, lm_fit_controls)
)
```

The above tables are produced using `screenreg()`. This produces output that you can easily read in your R console. You can also produce html tables:
```{r}
htmlreg(list(lm_fit, lm_fit_controls)) %>%
  htmltools::HTML()
```

By default, these `*reg()` functions produce tables with 95% confidence intervals. If you want to show standard errors instead, you can just say so:
```{r}
htmlreg(list(lm_fit, lm_fit_controls),
        include.ci = F) %>%
  htmltools::HTML()
```

Same if you want to show more digits after the decimal point:
```{r}
htmlreg(list(lm_fit, lm_fit_controls),
        include.ci = F,
        digits = 3) %>%
  htmltools::HTML()
```

There generally is no need to show more than 3.

Finally, you can clean up the variable names and add a title:
```{r}
htmlreg(list(lm_fit, lm_fit_controls),
        include.ci = F,
        digits = 3,
        caption = "My Regression Estimates",
        caption.above = T,
        custom.coef.names = c("Intercept",
                              "USSR Collapse",
                              "Democracy",
                              "GDP per Capita",
                              "Population")) %>%
  htmltools::HTML()
```


## Summarizing the ATE with Coefficient Plots

In some settings, communicating about your results with regression tables doesn't do you many favors. In presentations or posters, a coefficient plot can be better.

```{r}
tidy(lm_fit_controls) %>%
  ggplot() +
  aes(x = estimate,
      xmin = conf.low,
      xmax = conf.high,
      y = term) +
  geom_pointrange()
```

Generally, you'll want to clean up the data viz a bit, just like with regression tables:
```{r}
tidy(lm_fit_controls) %>%
  ggplot() +
  aes(x = estimate,
      xmin = conf.low,
      xmax = conf.high,
      y = term) +
  geom_pointrange() +
  # add a vertical line at zero
  geom_vline(xintercept = 0,
             linetype = 2) +
  # update the plot labels
  labs(x = "Estimate with 95% CI",
       y = NULL,
       title = "My Regression Estimates") +
  # update variable names
  scale_y_discrete(labels = c("Intercept",
                              "USSR Collapse",
                              "Democracy",
                              "GDP per Capita",
                              "Population"))
```


Say you wanted to highlight how controlling for democracy, GDP per capita, and population change the estimates. You can do that, too:
```{r}
bind_rows(
  tidy(lm_fit),
  tidy(lm_fit_controls)
) %>%
  filter(term == "ussr_collapse") %>%
  mutate(model = c("No Controls", "Controls")) %>%
  ggplot() +
  aes(x = estimate,
      xmin = conf.low,
      xmax = conf.high,
      y = term,
      color = model) +
  geom_pointrange(position = ggstance::position_dodge2v(.5)) +
  geom_vline(xintercept = 0,
             linetype = 2) +
  labs(x = "Estimate with 95% CI",
       y = NULL,
       title = "Comparison of Regression Estimates",
       color = "Model includes... ") +
  scale_y_discrete(label = "USSR Collapse")
```

