---
title: "Simulate Data for AC4"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = F,
                      message = F,
                      out.width = "75%",
                      dpi = 500)
library(tidyverse)
library(estimatr)
library(fabricatr)
library(randomizr)
library(coolorrr)
set_theme()
set_palette()
```


First, simulate the background data:
```{r}
fabricate(
  county_id = add_level(
    N = 2, 
    county = c("County A", "County B"),
    county_size = c(1000, 2000),
    county_urban = c(0.2, 0.5)
  ),
  hh_id = add_level(
    N = county_size,
    urban = rbinom(N, 1, county_urban),
    hhincome = 1000 * exp(rnorm(N, 
                         mean = log(urban * 50 + (1 - urban) * 80))),
    acres = exp(rnorm(N,
                      mean = log(urban * 0.5 + (1 - urban) * 6)))
)) -> Data
```


Confirm that income and and acres are exponentially distributed:
```{r}
hist(Data$hhincome)
hist(Data$acres)
```

Cool. Now let's assign treatment and specify the outcome:
```{r}
f <- function(x) 1 / (1 + exp(-x))
Data %>%
  mutate(
    letter = block_ra(blocks = paste(county, urban),
                      conditions = c(0, 1),
                      block_m = rep(100, 4)),
    app0 = rbinom(
      n(), 1,
      f(-5.8 + 0.1 * log(hhincome) + 0.2 * log(acres) - 0.5 * urban)
    ),
    app1 = rbinom(
      n(), 1,
      f(-3.5 + 0.1 * log(hhincome) + 0.3 * log(acres) - 1 * urban)
    ),
    app = app0 * (1 - letter) + app1 * letter
  ) -> Data
Data %>%
  group_by(county, urban) %>%
  summarize(
    sum(letter),
    n(),
    mean(letter)
  )
```


Alright, so we've assigned treatment and revealed outcomes. If we take the true ATE and compare it to the estimated one, do we get biases?
```{r}
trueATE <- mean(Data$app1 - Data$app0)
obsATE <- coef(lm(app ~ letter, Data))[2]
c(Truth = trueATE, Observed = obsATE)
```

Technically, we can only estimate the intent to treat effect or ITT, but I won't tell the students that (yet). Assuming compliance with the treatment weren't an issue, I'd do the following:

```{r}
fit <- lm_robust(
  app ~ letter * (ccounty + curban) + hhincome + acres, 
  data = Data %>%
    mutate(
      ccounty = cmean(county=="County B"),
      curban = cmean(urban)
    )
)
summary(fit)
```

```{r}
library(oesr)
oes_prep(fit, 
         treatment_vars = "letter",
         scale = "percentage") %>%
  oes_plot()
```

Lock it in:
```{r}
write_csv(
  Data,
  here::here("DPR 201", "Data", "montana_wildfires.csv")
)
```


How about we test this a little more systematically using {seerrr}:
```{r}
library(seerrr)
simTheData <- function() fabricate(
  county_id = add_level(
    N = 2, 
    county = c("County A", "County B"),
    county_size = c(1000, 2000),
    county_urban = c(0.2, 0.5)
  ),
  hh_id = add_level(
    N = county_size,
    urban = rbinom(N, 1, county_urban),
    hhincome = 1000 * exp(rnorm(N, 
                         mean = log(urban * 50 + (1 - urban) * 80))),
    acres = exp(rnorm(N,
                      mean = log(urban * 0.5 + (1 - urban) * 6))),
    letter = block_ra(blocks = paste(county, urban),
                      conditions = c(0, 1),
                      block_m = rep(100, 4)),
    app0 = rbinom(
      N, 1,
      f(-5.8 + 0.1 * log(hhincome) + 0.2 * log(acres) - 0.5 * urban)
    ),
    app1 = rbinom(
      N, 1,
      f(-3.5 + 0.1 * log(hhincome) + 0.3 * log(acres) - 1 * urban)
    ),
    app = app0 * (1 - letter) + app1 * letter 
  )
)

library(furrr)
plan(multicore, workers = 3)
simData <- future_map(
  1:500,
  ~ simTheData(),
  .options = furrr_options(seed = T)
)
```


Get the range of estimates:
```{r}
simEst1 <- estimate(
  simData, app ~ letter, "letter", estimator = lm
)
simEst2 <- estimate(
  simData, app ~ letter + county + urban, estimator = lm,
  "letter"
)
cmean <- function(x) x - mean(x, na.rm=T)
simEst3 <- estimate(
  simData, app ~ letter * (cmean(county == "County B") +
                             cmean(urban)),
  "letter",
  estimator = lm
)
```


Check for bias:
```{r}
theTruth <- simData %>%
  map(~ mean(.x$app1 - .x$app0)) %>%
  unlist() %>%
  mean()
simEval1 <- evaluate(
  simEst1, what = "bias", truth = theTruth
)
simEval2 <- evaluate(
  simEst2, what = "bias", truth = theTruth
)
simEval3 <- evaluate(
  simEst3, what = "bias", truth = theTruth
)
bind_rows(
  simEval1,
  simEval2,
  simEval3
) %>%
  mutate(
    model = c(1, 2, 3)
  ) %>%
  ggplot() +
  aes(x = model,
      y = bias,
      ymin = bias - 1.96 * std.error,
      ymax = bias + 1.96 * std.error) +
  geom_point() +
  geom_errorbar(width = 0.5) +
  geom_hline(
    yintercept =  0,
    lty = 2
  ) + 
  scale_x_continuous(
    breaks = 1:3,
    labels = c("No Controls", "Dummies", "Dummies with\nInteraction")
  ) +
  labs(
    x = NULL,
    y = "Bias",
    title = paste0("Results after ", 
                   scales::comma(length(simData)),
                   " simulations")
  )
```



