---
title: "March Madness, 2023"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")
```

## Introduction

Let's try out some predictions for March Madness 2023.

## Setup

First, I need to gather my tools and set some options for my session:
```{r}
library(tidyverse)
library(coolorrr)
set_theme()
set_palette()
```

## Data

Now I need to read in and merge some historical data on team performance in past years. The below files come from Kaggle: https://www.kaggle.com/competitions/mens-march-mania-2022/data

```{r}
## First, team identifiers
d1 <- read_csv("MTeams.csv")

## Then, team seeds
d2 <- read_csv("MNCAATourneySeeds.csv")

## Finally, outcomes
d3 <- read_csv("MRegularSeasonCompactResults.csv")
```

For each team in a given year, I want a record and the seed:
```{r}
d1 %>%
  select(TeamID, TeamName) %>%
  right_join(
    d2
  ) %>%
  mutate(
    Seed = str_remove(Seed, "[A-Z]") %>%
      str_remove("[a-z]") %>%
      as.numeric()
  ) -> Data

d3 %>%
  group_by(Season, WTeamID) %>%
  summarize(
    Score = mean(WScore),
    Differential = mean(WScore - LScore),
    N = n()
  ) %>%
  ungroup %>%
  rename(TeamID = WTeamID) -> win_d
d3 %>%
  group_by(Season, LTeamID) %>%
  summarize(
    Score = mean(LScore),
    Differential = mean(LScore - WScore),
    N = n()
  ) %>%
  ungroup %>%
  rename(TeamID = LTeamID) -> loss_d

bind_rows(
  win_d,
  loss_d
) %>%
  group_by(Season, TeamID) %>%
  summarize(
    AvgScore = sum(N * Score) / sum(N),
    Differential = sum(N * Differential) / sum(N)
  ) %>%
  ungroup -> outcomes

Data %>%
  left_join(outcomes, by = c("Season", "TeamID")) -> Data
```


## Explore some trends

It should probably be the case that teams with better records have better seeds:
```{r}
ggplot(Data) +
  aes(x = Differential,
      y = Seed) +
  geom_point() +
  geom_smooth(method = "loess",
              se = F) +
  labs(x = "Average Point Differential",
       y = "Tourney Seed",
       title = "A scatter plot")
```

How about a summary:
```{r}
library(socsci)
Data %>%
  group_by(Seed) %>%
  mean_ci(Differential) %>%
  ggplot() +
  aes(x = mean,
      y = Seed,
      xmin = lower,
      xmax = upper) +
  geom_pointrange() +
  geom_vline(xintercept = 0) +
  labs(x = "Average Differential",
       y = "Seed",
       title = "A dotwhisker plot") +
  scale_y_continuous(
    breaks = 1:16
  )
```

## Adding tourney performance

Okay, I need to add in performance in the tournaments:
```{r}
d4 <- read_csv("MNCAATourneyCompactResults.csv")
```

I think I want to make something like a directed dyad dataset:
```{r}
bind_rows(
  d4 %>%
    rename(TeamID1 = WTeamID,
           Score1 = WScore,
           TeamID2 = LTeamID,
           Score2 = LScore),
  d4 %>%
    rename(TeamID2 = WTeamID,
           Score2 = WScore,
           TeamID1 = LTeamID,
           Score1 = LScore)
) %>%
  select(1:6) -> tourney_data
```

Okay, let's find a way to match this with the season data:
```{r}
# Make a version for each team of the dyad
Data1 <- Data
Data2 <- Data
names(Data1) <- paste0(names(Data), 1)
names(Data2) <- paste0(names(Data), 2)

# I need the season column to be the same in each
Data1 <- Data1 %>% rename(Season = Season1)
Data2 <- Data2 %>% rename(Season = Season2)

# These should merge nicely with the tourney data
tourney_data %>%
  left_join(Data1, by = c("TeamID1", "Season")) %>%
  left_join(Data2, by = c("TeamID2", "Season")) -> full_data
```

## Look at the data again

Can we see a correlation between regular season performance and tourney outcomes?

Let's try a scatter plot:
```{r}
library(geomtextpath)
ggplot(full_data) +
  aes(x = AvgScore1,
      y = Score1) +
  geom_point(color = "darkgray",
             alpha = 0.2) +
  labs(x = "Average Season Points Scored",
       y = "Points Scored in Tourney") +
  geom_textabline(
    slope = 1,
    intercept = 0,
    label = "45 degrees",
    linetype = 2,
    color = "red",
    hjust = .8
  ) +
  geom_textsmooth(
    method = lm,
    label = "OLS Line of Best Fit",
    color = "darkblue"
  )
```

Yeah, just about. But...
```{r}
ggplot(full_data) +
  aes(x = Score1 - AvgScore1) +
  geom_density(fill = "gray",
               color = "black") +
  geom_vline(xintercept = 0,
             lty = 2) +
  labs(x = "Tourney Score minus Season Avg.")
```

It looks like teams tend to do slightly worse in March Madness than in the regular season. On average, they do about 5 points worse, and this difference is statistically significant, too:
```{r}
t.test(Score1 - AvgScore1 ~ 1, full_data)
```

Now, let's take a look at differentials:
```{r}
ggplot(full_data) +
  aes(x = Differential1 - Differential2,
      y = Score1 - Score2) +
  geom_point(color = "darkgray",
             alpha = 0.2) +
  geom_textabline(
    slope = 1,
    intercept = 0,
    label = "45 degrees",
    linetype = 2,
    color = "red",
    hjust = .8
  ) +
  geom_textsmooth(
    method = lm,
    label = "OLS Line of Best Fit",
    color = "darkblue"
  )
```

I wonder if seed is predicitive?
```{r}
ggplot(full_data) +
  aes(x = Seed1 - Seed2,
      y = Score1 - Score2) +
  geom_point(color = "darkgray",
             alpha = 0.2) +
  geom_textsmooth(
    method = lm,
    label = "OLS Line of Best Fit",
    color = "darkblue"
  )
```

Ooo... That's pretty good. I think I have some ideas for a predictive model.

But, let's do one more thing. In a previous figure I got the impression that there could be a secular upward trend in the regular season point differentials for Tourney teams. Is that worth modeling?
```{r}
ggplot(full_data) +
  aes(x = Season,
      y = Score1 - Score2) +
  geom_point(color = "darkgray",
             alpha = 0.2) +
  labs(x = "Season",
       y = "Tourney Differential") +
  geom_smooth(method = lm)
```

Nah. That's pretty flat. Okay...on with the model.

## Train the model

I'm going to make a model that predicts the tourney point differential as a function of the seed differential between the teams matched against each other and the difference in their regular season average point differential.

The model:
```{r}
model <- tourneyDiff1 ~ diffDiff1 + seedDiff1
```

Add the necessary columns to the data:
```{r}
full_data %>%
  mutate(tourneyDiff1 = Score1 - Score2,
         seedDiff1 = Seed1 - Seed2,
         diffDiff1 = Differential1 - Differential2) -> full_data
```

Fit the model to the data:
```{r}
fit <- lm(model, full_data)
```

Let's take a quick look:
```{r}
summary(fit)
```

Okay, it's a good idea to try some hypothetical predictions and also to translate predictions to win probabilities. To do that, I want to take a bootstrapping procedure.

First, my bootstrapping function:
```{r}
boot_lm <- function(form, data, R = 500) {
  ## The main fit
  fit <- lm(form, data)
  
  ## Bootstrap the data and refit
  boot_fits <- 1:R %>%
    map(~sample_n(data, nrow(data), T)) %>%
    map(~lm(form, .x))
  
  ## Return the output
  list(
    fit = fit,
    boot_fits = boot_fits
  )
}
```

Now, apply it:
```{r}
boot_fit <- boot_lm(model, full_data, R = 999)
```

Okay, now I need a function that will get the original plus bootstrapped predictions:
```{r}
boot_predict <- function(fit, newdata = NULL, ref = 0) {
  ## The main prediction
  pred <- predict(fit$fit, newdata)
  ## The bootstrapped predictions
  boot_preds <- fit$boot_fits %>%
    map(~ predict(.x, newdata)) 
  ## For each obs, get prob x > ref
  boot_probs <- 1:nrow(newdata) %>%
    map_dfr(
      ~ {
        obs <- .x
        boot_preds %>%
          map_dfr(
            ~ tibble(val = .x[obs]) 
          ) %>%
          summarize(prob = mean(val > 0)) %>%
          mutate(pred = pred[obs]) %>%
          select(pred, prob) 
      }
    )
  ## Return
  list(pred = pred,
       boot_preds = boot_preds,
       boot_probs = boot_probs)
}
```

Try it out:
```{r}
newdata <- tibble(
  diffDiff1 = -5:5,
  seedDiff1 = 3
)
boot_pred <- boot_predict(fit = boot_fit, newdata, ref = 0)
boot_pred$boot_probs
```

## Make predictions for 2023

Okay, first I need information about the season performance and seeds of current teams...

An hour of data collection later, I have it!

```{r}
googsheeturl <- "https://docs.google.com/spreadsheets/d/1JdIoZBExz5AygMH7E6QMfTCjQ75wkiRMpimdNq-2qiU/edit?usp=sharing"
library(googlesheets4)
gs4_deauth()
data23 <- range_speedread(googsheeturl)
glimpse(data23)
```

Okay, let's create seedDiff1 and diffDiff1:
```{r}
data23 %>%
  mutate(
    seedDiff1 = Seed1 - Seed2,
    diffDiff1 = Differential1 - Differential2
  ) -> data23
```

Now, I can make some first round predictions:
```{r}
fst_pred <- boot_predict(boot_fit, data23)
fst_pred$boot_probs %>%
  cbind(., data23)
```

Okay, so I need to keep the teams that are predicted to win:
```{r}
fst_pred$boot_probs %>%
  cbind(., data23) %>%
  mutate(
    Winner = ifelse(prob > 0.5, 
                    TeamName1, 
                    TeamName2),
    Prediction = paste0("+",round(abs(pred), 2)),
    Probability = ifelse(prob > 0.5,
                  paste0(round(100 * prob, 3), "%"),
                  paste0(round(100 * (1 - prob), 3), "%")) 
  ) %>%
  select(
    Winner, Prediction, Probability
  ) -> fst_round_winners
fst_round_winners %>%
  kableExtra::kable("html", caption = "1st Round Predictions")
```

Now, set things up for second round. I need to match the team above with the team below in pairs.
```{r}
r2_data <- tibble(
  TeamName = fst_round_winners$Winner,
  side = paste0("TeamName",rep(1:2, len = 32))
) %>%
  pivot_wider(
    values_from = TeamName,
    names_from = side
  ) %>%
  unnest %>%
  mutate(
    Differential1 = 0,
    Seed1 = 0,
    Differential2 = 0,
    Seed2 = 0
  )
keep1 <- which(r2_data$TeamName1 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r2_data$TeamName1) 
r2_data[keep1, 3:4] <- data23[keep2, 2:3]
keep1 <- which(r2_data$TeamName2 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r2_data$TeamName2) 
r2_data[keep1, 5:6] <- data23[keep2, 2:3]
keep1 <- which(r2_data$TeamName1 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r2_data$TeamName1) 
r2_data[keep1, 3:4] <- data23[keep2, 5:6]
keep1 <- which(r2_data$TeamName2 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r2_data$TeamName2)
r2_data[keep1, 5:6] <- data23[keep2, 5:6]
r2_data
```

There we go! Now I need the new predictors:
```{r}
r2_data %>%
  mutate(
    seedDiff1 = Seed1 - Seed2,
    diffDiff1 = Differential1 - Differential2
  ) -> r2_data
```

Now I can make round 2 predictions:
```{r}
snd_pred <- boot_predict(boot_fit, r2_data)
```

Check out the predictions:
```{r}
snd_pred$boot_probs %>%
  cbind(., r2_data) %>%
  mutate(
    Winner = ifelse(prob > 0.5, 
                    TeamName1, 
                    TeamName2),
    Prediction = paste0("+",round(abs(pred), 2)),
    Probability = ifelse(prob > 0.5,
                  paste0(round(100 * prob, 3), "%"),
                  paste0(round(100 * (1 - prob), 3), "%")) 
  ) %>%
  select(
    Winner, Prediction, Probability
  ) -> snd_round_winners
snd_round_winners %>%
  kableExtra::kable("html", caption = "2nd Round Predictions")
```

Same thing as before. Line up teams by who they'll play:
```{r}
r3_data <- tibble(
  TeamName = snd_round_winners$Winner,
  side = paste0("TeamName",rep(1:2, len = 16))
) %>%
  pivot_wider(
    values_from = TeamName,
    names_from = side
  ) %>%
  unnest %>%
  mutate(
    Differential1 = 0,
    Seed1 = 0,
    Differential2 = 0,
    Seed2 = 0
  )
keep1 <- which(r3_data$TeamName1 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r3_data$TeamName1) 
r3_data[keep1, 3:4] <- data23[keep2, 2:3]
keep1 <- which(r3_data$TeamName2 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r3_data$TeamName2) 
r3_data[keep1, 5:6] <- data23[keep2, 2:3]
keep1 <- which(r3_data$TeamName1 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r3_data$TeamName1) 
r3_data[keep1, 3:4] <- data23[keep2, 5:6]
keep1 <- which(r3_data$TeamName2 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r3_data$TeamName2)
r3_data[keep1, 5:6] <- data23[keep2, 5:6]
r3_data
```

There we go! Now I need the new predictors:
```{r}
r3_data %>%
  mutate(
    seedDiff1 = Seed1 - Seed2,
    diffDiff1 = Differential1 - Differential2
  ) -> r3_data
```

Now I can make round 3 predictions:
```{r}
trd_pred <- boot_predict(boot_fit, r3_data)
```

Check out the predictions:
```{r}
trd_pred$boot_probs %>%
  cbind(., r3_data) %>%
  mutate(
    Winner = ifelse(prob > 0.5, 
                    TeamName1, 
                    TeamName2),
    Prediction = paste0("+",round(abs(pred), 2)),
    Probability = ifelse(prob > 0.5,
                  paste0(round(100 * prob, 3), "%"),
                  paste0(round(100 * (1 - prob), 3), "%")) 
  ) %>%
  select(
    Winner, Prediction, Probability
  ) -> trd_round_winners
trd_round_winners %>%
  kableExtra::kable("html", caption = "3rd Round Predictions")
```


Same thing as before. Line up teams by who they'll play:
```{r}
r4_data <- tibble(
  TeamName = trd_round_winners$Winner,
  side = paste0("TeamName",rep(1:2, len = 8))
) %>%
  pivot_wider(
    values_from = TeamName,
    names_from = side
  ) %>%
  unnest %>%
  mutate(
    Differential1 = 0,
    Seed1 = 0,
    Differential2 = 0,
    Seed2 = 0
  )
keep1 <- which(r4_data$TeamName1 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r4_data$TeamName1) 
r4_data[keep1, 3:4] <- data23[keep2, 2:3]
keep1 <- which(r4_data$TeamName2 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r4_data$TeamName2) 
r4_data[keep1, 5:6] <- data23[keep2, 2:3]
keep1 <- which(r4_data$TeamName1 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r4_data$TeamName1) 
r4_data[keep1, 3:4] <- data23[keep2, 5:6]
keep1 <- which(r4_data$TeamName2 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r4_data$TeamName2)
r4_data[keep1, 5:6] <- data23[keep2, 5:6]
r4_data
```

There we go! Now I need the new predictors:
```{r}
r4_data %>%
  mutate(
    seedDiff1 = Seed1 - Seed2,
    diffDiff1 = Differential1 - Differential2
  ) -> r4_data
```

Now I can make round 4 predictions:
```{r}
fth_pred <- boot_predict(boot_fit, r4_data)
```

Check out the predictions:
```{r}
fth_pred$boot_probs %>%
  cbind(., r4_data) %>%
  mutate(
    Winner = ifelse(prob > 0.5, 
                    TeamName1, 
                    TeamName2),
    Prediction = paste0("+",round(abs(pred), 2)),
    Probability = ifelse(prob > 0.5,
                  paste0(round(100 * prob, 3), "%"),
                  paste0(round(100 * (1 - prob), 3), "%")) 
  ) %>%
  select(
    Winner, Prediction, Probability
  ) -> fth_round_winners
fth_round_winners %>%
  kableExtra::kable("html", caption = "4th Round Predictions")
```


Same thing as before. Line up teams by who they'll play:
```{r}
r5_data <- tibble(
  TeamName = fth_round_winners$Winner,
  side = paste0("TeamName",rep(1:2, each = 2))
) %>%
  pivot_wider(
    values_from = TeamName,
    names_from = side
  ) %>%
  unnest %>%
  mutate(
    Differential1 = 0,
    Seed1 = 0,
    Differential2 = 0,
    Seed2 = 0
  )
keep1 <- which(r5_data$TeamName1 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r5_data$TeamName1) 
r5_data[keep1, 3:4] <- data23[keep2, 2:3]
keep1 <- which(r5_data$TeamName2 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r5_data$TeamName2) 
r5_data[keep1, 5:6] <- data23[keep2, 2:3]
keep1 <- which(r5_data$TeamName1 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r5_data$TeamName1) 
r5_data[keep1, 3:4] <- data23[keep2, 5:6]
keep1 <- which(r5_data$TeamName2 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r5_data$TeamName2)
r5_data[keep1, 5:6] <- data23[keep2, 5:6]
r5_data
```

There we go! Now I need the new predictors:
```{r}
r5_data %>%
  mutate(
    seedDiff1 = Seed1 - Seed2,
    diffDiff1 = Differential1 - Differential2
  ) -> r5_data
```

Now I can make round 5 predictions:
```{r}
fvth_pred <- boot_predict(boot_fit, r5_data)
```

Check out the predictions:
```{r}
fvth_pred$boot_probs %>%
  cbind(., r5_data) %>%
  mutate(
    Winner = ifelse(prob > 0.5, 
                    TeamName1, 
                    TeamName2),
    Prediction = paste0("+",round(abs(pred), 2)),
    Probability = ifelse(prob > 0.5,
                  paste0(round(100 * prob, 3), "%"),
                  paste0(round(100 * (1 - prob), 3), "%")) 
  ) %>%
  select(
    Winner, Prediction, Probability
  ) -> fvth_round_winners
fvth_round_winners %>%
  kableExtra::kable("html", caption = "5th Round Predictions")
```


Same thing as before. Line up teams by who they'll play:
```{r}
r6_data <- tibble(
  TeamName = fvth_round_winners$Winner,
  side = paste0("TeamName",rep(1:2, len = 2))
) %>%
  pivot_wider(
    values_from = TeamName,
    names_from = side
  ) %>%
  unnest %>%
  mutate(
    Differential1 = 0,
    Seed1 = 0,
    Differential2 = 0,
    Seed2 = 0
  )
keep1 <- which(r6_data$TeamName1 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r6_data$TeamName1) 
r6_data[keep1, 3:4] <- data23[keep2, 2:3]
keep1 <- which(r6_data$TeamName2 %in% data23$TeamName1)
keep2 <- which(data23$TeamName1 %in% r6_data$TeamName2) 
r6_data[keep1, 5:6] <- data23[keep2, 2:3]
keep1 <- which(r6_data$TeamName1 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r6_data$TeamName1) 
r6_data[keep1, 3:4] <- data23[keep2, 5:6]
keep1 <- which(r6_data$TeamName2 %in% data23$TeamName2)
keep2 <- which(data23$TeamName2 %in% r6_data$TeamName2)
r6_data[keep1, 5:6] <- data23[keep2, 5:6]
r6_data
```




There we go! Now I need the new predictors:
```{r}
r6_data %>%
  mutate(
    seedDiff1 = Seed1 - Seed2,
    diffDiff1 = Differential1 - Differential2
  ) -> r6_data
```

Now I can make final predictions:
```{r}
final_pred <- boot_predict(boot_fit, r6_data)
```

Check out the predictions:
```{r}
final_pred$boot_probs %>%
  cbind(., r6_data) %>%
  mutate(
    Winner = ifelse(prob > 0.5, 
                    TeamName1, 
                    TeamName2),
    Prediction = paste0("+",round(abs(pred), 2)),
    Probability = ifelse(prob > 0.5,
                  paste0(round(100 * prob, 3), "%"),
                  paste0(round(100 * (1 - prob), 3), "%")) 
  ) %>%
  select(
    Winner, Prediction, Probability
  ) -> final_round_winners
final_round_winners %>%
  kableExtra::kable("html", caption = "Final Round Predictions")
```


