---
title: "Lab 06"
author: "Miles D. Williams"
date: '2023-03-10'
output: 
  html_document:
    toc: yes
    code_folding: hide
---

```{r setup, include=FALSE}
## The setup
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")

## My tools
library(tidyverse)
library(coolorrr)

## Customize plotting themes and colors
set_theme()
set_palette(
  binary = c("blue", "red"),
  from_coolors = F
)
```


First, I need to read in the data:
```{r}
housedata <- read_csv("House_all_members.csv")
```

Now I need to create a year column:
```{r}
housedata %>%
  mutate(
    year = 1789 + 2 * (congress - 1)
  ) -> housedata
```

Confirm that the years have the appropriate range:
```{r}
range(housedata$year)
```

Make a "polished" plot of the nominate_dim1 scores over time:
```{r}
ggplot(housedata) +
  aes(x = year,
      y = nominate_dim1) +
  geom_point(alpha = 0.2,
             color = "gray") +
  labs(x = NULL,
       y = "DW-scores",
       title = "Ideology of Congress over Time")
```

Do the same by party code and use geom_gitter:
```{r}
ggplot(housedata) +
  aes(x = year,
      y = nominate_dim1,
      color = factor(party_code)) +
  geom_jitter(alpha = 0.1,
              show.legend = F) +
  labs(x = NULL,
       y = "DW-scores",
       title = "Ideology of Congress over Time")
```

Just focus on polarization after Reconstruction (post 1876):
```{r}
house1876 <- housedata %>%
  filter(year > 1876)
```

Make a twoparty dummy:
```{r}
`%nin%` <- Negate(`%in%`)
house1876 %>%
  mutate(
    twoparty = ifelse(
      party_code %nin% c(100, 200),
      NA,
      ifelse(party_code==200, 1, 0)
    )
  ) -> house1876
```

Now add rep_or_dem:
```{r}
house1876 %>%
  mutate(
    rep_or_dem = ifelse(twoparty==0,
                        "Democrat",
                        "Republican")
  ) -> house1876
```

Reps should be positive and Dems negative:
```{r}
house1876 %>%
  group_by(rep_or_dem) %>%
  summarize(mean(nominate_dim1, na.rm=T))
```

Now let's make another plot with colors by the two parties:
```{r}
ggplot(house1876) +
  aes(x = year,
      y = nominate_dim1,
      color = rep_or_dem) +
  geom_jitter(alpha = 0.1) +
  labs(x = NULL,
       y = "DW-score",
       title = "Ideology in Congress over Time",
       color = NULL) +
  ggpal(type = "binary",
        na.translate = F)
```

Alright, one more thing. Adjust the x-axis to be more granular and at an angle:
```{r}
ggplot(house1876) +
  aes(x = year,
      y = nominate_dim1,
      color = rep_or_dem) +
  geom_jitter(alpha = 0.1) +
  labs(x = NULL,
       y = "DW-score",
       title = "Ideology in Congress over Time",
       color = NULL) +
  ggpal(type = "binary",
        na.translate = F) +
  scale_x_continuous(
    breaks = seq(1877, 2017, 10)
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Now let's test some claims about polarization over time. I want to show the average for each party and the total average over time to compare:
```{r}
house1876 %>%
  group_by(congress, rep_or_dem) %>%
  mutate(party_mean = mean(nominate_dim1, na.rm=T)) %>%
  group_by(congress) %>%
  mutate(
    congress_mean = mean(nominate_dim1, na.rm=T)
  ) -> house1876

house1876 %>%
  drop_na(rep_or_dem) %>%
  ggplot() +
  aes(x = year) +
  geom_line(aes(y = party_mean,
                     color = rep_or_dem)) +
  geom_line(aes(y = congress_mean),
                 color = "black") +
  ggpal(type = "binary") +
  scale_x_continuous(
    breaks = seq(1877, 2017, by = 10) 
  ) +
  labs(x = NULL,
       y = "Mean DW-Scores",
       title = "Mean DW-Scores over Time by Party",
       color = NULL,
       caption = "Note: Black line is the total mean") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Let's do some statistical tests. Is one party more polarized than the other?
```{r}
house1876 %>%
  ungroup %>%
  select(year, congress_mean, party_mean, rep_or_dem)%>%
  distinct()%>%
  drop_na() %>%
  pivot_wider(values_from = party_mean,
              names_from = rep_or_dem) -> house1876cong
```

Now let's recreate the last figure but in a kinda-sorta hacky way:
```{r}
set_palette(
  qualitative = c("black", "blue", "red"),
  binary = c("blue", "red"),
  from_coolors = F
)
ggplot(house1876cong) +
  aes(x = year) +
  geom_line(
    aes(y = congress_mean, 
        color = "All Members")
  ) +
  geom_line(
    aes(y = Republican,
        color = "Republicans")
  ) +
  geom_line(
    aes(y = Democrat,
        color = "Democrats")
  ) +
  ggpal() +
  labs(x = NULL,
       y = "Mean DW-score",
       title = "Mean DW-scores over Time by Party",
       color = NULL)
```

Is one party more polarized than the other?
```{r}
lm(abs(Democrat) - abs(Republican) ~ 1, house1876cong) %>%
  summary()
```

It looks like the Reps are farther to the right than the Dems are to the left. But the difference is actually quite small.