---
output: 
  stevetemplates::syllabus:
    latex_engine: xelatex
    dev: cairo_pdf
geometry: margin=1in

title: "DPR 201: Data & Design for Social Impact"
author: Dr. Miles D. Williams
date: "Fall 2022"

email: "williamsmd@dension.edu"
web: "milesdwilliams15.github.io"
officehours: "MWF 2:30-3:30 pm"
office: "Higley 401C"
classroom: "Higley 325"
classhours: "MWF 9:30-10:20 am"

fontfamily: mathpazo
fontsize: 11pt
header-includes:
   - \linespread{1.05}
   - \usepackage{amsmath,graphicx,amssymb}
---

<!-- CCCCCCCCnncdxxxxxxx -->

<!-- Fun fact: my cat added this comment as she walked over the keyboard while I was working on this template. -->

<!-- I'm going to keep it going forward. Long live Queen Bea. -->

```{r setup, include=FALSE}
# knitr::opts_chunk$set(cache=FALSE, dev='pdf')
knitr::opts_chunk$set(cache=F,
                      fig.path = 'figs/',
                      cache.path='cache/',
                      warning=F,
                      message=F)

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )


advdate <- function(obj, adv) {
 tmon <- obj + 7*(adv-1)
 tfri <- obj + 4 + 7*(adv-1)
 tmon <- format(tmon, format="%m/%d")
 tfri <- format(tfri, format="%m/%d")
 zadv <- sprintf("%02d", adv)
 tmp <- paste("Week ",zadv,sep='',", ", tmon," - ",tfri)
 return(tmp)
}

options(scipen=999)
library(tidyverse)
library(stevemisc)
#library(ggpmisc)
#library(anonymizer)
# library(ggcal)
#library(stringr)
#library(kfigr)
#library(broom)
library(lubridate)



# library(RefManageR)
# library(knitcitations)
# library(rcrossref)
# bib <- ReadBib(here::here("bib/refs.bib"))
#myopts <- BibOptions(bib.style = "authoryear", style="latex", first.inits=FALSE, max.names = 20)

# Create a calendar for your syllabus ----
# Source: http://svmiller.com/blog/2020/08/a-ggplot-calendar-for-your-semester/

# 1) what is the first Monday of the semester?
# Any number of ways to identify dates in R, but we'll use {lubridate} and the ymd() function here.
# Format: YYYYMMDD. In this example, 4 January 2021 (but no mas!).

mon <- ymd(20220829)

# What are some dates you won't be here? In this example, I had a conference on 7 January 2021.
# Spring Break was 15 March 2021 to 19 March 2021.
not_here_dates <- c(
  # Fall break,
  c(ymd(20221017), ymd(20221018)),
  # Tanksgivin'
  seq(ymd(20221121), ymd(20221125), by = 1))

# You can adjust this as you see fit. Basically: add assignment types (e.g. papers, quizzes).
# writing (MA) assignments:
ma_dates <- c(ymd(20221007), 
              ymd(20221028),
              ymd(20221118),
              ymd(20221215))
# graph challenges:
gc_dates <- c(ymd(20220918),
              ymd(20220925),
              ymd(20221002),
              ymd(20221016),
              ymd(20221106),
              ymd(20221204))

# What are the full dates of the semester? Here, I'll exclude exam week as I like to do.
# In this case: Aug. 29 to Dec. 13
semester_dates <- seq(ymd(20220829), ymd(20221213), by=1)

# Custom function for treating the first day of the month as the first week 
# of the month up until the first Sunday (unless Sunday was the start of the month)
wom <- function(date) {
    first <- wday(as.Date(paste(year(date),month(date),1,sep="-")))
    return((mday(date)+(first-2)) %/% 7+1)
  }

# Create a data frame of dates, assign to Cal
tibble(date = seq(ymd(20220801), ymd(20221231), by=1))  %>%
  mutate(mon = lubridate::month(date, label=T, abbr=F), # get month label
         wkdy = weekdays(date, abbreviate=T), # get weekday label
         wkdy = fct_relevel(wkdy, "Sun", "Mon", "Tue", "Wed", "Thu","Fri","Sat"), # make sure Sunday comes first
         semester = ifelse(date %in% semester_dates, 1, 0), # is date part of the semester?
         mas = ifelse(date %in% ma_dates, 1, 0), # is it an MA assignment?
         gcs = ifelse(date %in% gc_dates, 1, 0), # is it a GC assignment?
         not_here = ifelse(date %in% not_here_dates, 1, 0), # is it a day off?
         day = lubridate::mday(date), # get day of month to add later as a label
         # Below: our custom wom() function
         week = wom(date)) -> Cal

# Create a category variable, for filling.
# I can probably make this a case_when(), but this will work.

Cal %>%
  mutate(category = NA,
         category = ifelse(semester == 1, "Semester", category),
         category = ifelse(semester == 1 & wkdy %in% c("Mon", "Wed", "Fri"), "Class Day", category),
         category = ifelse(mas == 1, "MA Due", category),
         category = ifelse(gcs == 1, "GC Due", category),
         category = ifelse(not_here == 1, "Breaks", category),
         category = ifelse(is.na(category), "NA", category)) -> Cal 

Cal %>% 
  ggplot(.,aes(wkdy, week)) +
  # custom theme stuff below
  # theme_steve_web() + 
  theme_bw() +
  theme(panel.grid.major.x = element_blank()) +
  # geom_tile and facet_wrap will do all the heavy lifting
  geom_tile(alpha=0.8, aes(fill=category), color="black", size=.45) +
  facet_wrap(~mon, scales="free", ncol=3) +
  # fill in tiles to make it look more "calendary" (sic)
  geom_text(aes(label=day),family="Open Sans") +
  # put your y-axis down, flip it, and reverse it
  scale_y_reverse(breaks=NULL) +
  # manually fill scale colors to something you like...
  scale_fill_manual(values=c("Class Day"="steelblue", 
                             "Semester"="lightsteelblue",
                             "Breaks"="forestgreen",
                             "NA" = "white", # I like these whited out...
                             "MA Due"="indianred4",
                             "GC Due"="gold"),
                    #... but also suppress a label for a non-class semester day
                    breaks=c("Class Day","MA Due","GC Due","Breaks")) +
  labs(fill = "", x="", y="",
       caption = "Notable dates: Fall Break (17-18 Oct.), Thanksgiving (21-25 Nov.), Finals (14-19 Dec.)\nMA = Main Assigment; GC = Graph Challenge")  -> class_cal

```

# Course Description

Political data surround us everywhere we look. How do we use this data for maximum social impact? We start by thinking clearly with data and using a good research design. To do these things well, you need to be armed with the appropriate tools and enough conceptual understanding of them to know which to use and how. 

In this class, we'll cover the basics of research design. You'll learn about **correlation** and **causation**, how to know when one implies the other, and how to estimate them in the face of **bias** and **noise** that inevitably find their way into out data.

# Course Objectives

In this course, you'll develop a number of skills.

#### Quantitative

You will develop your quantitative skills in the treatment of data. You'll learn how data are gathered, assembled into datasets, and most effectively visualized to describe and draw inferences for maximum impact.

#### Writing

A picture may be worth a thousand words, but it helps if words are included. When you visualize data you will also write up your results. This will enhance your writing skills and reinforce the idea that data visualization is not a mechanical skill; it is a method of inquiry intended for public communication.

#### Critical Thinking

Critical thinking involves "confronting multiple, competing perspectives and adjudicating between them." This is the essence of what is required to organize data, create effective visualizations, and draw conclusions.

#### Statistical Programming

You will use R and RStudio to wrangle and visualize data. By the time you have finished this course you may not be an expert programmer, but you will be well versed in the basics of R and how to use the `{tidyverse}` suite of packages to assemble, organize, and visualize data.

# Analysis Software

Students will use R and RStudio, which are free and open source! We will primarily use the RStudio version available through a web browser installed on a Denison server: `r.denison.edu`. This can be accessed on campus on any computer or through a VPN off campus. To use the VPN:

1.  Go to myDenison.
2.  Head to MyApps.
3.  Select Remote Access.
4.  Follow the instructions to download and use the VPN.

All necessary R packages for the course are already installed in the server version. While using the **Denison server version is highly recommended**, you are also welcome to download and install local versions of R and RStudio on your laptop. Follow [this guide](https://www.rstudio.com/products/rstudio/download/).

# Required Readings

In addition to reading some short pieces written by journalists, academics, and others available online, we will primarily draw on Kieran Healy's excellent book on data viz:

> Healy, Kieran. 2019. *Data Visualization: A Practical Introduction*. Princeton: Princeton University Press.

# Resources

## We Have an TA!

Zach Broeren is the Teaching Assistant for DPR 101. If you have questions about assignments or need help troubleshooting code, Zach is an excellent resource.

**Office Hours**: 11:00 am -- 1:00 pm

**Location**: Higley 408

**Contact**: [broere_z1\@denison.edu](mailto:broere_z1@denison.edu){.email}

## Accessibility

Students with a documented disability should complete a Semester Request for Accommodations through their Accommodate MyAccommodations app on MyDenison. It is the student's responsibility to contact me privately as soon as possible to discuss specific needs and make arrangements well in advance of an evaluation. I rely on the Academic Resource Center (ARC) located in 020 Higley Hall, to verify the need for reasonable accommodations based on the documentation on file in that office. Reasonable accommodations cannot be applied retroactively and therefore ideally should be enacted early in the semester as they are not automatically carried forward from a previous term and must be requested every semester.

## Writing Center

Staffed by student Writing Consultants, the Writing Center is a free resource available to all Denison students. Writing Consultants from a range of majors work with writers one-on-one in all phases of the writing process, including (but not limited to): deciphering assignments, discussing ideas, developing an argument, integrating research and sources, working with faculty feedback, and/or polishing a draft. In addition, Consultants are happy to help with all types of writing, from lab reports, research papers, and informal writing assignments to cover letters, personal statements, and other application materials. The Center welcomes writers from all backgrounds and levels of college preparation, including multilingual writers. Should a multilingual writer need writing assistance that exceeds the abilities of consultants, the writer can be referred to the Coordinator for Multilingual Learning, Kaly Thayer ([thayerk\@denison.edu](mailto:thayerk@denison.edu){.email}). Writing Center consultations will take place in person in in the Atrium level of the Library; please visit the Writing Center's page (<https://my.denison.edu/campus-resources/writing-center>) on MyDenison for specific information regarding hours of availability and how to schedule an appointment. The Writing Center strongly recommends signing up for appointments in advance.

## Multilingual Support (L2)

Students who use English in addition to other languages are welcome to use the resources available at the Multilingual Learning Office. Kaly Thayer, the Assistant Director of Multilingual Learning, and Anna Adams, the English Language Support Specialist, as well the student consultants who work with them, are trained and experienced in helping students address the different issues that arise when working in more than one language. If English is not your first or only language, please consider utilizing this resource, which is available to ALL Denison students. Ms. Thayer, Ms. Adams, and the student consultants offer a variety of support for L2 students, including consulting with you about your written language (grammar, syntax, word-choices), strategies to manage your reading assignments, assistance with class conversation and presentations, and help devising ways to develop and effectively use all your skills in English. You can set up an appointment via MyDenison - Campus Resources - Multilingual Learning, or by emailing the Multilingual Learning Office directly at [englishhelp\@denison.edu](mailto:englishhelp@denison.edu){.email}.

## Reporting Sexual Assault

Essays, journals, and other coursework submitted for this class are generally considered confidential pursuant to the University's student record policies. However, students should be aware that University employees are required by University policy to report allegations of discrimination based on sex, gender, gender identity, gender expression, sexual orientation or pregnancy to the Title IX Coordinator or a Deputy Title IX Coordinator. This includes reporting all incidents of sexual misconduct, sexual assault and suspected abuse/neglect of a minor. Further, employees are to report these incidents that occur on campus and/or that involve students at Denison University whenever the employee becomes aware of a possible incident in the course of their employment, including via coursework or advising conversations. There are others on campus to whom you may speak in confidence, including clergy and medical staff and counselors at the Wellness Center. More information on Title IX and the University's Policy prohibiting sex discrimination, including sexual harassment, sexual misconduct, stalking and retaliation, including support resources, how to report, and prevention and education efforts, can be found at: <https://denison.edu/campus/title-ix> .

## R Resources & Cheat Sheets

[How to Google R Stuff](https://www.r-bloggers.com/2021/12/how-to-google-r-stuff/) $\Diamond$ [`{tidyr}`](https://github.com/rstudio/cheatsheets/blob/main/tidyr.pdf) $\Diamond$ [`{dplyr}`](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf) $\Diamond$ [`{ggplot2}`](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf) $\Diamond$ [`{rmarkdown}`](https://github.com/rstudio/cheatsheets/blob/main/rmarkdown-2.0.pdf) $\Diamond$ [R color cheatsheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf) $\Diamond$ [`{paletteer}`](https://emilhvitfeldt.github.io/r-color-palettes/discrete.html)

# Course Policy

The course policy and requirements are detailed below. It all basically boils down to: (1) show up to class, (2) learn some stuff, and (3) don't cheat or trick me into believing you've accomplished 2.

## Grading Policy

Grades at Denison are based on a standard 4.0 scale. You can read more about Denison's grading system [here](https://catalog.denison.edu/catalog/academic-program/grading-system-evaluation/). Generally, a 90 corresponds to an A$-$, an 80 to a B$-$, etc. The specifics will vary as I curve the scale depending on the overall class distribution.

*Attendance and Participation* **10%**

> This should go without saying, but students should show up to class and ask questions and answer them. This course works best with lots of two-way conversation between you and me. Data programming and visualization involve a lot of trial and error, but it helps to talk problems through. To make this happen, you need to show up to class. You will have **1 free absence** in this class (no questions asked---really). Beyond this first absence you can expect to lose 1/2 of your participation and attendance grade for each of 2 subsequent unexcused absences. If you do the math, that means if you have two unexcused absences (after your first freebie), you lose all of your participation and attendance points. The way to avoid **unexcused** absences (other than not missing class) is to give me advanced warning of an upcoming absence---either via email or written notice at least **24 hours in advance**. If you only come to class and never say a word, you will receive no more than 66% of your attendance and participation points (that's a D, FYI).

*4 Main Assignments* **15% Each**

> You'll have four writing assigments to complete in this course. The prompts for these assigments are located at the end of this syllabus. These involve wrangling some political data, describing trends with approximately five figures, and a five page written summary (with your figures included) of what you find.

*6 Graph Challenges(!)* **4% Each**

> These will be scattered throughout the semester, and are intended to be fun and challenging. You'll be given data or a data-frame (you'll learn what that is in the course) and make a graph that tells a story from the data better than what could be communicated by just looking at the data on its own. You'll get the necessary data for these a week in advance of when they are due.

*Post a Viz* **6%**

> This is a low stakes weekly assigment to post before class on Mondays. You're job is simple. Find a visualization out in nature, post it to Canvas with its source url, and write a paragraph describing it. Did its creators make smart display choices? Is there anything that makes it interesting? Is it just terrible? This assigment will be graded based on completion (unless it's clear that you didn't try at all).

$$
\begin{aligned}
6\% & \text{ (Post a Viz)} \\
10\%  & \text{ (Attendance + Participation)} \\
4 \times 15\% = 60\%  & \text{ (Main Assigments)} \\
\underline{+ \space 6 \times 4\% = 24\%} &  \text{ (Graph Challenges)} \\
100 \% & \text{ (Total Assigments)}
\end{aligned}
$$

## E-mail Policy

I have a simple email policy, and it is targeted at achieving one goal: **maximizing your and my work-life balance**. The policy is this:

> I promise a timely response to **relevant** emails I receive between **9:00 AM** and **5:00 PM** Monday to Friday.

You may not think professors have lives (but we in fact do, and I like to live mine outside of normal working hours). That doesn't mean I expect students to abide by chrono-normative standards.[^1] But, this does mean that if you email me outside of these windows, I will not respond until the next 9-5 workday.[^2]

[^1]: By this, I simply mean classic societal expectations about working vs. leisure hours.

[^2]: FYI, if you email me on a Saturday that you will be unable to attend class on Monday for an excusable reason, don't worry that I won't see your email until Monday. I'll be able to see when you sent me your notice.

So that explains non-responses to emails sent outside of my normal working outs; what about **irrelevant** emails? What is an irrelevant email? It's an email that asks a question that can be answered by the student without need of consulting me.

1.  Did you miss class and want to know what we covered? ***Read the syllabus***.
2.  Do you want to know how much of your total grade an assignment is worth? ***Read the syllabus***.
3.  Are you curious about a course policy? ***Read the syllabus***.
4.  Are you curious about a campus-wide policy? [***Consult Denison's academic policies***](https://denison.edu/academics/curriculum/academic-policies).
5.  Did you miss class, or are you going to miss class, and do you want to notify me of why? ***Read the attendance policy in the syllabus***. You have 1 free absence baked into your grade. I don't need or want an explanation for this first absence. It's yours to use as needed or desired.
6.  Did you miss class for the second (or more) time and you want to retroactively get that absence excused? ***You should have notified me prior to the absence***. Barring the most extreme of emergencies,[^3] most excusable absences have advanced notice for you. It's your responsibility to give me advanced notice, too. Notice of an excusable absence after the fact is too late. You may give me your advanced notice up to 24 hours in advance of class via email or in writing. **A verbal notice will not count.** I'll inevitably forget that you told me, and there will be nothing in writing to prove that you notified me.
7.  If a student would like to protest a grade on an assignment but fails to reference a specific point of objection, this does not require a response from me. As is noted in Denison's [Academic Grievance Policy](https://denison.edu/academics/curriculum/academic-policies), "Simple disagreement with the instructor's philosophy in assigning grades, or with the instructor's professional judgment about the quality of the work does not constitute a cause for petitioning." If you would like to improve your understanding of course material, or would like to preempt a poor grade, you are strongly encouraged to visit me during my office hours.
8.  Do you want to know if you can have an extension on an assignment? ***The answer will always be no***. There's too much subjective discretion involved on my part in deciding whether someone deserves an extension. Since there's no objective or consistent way of doing this, I won't do it at all.
9.  Do you want to know if you can have your grade rounded up? ***The answer will always be no***.[^4]
10. Do you want extra credit, which amounts to asking the professor to do more work? ***The answer will always be no***.

[^3]: Medical emergencies, for example, would fall into this category. Note, however, if you have to be absent for an emergency, and it is your first absence, **you do NOT need to notify me**.

[^4]: ["Grade grubbing"](https://marktomforde.com/academic/undergraduates/GradeGrubbing.html) is a bad look.

## Make-Up Assigment Policy

There are **NO** make-ups for missed assigments. Don't bother asking. But, if you anticipate missing an exam and notify me ***in advance***, we can meet to work out a solution. "I'm really busy" is not a valid excuse. There's a lot of work to do in this course, but it is spaced out so that not all of this work is crammed into a single week. If you have a lot of assignments coming due for other classes, **PLAN AHEAD**.

## Attendance Policy

Come to class. My rule is that you get only 1 free unexcused absence. Beyond this, you forfeit 50% of your attendance and participation points for each of up to two subsequent unexcused absences. If you do the math, that means three unexcused absences will result in a total loss of your attendance grade. If you attend class, but never participate in class discussion, expect to earn no more than 66% of your attendance and participation points.

## Computer-based Excuses

Excuses for late or missed assignments based on CD, flash drive, or hard drive errors are **not acceptable**. The Denison network and server is reliable and accessible. If you use your Google Drive and the RStudio server, all your work will be backed up and easy to access.

## Late Assigments

We have a lot of work to do in this class. So turn in your work when it's due. This is meant to help you. I love to procrastinate just as much as anyone else---but if you procrastinate in this course, you will drown. As incentive for keeping on top of your assignments, each day your assignment is late and unexcused (including weekends and holidays) you will lose 5 percent from your final grade for that assignment. There will be no exceptions made for work that is submitted only minutes after a deadline. If something is due by midnight on a Friday (12:00 AM), the moment the clock strikes 12:01 AM your assignment is a day late and you will automatically lose 5% from your grade.

## Electronic Submission

You will submit all of your assignments electronically via Canvas.

## Academic Dishonesty Policy

Don't cheat. Just don't do it.

It should go without saying, but ***plagiarism*** is also a form of cheating and it includes:

1.  Copying or paraphrasing the ideas of others without citation or attribution.
2.  Copying or paraphrasing the ideas of ***other students in the class***.

I've had to deal with students plagiarizing before. It's painful for me and it puts a blight on the record of the student. It's not only cheating, it's stealing.

When in doubt about whether something constitutes cheating, consult Denison's [Code of Academic Integrity](https://denison.edu/academics/curriculum/integrity).[^5] Be advised that this same Code of Academic Integrity requires that instructors notify the Associate Provost of cases of academic dishonesty. **Any incidence of academic dishonesty will result in failure of the course and referral to the Denison judicial process.**

[^5]: Of course, if you have to ask yourself if something counts as cheating, then it probably is...

\newpage

# Class Schedule[^6]

[^6]: As the instructor, I retain the right to amend this schedule if the need arises.

\hrulefill

```{r calendar, echo=F, eval=T,  warning=F, fig.width=8, fig.height=6}

class_cal + 
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    legend.position = 'top'
  )
ggsave(
  here::here('schedule.png'),
  height = 6,
  width = 8
)
```

```{=tex}
\singlespacing
\setlength{\parskip}{0pt}
```
\bigskip

\hrulefill

# Getting Started

## `r advdate(mon, 1)`: Intro to Data Viz

#### M

Syllabus day and course intro.

#### W

Read: Healy Ch. 1 + Tufte [Ch. 1](https://drive.google.com/file/d/1-3OdDZ-HfQEmCHeCzKVZbl99cyd8-QCG/view?usp=sharing) & [Ch. 2](https://drive.google.com/file/d/1-49qU-zwzv766fLK376EgSbOwgiZI2Bf/view?usp=sharing). In-class: We'll talk about what makes a good data visualization.

#### F

Read: [Eight graphs that tell the story of U.S. economic inequality](https://equitablegrowth.org/eight-graphs-that-tell-the-story-of-u-s-economic-inequality/); [Congressional polarization](https://www.vox.com/2015/4/23/8485443/polarization-congress-visualization); [Polarization across 9 countries](https://www.vox.com/2020/1/24/21076232/polarization-america-international-party-political). In-class: Visit from our TA + intro to RStudio.

## `r advdate(mon, 2)`: R + RStudio

#### M

Read: Healy Ch. 2.; [Top 50 ggplot2 visualizations](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html). In-class: RMarkdown, packages, and reading data.

#### W

Read: [Social lobying](https://drive.google.com/file/d/1-A7ivDIb2KWuO5f7G5inMo3B4OgXPePs/view?usp=sharing); [QAnnon beliefs](https://religioninpublic.blog/2021/09/17/what-is-happening-with-qanon-beliefs-post-trump/); [Aid skeptical](https://religioninpublic.blog/2021/01/21/white-conservative-religious-and-aid-skeptical/). In-class: The nitty gritty details of knitting documents.

#### F

Read: [Religion and culture wars](https://religioninpublic.blog/2021/06/21/religion-and-the-culture-wars-a-re-examination/); [Companies punishing Rs for Jan. 6](https://drive.google.com/file/d/1-Et2FXCkNpsXYnQZ5Nsq-jFRInSYbdY5/view?usp=sharing). Do: Follow prompt on Canvas for editing and knitting your first R Markdown file. In-class: Make your first figure!

# Unit 1 --- Cross-sectional Data

## `r advdate(mon, 3)`: Making Plots

#### M

Read: Healy Ch. 3. In-class: How ggplot works + tidy data.

#### W

In-class: Mappings and layers. Do: Tidy data task on Canvas.

#### F

In-class: More on mapping aesthetics + saving your work.

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{First graph challenge due on Canvas Sunday 09.18 by midnight.}}

\end{center}
```
\bigskip

## `r advdate(mon, 4)`: Displaying Multiple Variables

#### M

In-class: Exploring state-level data with joins and filters.

#### W

Read: Healy Ch. 4; [this](https://medium.com/nightingale/how-to-choose-the-colors-for-your-data-visualizations-50b2557fa335), [this](https://phys.org/news/2021-08-perception-scientific-results-distorted-graphics.html), and [this](https://www.visualisingdata.com/2015/01/make-grey-best-friend/) on color in figures. In-class: Visualizing state-level data with colors and different data types.

#### F

Read: [R color cheatsheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf), check out [this](https://coolors.co/) and [this](https://emilhvitfeldt.github.io/r-color-palettes/discrete.html) resource for generating palettes (and see [this one](https://www.color-blindness.com/coblis-color-blindness-simulator/) for accommodating colorblindness). Do: Use the palette picker to bring to class FOUR palettes picked out --- 1) one five value qualitative palette, 2) one divergent palette, 3) one sequential (i.e., low to high), and 4) one for just two groups (just pick two colors manually from the R color cheatsheet that would work well).

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{Second graph challenge due on Canvas Sunday 09.25 by midnight.}}

\end{center}
```
\bigskip

## `r advdate(mon, 5)`: Maps & Small Multiples

#### M

Read: Healy Ch. 7. In-class: Plotting data on maps.

#### W

In-class: [Small multiples](https://towardsdatascience.com/how-to-make-beautiful-small-multiple-us-maps-in-r-ad7e557cd463) and [`{geofacets}`](https://cran.r-project.org/web/packages/geofacet/vignettes/geofacet.html). See a fun example [here](https://www.axios.com/2017/12/15/the-emoji-states-of-america-1513302318).

#### F

In-class: Work day.

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{Third graph challenge due on Canvas Sunday 10.02 by midnight.}}

\end{center}
```
\bigskip

# Unit 2 --- Survey Data

## `r advdate(mon, 6)`: Survey Data + Reading & Manipulating

#### M

Read: Survey codebook and design statement (tbd); also read about survey research in Johnson and Reynolds (tbd -- in Canvas readings). In-class: Intro to using [`{socsci}`](https://github.com/ryanburge/socsci).

#### W

Read: [Armaly, Buckley and Enders](https://drive.google.com/file/d/1-F-ZFpWGYATCrRfrReyknpAGbPQ_xm9u/view?usp=sharing), paying special attention to the "Data and Measurement" section. In-class: Making indices and showing them.

#### F

Read: [Weitz-Shapiro and Winters](https://drive.google.com/file/d/1-FOMTh8WFmq3FPRymmPACqP2jPb5EdfI/view?usp=sharing). In-class: More on indices.

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{MA1 due on Canvas Friday 10.07 by midnight.}}

\end{center}
```
\bigskip

## `r advdate(mon, 7)`: Displaying 1 Variable + "Graph Tables"

#### M

Read: Healy Ch. 5. In-class: Making "graph tables" with `{socsci}`.

#### W

Read: Healy Ch. 8. In-class: Refining your plots.

#### F

Work Day!

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{Fourth graph challenge due on Canvas Sunday 10.16 by midnight.}}

\end{center}
```
\bigskip

## `r advdate(mon, 8)`: Graph Finishing

#### M

**No Class.** Fall break!

#### W

Work Day!

#### F

Work Day!

# Unit 3 --- Longitudinal Data

## `r advdate(mon, 9)`: Temporal Data, Gathering, & Aggregating

#### M

Read: [Can we trust measures of democracy?](https://www.washingtonpost.com/news/monkey-cage/wp/2015/06/23/how-do-you-measure-democracy/) And read about [7 different types](https://humansofdata.atlan.com/2016/11/visualizing-time-series-data/) of temporal data visualizations. In-class: Using [`{democracyData}`](https://github.com/xmarquez/democracyData) to explore tends in democracy.

#### W

Read: Temporal visualizations of [employment](https://www.economist.com/graphic-detail/2021/08/20/so-far-ending-pandemic-unemployment-aid-has-not-yielded-extra-jobs) and [minimum wage](https://www.abc.net.au/news/2021-10-13/nobel-prize-in-economics-2021-david-card-minimum-wage/100531994). In-class: Shifting between time and groups, and aggregating and adjusting for baselines.

#### F

**No Class.** I will be at a conference.

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{MA2 due on Canvas Friday 10.28 by midnight.}}

\end{center}
```
\bigskip

## `r advdate(mon, 10)`: Manipulating & Joining

#### M

In-class: Add demography for countries using [`{pwt10}`](https://www.rug.nl/ggdc/productivity/pwt/?lang=en).

#### W

In-class: Labeling in time.

#### F

In-class: Choosing case studies.

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{Fifth graph challenge due on Canvas Sunday 11.06 by midnight.}}

\end{center}
```
\bigskip

## `r advdate(mon, 11)`: Visualizing Multiple Trends

#### M

In-class: Groups and multiple facets.

#### W

In-class: Groups and multiple facets continued.

#### F

In-class: Animated plots for multiple trends with [`{gganimate}`](https://gganimate.com/).

# Unit 4 --- Choose Your Own Adventure

## `r advdate(mon, 12)`: Finding & Pitching

#### M

In-class: Review data available --- I'll provide a list.

#### W

In-class: Pitch your project in 5 min.

#### F

In-class: Pitch your project in 5 min. continued.

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{MA3 due on Canvas Friday 11.18 by midnight.}}

\end{center}
```
\bigskip

## `r advdate(mon, 13)`: Thanksgiving

***No Class (Thanksgiving)*** If you feel like it, spend this time either working ahead and/or reviewing past material. But by all means, **take a break, too!**

## `r advdate(mon, 14)`: Descriptives & Explanations.

#### M

Lab Work.

#### W

Lab Work.

#### F

Lab Work + MA4 presentations.

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{Sixth graph challenge due on Canvas Sunday 12.04 by midnight.}}

\end{center}
```
\bigskip

## `r advdate(mon, 15)`: Descriptives & Explanations Cont.

#### M

Lab Work

#### W

Lab Work

#### F

Lab Work

## `r advdate(mon, 16)`: Finals Week

#### M

Reflections and course evaluations.

\bigskip

```{=tex}
\begin{center}

\fbox{\textbf{MA4 due on Canvas Thursday 12.15 by 11:00 am.}}

\end{center}
```
\bigskip

\clearpage

# Quick Description of Main Assignments.

#### MA1

We often have data organized at higher administrative units than the individual, such as the state/nation or we need to have the data aggregated to these useful units. Part of the struggle is wrangling the data to attach to these units, which might include having the appropriate identifiers. One great puzzle of American politics is that women are 51 percent of the population and constitute less than half that amount of elected representatives. Let's draw on the American laboratories of democracy (the states) and get data about women in state legislatures (WISL). Let's also compare how the US fares compared to other countries -- is it distinctive or is this common for industrialized, developed countries? We'll explore a variety of explanations, including partisanship, institutions, and attitudes to help explain the variation in WISL. We'll at least use bar graphs, maps, scatterplots. Your write up should be about 5 pages and include about 5 figures. Due Friday 10.07 at Midnight.

#### MA2

One of the core public problems of our time is racism, and racial resentment has been thought to drive a considerable amount of our politics. We can infer this from policy decisions pursuing disenfranchisement, but it's optimal to measure it directly. Using the 2018 CCES, you will display how much using their battery of measures: how much of each item, how much in total, how much racial resentment by party affiliation and some other demographic measures. We can also link that at the state-level to policy changes about the electoral process. Then write words to help readers understand what is shown and what it means. The write up should be about 5 pages to accompany your \~5 figures. We'll show bar graphs, dumbbell plots, and a ridgeline (and/or histogram). Due Friday 10.28 at Midnight.

#### MA3

Many policymakers and academics have been sounding the alarm about democratic backsliding in the United States and several countries that have recently transitioned to democracy. But the severity (and existence!) of the problem depends on how you choose to measure democracy. Has democracy really declined in places like the United States and in countries in Eastern Europe (and beyond)? Draw conclusions about trends in democracy over time using different indices provided in `{democracyData}`, and use country-level variables provided in `{pwt10}` to filter to carefully chosen comparisons between countries. The write up should be about 5 pages and include \~5 figures. Due Friday 11.18 at Midnight.

#### MA4

You will draw from Denison survey data for this choose-your-own-adventure assignment. The advantage of survey data is that they allow so many more narratives to develop than any other data source. There is little central direction on this assignment, but you will be asked to provide a pitch for your chosen investigation. As with the other papers, this should be about 5 pages with somewhere in the neighborhood of 5 figures. This will be due as your final assignment Dec. 15 by 11:00 am.
