---
title: "DPR 201 Class Notes"
format: html
---

Here's a Quarto file for you to take class notes. It includes headers that correspond to different subjects we'll cover. This is a great place to try out example code in the lecture notes and to follow along with code-alongs in class.

## Looking at Data

```{r}
library(tidyverse)

usaidhr <- read_csv(
  "https://tinyurl.com/usaidhr"
)
```


```{r}
glimpse(usaidhr)
```


### scatter plots

```{r}
ggplot(usaidhr) +
  aes(x = physint, y = aid) +
  # geom_point() +
  geom_smooth(
    method = "lm",
    se = F
  ) +
  labs(
    title = "Here's a title",
    subtitle = "A subtitle",
    x = "Physical Integrity",
    y = "U.S. Aid",
    caption = "Data for 1945-2023"
  )
```

```{r}
usaidhr |>
  filter(
    aid == max(aid)
  )
```

### Line plot

```{r}
ggplot(usaidhr) +
  aes(x = year, aid) +
  geom_smooth()
```

```{r}
ggplot(usaidhr) +
  aes(x = year, y = aid) +
  stat_summary(
    geom = "line",
    fun = sum
  )
```

### column plot

```{r}
ggplot(usaidhr) +
  aes(x = aid, y = reorder(leader, -year)) +
  geom_col()
```


## Modeling Data with Linear Regression


## Quantifying the Role of Chance


## From Relationships to Predictions


## Picking the Best Forecast


## From Relationships to Causation


## Randomized Trials


## Selection on Observables


## Regression Discontinuity


## Difference-in-Differences
