---
title: "AC3 Solution"
format: pdf
editor: visual
---

## Get data

First I need to get the data:

```{r}
## open {tidyverse}
library(tidyverse)

## read in data
read_csv(
  "https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20201/Data/prez_data.csv"
) -> Data
```

With this data I need to do three things:

1.  Create a new outcome variable that is the incumbent's vote margin adjusted to only be the margin between the Democratic and Republican candidates.
2.  Make an election forecast and use bootstrapping to get a standard error and confidence intervals for my prediction.
3.  Determine whether Biden is projected to do statistically better than his Republican challenger in 2024.

### Make new outcome

To make a measure of the incumbent's vote margin adjusted to only be fore the Republican and Democratic margin, I need to take the incumbent vote and divide it by the sum of the Democratic and Republican vote and subtract 0.5.

```{r}
## read data into mutate()
Data |>
  mutate(
    ## add new column to the data called inc_margin
    inc_margin = inc_vote / (dem + rep) - 0.5
  ) -> Data ## save
```

### Estimate forecast model and bootstrap

Next I need to construct my forecast model.

```{r}
## fit the model with the data
lm(
  inc_margin ~ third_party + inc_running + inc_approval +
    gdp_pct_growth,
  data = Data
) -> forecast_model

## check out its summary
summary(forecast_model)
```

With this model, I'll make a new forecast for 2024:

```{r}
tibble(
  third_party = 0,
  inc_running = 1,
  inc_approval = 0.41 - 0.54,
  gdp_pct_growth = 0.03
) -> forecast_data

forecast_data |>
  mutate(
    inc_margin_fit = predict(forecast_model, forecast_data)
  ) -> forecast_data
```

Now do bootstrapping.

```{r}
tibble(
  its = 1:1000,
  boot_fit = map(
    its,
    ~ {
      ## bootstrap the data
      Data |>
        sample_n(
          size = nrow(Data),
          replace = T
        ) -> boot_Data
      
      ## fit the model to the bootstrapped data
      lm(
        inc_margin ~ third_party + inc_running + inc_approval +
          gdp_pct_growth,
        data = boot_Data
      ) -> boot_model
      
      ## get a bootstrapped prediction
      predict(
        boot_model,
        forecast_data
      ) -> boot_fit
      
      ## return
      boot_fit
    }
  )
) |>
  unnest(
    boot_fit
  ) -> boot_fits
```

Now add std. error and CIs. It looks like Biden's chances are not statistically better than his opponents.

```{r}
forecast_data |>
  mutate(
    std_error = sd(boot_fits$boot_fit),
    conf_lo = inc_margin_fit - 1.96 * std_error,
    conf_hi = inc_margin_fit + 1.96 * std_error
  ) -> forecast_data

## report
forecast_data |>
  select(inc_margin_fit:conf_hi) |>
  kableExtra::kbl(
    digits = 3
  )
```
