---
title: "Analysis Challenge 6"
subtitle: "Confounding"
author: "Miles D. Williams"
date: '2023-03-28'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      message = F,
                      dpi = 500,
                      out.width = "50%")
library(tidyverse)
library(ggdag)
theme_set(theme_void())
```

For this analysis challenge, you'll be presented with a few different causal frameworks. Your task, for each case, is to use that framework to decide which variables you should control for in a multiple regression model, and which you should leave out.

For each scenario you will 

1. Indicate the variables you will control for
2. Justify your choice in a short paragraph

For two, your reasoning should be guided by one of the following considerations we discussed in the lecture regarding *selection on observables* or SOO designs:

- Minimizing ommited variable bias
- Avoiding controlling for mechanisms
- Avoiding post-treatment bias

\clearpage

## Secnario 1

```{r}
dagify(
  Outcome ~ Treatment + Z1 + Z2,
  Treatment ~ Z1 + Z2
) %>%
  ggdag(
    node = F,
    text_col = "Black"
  )
```

Variables you'd control for:

Reasoning:

\clearpage

## Scenario 2

```{r}
dagify(
  Outcome ~ Treatment + Z1 ,
  Treatment ~ Z1,
  Z2 ~ Treatment
) %>%
  ggdag(
    node = F,
    text_col = "Black"
  )
```

Variables you'd control for:

Reasoning:

\clearpage

## Scenario 3

```{r}
dagify(
  Outcome ~ Treatment,
  Treatment ~ Z1 + Z2
) %>%
  ggdag(
    node = F,
    text_col = "Black"
  )
```

Variables you'd control for:

Reasoning:


\clearpage

## Scenario 4

```{r}
dagify(
  Outcome ~ Treatment + Z1 + Z2
) %>%
  ggdag(
    node = F,
    text_col = "Black"
  )
```

Variables you'd control for:

Reasoning:


\clearpage

## Scenario 5

```{r}
dagify(
  Outcome ~ Z1 + Z2,
  Z1 ~ Treatment,
  Z2 ~ Treatment
) %>%
  ggdag(
    node = F,
    text_col = "Black"
  )
```

Variables you'd control for:

Reasoning:


\clearpage

## Scenario 6

```{r}
dagify(
  Outcome ~ Treatment + Z1 + Z2,
  Z1 ~ Treatment,
  Z2 ~ Treatment
) %>%
  ggdag(
    node = F,
    text_col = "Black"
  )
```

Variables you'd control for:

Reasoning: