---
title: "Untitled"
author: "Miles D. Williams"
date: '2023-04-04'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(estimatr)
library(texreg)
coolorrr::set_theme()
```


Okay, let's see if I can make any sense of this. Here's the data:
```{r}
url <- "https://raw.githubusercontent.com/AndrewBertoli/Nationalism-RD-Replication-Files/master/WorldCupData.csv"
data <- read_csv(url)
```

What's this look like?
```{r}
ggplot(data) +
  aes(x = PointsFromCutpoint,
      y = Treat) +
  geom_point()
```

Okay, so that's the discontinuity. Outcome should be aggression.
```{r}
## From the original code
data$AGGAfter=with(data,  AGGP.5 + AGGP1 + AGGP1.5 + AGGP2 + AGGP2.5)

data$AGGBefore=with(data, AGG0 + AGGM.5 + AGGM1 + AGGM1.5 + AGGM2)
```

This format isn't exactly tidy:
```{r}
ggplot(data) +
  aes(x = PointsFromCutpoint,
      y = AGGAfter) +
  geom_point()
```

Or maybe it is. Actually, yeah this makes sense! So I should be able to easily construct a RDD like so:

```{r}
rdd_fit <- 
  lm_robust(
    AGGAfter ~ Treat * PointsFromCutpoint,
    data = data
  )
screenreg(
  rdd_fit,
  include.ci = F,
  digits = 3
)
```

Conflict before the cup could be used for a placebo test:
```{r}
rdd_fit2 <- 
  lm_robust(
    AGGAfter ~ Treat * PointsFromCutpoint,
    data = data %>% filter(Use==1)
  )
rdd_fit3 <- 
  lm_robust(
    AGGBefore ~ Treat * PointsFromCutpoint,
    data = data
  )
screenreg(
  list(main = rdd_fit, 
       trimmed = rdd_fit2,
       placebo = rdd_fit3),
  include.ci = F,
  digits = 3
)
```


Some better figures:
```{r}
data %>%
  mutate(
    diff = AGGAfter - AGGBefore
  ) %>%
  group_by(diff, PointsFromCutpoint, Treat) %>%
  summarize(
    n = n()
  ) %>%
  ungroup() %>%
  mutate(
    pred = fitted(
      lm(diff ~ PointsFromCutpoint * Treat,
         weights = n)
    )
  ) %>%
  ggplot() +
  aes(x = PointsFromCutpoint,
      y = diff) +
  geom_point(
    aes(size = n),
    size = 1
  ) +
  geom_line(
    aes(y = pred),
    color = "blue"
  )
```

```{r}
ggplot(data) +
  aes(
    x = PointsFromCutpoint,
    y = AGGAfter - AGGBefore,
    group = Treat
  ) +
  geom_smooth(
    method = lm,
    level = 0.83
  ) +
  geom_vline(
    xintercept = 0,
    lty = 2
  ) +
  labs(
    x = "Points from Cutpoint",
    y = "Frequency of MID Initiation",
    title = "Making it to the World Cup and Aggression",
    caption = "Data: Bertoli (2017), ISQ"
  )
```


## With cleaner data

```{r}
clean_data <- 
  data %>%
    transmute(
      ## Outcome data:
      pre_agg = AGGBefore,
        # No. MIDs post qualification +2 years
      post_agg = AGGAfter,
        # No. MIDs pre qualification -2 years
      change_agg = post_agg - pre_agg,
        # The post-pre difference
      
      ## Treatment and running variable
      treatment = Treat,
        # 1 = went to the World Cup
      cutpoint_margin = PointsFromCutpoint,
        # Point margin for threshold necessary to qualify
      
      ## Sample
      use = Use,
      
      ## Covariates
      military_spending = Milex,
      total_pop = Tpop,
      urban_pop = Upop,
      us_ally = USAlly,
      entrance_year = EntranceYear,
      imports = Imports,
      exports = Exports,
      ongoing_civwar = EngagedCivilWar,
      soccer_popular = SoccerMostPopular,
      infant_mortality = InfantMortality,
      life_expectancy = LifeExpectancy,
      democracy = Democracy,
    ) 
clean_data
```

Visualize the regression discontinuity with geom_smooth. Set `method = lm` and `level = 0.83`, and inside `aes()` map the group aesthetic to treatment status.
```{r}
ggplot(clean_data) +
  aes(x = cutpoint_margin, y = change_agg) +
  geom_smooth(
    aes(group = treatment),
    method = lm,
    level = 0.83
  ) +
  geom_vline(
    xintercept = 0,
    lty = 2
  ) +
  labs(
    x = "Point Margin to Qualify",
    y = "Change in MID Initiation",
    title = "World Cup Qualification and Military Aggression"
  )
```

Do it again, but this time use `method = "loess"`:
```{r}
ggplot(clean_data) +
  aes(x = cutpoint_margin, y = change_agg) +
  geom_smooth(
    aes(group = treatment),
    method = "loess",
    level = 0.83
  ) +
  geom_vline(
    xintercept = 0,
    lty = 2
  ) +
  labs(
    x = "Point Margin to Qualify",
    y = "Change in MID Initiation",
    title = "World Cup Qualification and Military Aggression"
  )
```

Do these ways of analyzing the data provide the same or different inferences about the effect of the treatment?

Some researchers recommend an alternative to measuring the LATE using the conventional regression approach. They argue that if we narrow the bandwidth enough and can assume treatment is truly as-if random within this bandwidth, we can analyze the data just as if it were generated in a randomized experiment.

Narrow the data down to those within the +/-2 point range and do a simple difference in means test:
```{r}
difference_in_means(
  change_agg ~ treatment, 
  data = clean_data %>% 
    filter(between(cutpoint_margin, -2, 2))
) %>%
  screenreg()
```

Is this sufficient to make apples-to-apples comparisons?
```{r}
samp_data <- clean_data %>%
  filter(between(cutpoint_margin, -2, 2))
cov_names <- names(samp_data)[7:18]
bal_fits <- lm_robust(
  as.matrix(samp_data[, cov_names]) ~ treatment,
  data = samp_data
)
bal_fits %>%
  tidy() %>%
  filter(term == "treatment") %>%
  ggplot() +
  aes(
    x = statistic,
    xmin = statistic - 1.96,
    xmax = statistic + 1.96,
    y = outcome
  ) +
  geom_pointrange() +
  geom_vline(
    xintercept = c(-2, 0, 2), 
    lty = c(1, 2, 1),
    color = c("darkblue", "black", "darkblue")
  )
```

## To Do

- save the cleaned data and make available to students:

```{r}
write_csv(
  clean_data,
  here::here("DPR 201", "Data", "world_cup.csv")
)
```

- make MA prompt 