---
title: "Analysis Challenge 8"
subtitle: "Difference-in-Differences"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using the Congressional data below ([read more about the data](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20201/Data/Files%20for%20Data%20Exercises/13_README.txt)), compare the estimates of the effect of Republican party membership on conservative voting behavior using three different designs:

1. Selection on observables or SOO
2. Regression discontinuity or RD
3. Difference-in-differences or DD

Here's the code to read in the data and open the packages you'll need:

```{r}
library(tidyverse)
library(estimatr)
congress_dt <- read_csv("https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20201/Data/CongressionalData.csv")
```

The outcome variable is `cvp`. This stands for "conservative vote probability" and is a score given to each member of Congress in each term summarizing their roll-call voting. It measures the probability that each member voted conservatively across all bills relative to the median member in that session. That means it is a *relative* probability and can take either positive or negative values. Positive means that a member voted more conservatively than the median member of Congress in a given term.

The causal variable of interest is `republican` which is 1 if the member was a Republican and 0 if they were a Democrat.

For the SOO design, you'll estimate a multiple regression model that controls for `rep_twoparty_pres`. This variable is the average Republican two-party vote share for US President in congressional districts from 2004 and 2008. You can think of it as a proxy (indirect measure) of the political ideology of a district.

```{r}
soo_fit <- lm_robust(
  cvp ~ republican + rep_twoparty_pres,
  data = congress_dt
)
```


For the RD design, you'll use `rep_voteshare` as your running variable. Note that you'll need to adjust for the fact that the discontinuity is at the 0.5 vote share value. 

```{r}
rdd_fit <- lm_robust(
  cvp ~ republican * I(rep_voteshare - 0.5),
  data = congress_dt
)
```


Finally, for the DD design, you can follow the example from our Monday lecture notes.

```{r}
dnd_fit <- lm_robust(
  cvp ~ republican,
  fixed_effects = ~ cong + state_dist,
  data = congress_dt
)
```


After you estimate all three designs, compare your estimates of the effect of party membership on member voting to answer the question: *Do these research designs provide similar or very different estimates?*

```{r}
library(texreg)
screenreg(
  list(
    "SOO" = soo_fit,
    "RD" = rdd_fit,
    "D-in-D" = dnd_fit
  ),
  include.ci = F,
  custom.coef.map = list(
    "republican" = "Republican"
  )
)
```


In providing your answer make sure that you fully and accurately interpret the regression coefficient for `republican` and the associated p-value. **1 extra bonus point if you compare the results using a coefficient plot!**

```{r}
bind_rows(
  tidy(soo_fit),
  tidy(rdd_fit),
  tidy(dnd_fit)
) %>%
  filter(term == "republican") %>%
  ggplot() +
  aes(x = )
```


Upload a knitted .Rmd file to Canvas that includes your code and comparison of the results.
