---
title: "Main Assignment 1"
author: "[YOUR NAME HERE]"
date: '`r Sys.Date()`'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")
library(tidyverse)
library(socviz)
county_map_data <- left_join(county_map, county_data)
county_map_data <- county_map_data %>%
  mutate(
    change_dem = per_dem_2016 - per_dem_2012,
    change_gop = per_gop_2016 - per_gop_2012
  )

# STOP! EDIT THIS TO PICK YOUR STATE
data_for_state <- county_map_data %>%
  filter(state == "AL")
```

## Introduction

What were the characteristics of counties in [US STATE] that were correlated with shifts in support from the [REPUBLICANS/DEMOCRATS] in 2012 to [DEMOCRATS/REPUBLICANS] in 2016 for the office of US President? Finding an answer to this question matters because...


## Data and Design

I used county-level data from the `{socviz}` R package to examine how demographic characteristics of counties in [US STATE] correlated with shifts in support away from the [REPUBLICANS/DEMOCRATS] to the [DEMOCRATS/REPUBLICANS] between 2012 and 2016. Change in support was measured as the difference in the share of the vote that went to the [REPUBLICAN/DEMOCRATIC] candidate in 2016 relative to 2012.

I specifically looked at the following four additional variables to see how well they explained shifts in support:

1. VARIABLE 1
2. VARIABLE 2
3. VARIABLE 3
4. VARIABLE 4

I chose to look at these because...

To provide a sense for where the biggest shifts in electoral support occurred in [US STATE], I first drew a map of the state and used color to show how much support changed in 2016 relative to 2012 in favor of the [REPUBLICAN/DEMOCRATIC] nominee for President in 2016. I then used a series of [BAR PLOTS/BOXPLOTS/SMOOTHED REGRESSION PLOTS] to show the correlation between shifts in support and the four demographic characteristics listed above.


## Results

Here's a map of [US STATE]. It shows...

```{r}
# write your code here to plot a state map
```


Write text about your next plot here...

```{r}
# code for your second figure
```


Write text about your next plot here...

```{r}
# code for your third figure
```


Write text about your next plot here...

```{r}
# code for your fourth figure
```


Write text about your next plot here...

```{r}
# code for your fifth figure
```


## Conclusion

[Restate the research question in your own words. Restate why you think this issue is important. Summarize what you found.]