---
title: "Main Assignment 2"
subtitle: "Conditions for Civil War: Democracy, the Economy, or Leaders?"
author: "[YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")

## Packages you need
library(tidyverse)
library(geomtextpath)
library(peacesciencer)
library(coolorrr)
set_theme() # You can turn this off and select your own theme!

## Preping the data
create_stateyears(
  system = 'gw',
  subset_years = 1946:2015
) %>%
  add_ucdp_acd(
    type = "intrastate"
  ) %>%
  add_sdp_gdp() %>%
  add_democracy() %>%
  add_archigos() -> Data
left_join(
  Data,
  archigos %>% mutate(jan1obsid = obsid),
  by = c("gwcode", "jan1obsid")
) %>%
  mutate(
    age = year - yrborn
  ) -> Data
Data %>%
  group_by(gwcode) %>%
  mutate(anywars = any(ucdpongoing)+0,
         nwars = sum(ucdpongoing)) %>%
  ungroup -> Data
stand <- function(x) (x - min(x, na.rm=T)) /
  (max(x, na.rm = T) - min(x, na.rm = T))
Data %>%
  mutate(
    across(c(polity2, v2x_polyarchy, xm_qudsest),
           stand)
  ) -> Data

## This is where you come in: Pick a country to look at!
## Ideally, you should pick a country that gives you some
## variation in conflict. At least 10 years at war is 
## a good place to start. The nwars column can help you
## narrow down your choices
country <- "Guatemala" # Here's an example
Data %>%
  filter(statename == country) -> countryData
```

## Introduction

Does quality of democracy (or lack thereof) explain why countries fall into civil war? In this paper, I look at the case of [COUNTRY NAME] to see how periods of conflict correlate with different measures of quality of democracy. I also examine alternative explanations---specifically, economic wellbeing and political leader characteristics. 

Understanding the conditions under which conflict is likely to errupt matters because...

## Data and Design

For this study I'm using data made available in the `{peacesciencer}` R package. My main outcome is whether there is an ongoing civil war in [COUNTRY] during a given year from [START YEAR] to [END YEAR].

I cross-reference the incidence of civil war with two popular measures of democracy from Polity and V-Dem. Polity measures democratic quality as [LOOK UP THE POLITY DATASET AND DESCRIBE HOW IT DEFINES DEMOCRACY]. V-Dem conversely measures democratic quality as [LOOK UP THE V-DEM DATASET AND DESCRIBE HOW IT DEFINES DEMOCRACY].

I also cross-reference the incidence of civil war with [THE ECONOMIC VARIABLE YOU'LL USE] and [THE LEADER VARIABLE YOU'LL USE]. These variables may provide an alternative explanation for why countries fall into civil war.

## Analysis: Democracy and Conflict

Write about your first figure here...

```{r}
## Use two geom_statsmooth() layers to show the global trend
## in democracy using v2x_polyarchy alongside the trend for
## the country you picked to look at.
```

Write about your second figure here...

```{r}
## Use two calls to geom_textsmooth() to compare how the relationship
## between v2x_polyarchy and ongoing conflict looks vs. polity2 and 
## ongoing conflict.
```

## Alternative Explanations

Write about your third figure here...

```{r}
## Pick an economic variable in the data and show its relationship
## with ongoing conflict in your chosen country.
```

Write about your fourth figure here...

```{r}
## Pick a leader-specific variable (like age or transition type) and
## show its relationship with ongoing conflict.
```


## Conclusion

Restate the research question in your own words. Restate why you think this issue is important. Summarize what you found.