---
output: 
  stevetemplates::syllabus:
    latex_engine: xelatex
    dev: cairo_pdf
geometry: margin=1in

title: "DPR 190: War, Wealth, and World Politics"
author: Dr. Miles D. Williams
date: "Fall 2026"

email: "williamsmd@denison.edu"
web: "milesdwilliams15.github.io"
officehours: "MWF 8:30 am - 9:20 am"
office: "Knobel 101E" 
classroom: "Fellows 205"
classhours: "MWF 1:30-2:20 pm"

fontfamily: mathpazo
fontsize: 11pt
header-includes:
   - \linespread{1.05}
   - \usepackage{amsmath}
   - \usepackage{multicol}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=F,
                      warning=F,
                      message=F,
                      echo=F)

advdate <- function(obj, adv) {
 tmon <- obj + 7*(adv-1)
 tfri <- obj + 4 + 7*(adv-1)
 tmon <- format(tmon, format="%m/%d")
 tfri <- format(tfri, format="%m/%d")
 zadv <- sprintf("%02d", adv)
 tmp <- paste("Week ",zadv,sep='',", ", tmon," - ",tfri)
 return(tmp)
}

options(scipen=999)
library(tidyverse)
library(stevemisc)
library(lubridate)

## First Monday of semester

mon <- ymd(20260824)

## dates of no class
not_here_dates <- c(
  # APSA
  seq(ymd(20260902), ymd(20260904), by = 1),
  # Paternity leave
  seq(ymd(20260923), ymd(20260929), by = 1),
  # Fall break,
  c(ymd(20261015), ymd(20261016)),
  # Thanksgiving
  seq(ymd(20261123), ymd(20261127), by = 1)
)
finals_week <- 
  seq(ymd(20261212), ymd(20261217), by = 1)

# You can adjust this as you see fit. Basically: add assignment types (e.g. papers, quizzes).
# writing (MA) assignments:
ma_dates <- c(ymd(20261009), 
              ymd(20261106),
              ymd(20261213),
              ymd(20261214)) 
  # Final date set to ...
# graph challenges:
gc_dates <- c(
  ymd(20260911),
  ymd(20260918),
  ymd(20260921),
  ymd(20261002),
  ymd(20261019),
  ymd(20261023),
  ymd(20261030),
  ymd(20261113),
  ymd(20261120)
)

# What are the full dates of the semester? Here, I'll exclude exam week as I like to do.
semester_dates <- 
  seq(ymd(20260827), ymd(20261211), by=1)

# Custom function for treating the first day of the month as the first week 
# of the month up until the first Sunday (unless Sunday was the start of the month)
wom <- function(date) {
    first <- wday(as.Date(paste(year(date),month(date),1,sep="-")))
    return((mday(date)+(first-2)) %/% 7+1)
  }

# Create a data frame of dates, assign to Cal
tibble(date = seq(ymd(20260801), ymd(20261231), by = 1))  %>%
  mutate(mon = lubridate::month(date, label=T, abbr=F), # get month label
         wkdy = weekdays(date, abbreviate=T), # get weekday label
         wkdy = fct_relevel(wkdy, "Sun", "Mon", "Tue", "Wed", "Thu","Fri","Sat"), # make sure Sunday comes first
         semester = ifelse(date %in% semester_dates, 1, 0), # is date part of the semester?
         mas = ifelse(date %in% ma_dates, 1, 0), # is it an MA assignment?
         gcs = ifelse(date %in% gc_dates, 1, 0), # is it a GC assignment?
         not_here = ifelse(date %in% not_here_dates, 1, 0), # is it a day off?
         day = lubridate::mday(date), # get day of month to add later as a label
         # Below: our custom wom() function
         week = wom(date)) -> Cal

# Create a fill category and color category variable, for filling/coloring.
# I can probably make this a case_when(), but this will work.

Cal %>%
  mutate(fcategory = NA,
         ccategory = NA,
         fcategory = ifelse(semester == 1, "Semester", fcategory),
         fcategory = ifelse(semester == 1 & wkdy %in% c("Mon", "Wed", "Fri"), "Class Day", fcategory),
         ccategory = ifelse(mas == 1, "MA Due", ccategory),
         ccategory = ifelse(gcs == 1, "GC Due", ccategory),
         fcategory = ifelse(not_here == 1, "Breaks", fcategory),
         fcategory = ifelse(is.na(fcategory), "NA", fcategory),
         ccategory = ifelse(is.na(ccategory), "NA", ccategory)) -> Cal 

Cal %>% 
  ggplot(.,aes(wkdy, week)) +
  # custom theme stuff below
  # theme_steve_web() + 
  theme_bw() +
  theme(panel.grid.major.x = element_blank()) +
  # geom_tile and facet_wrap will do all the heavy lifting
  geom_tile(alpha=0.8, aes(fill=fcategory,color=ccategory), size=1) +
  facet_wrap(~mon, scales="free", ncol=3) +
  # fill in tiles to make it look more "calendary" (sic)
  geom_text(aes(label=day,color=ccategory),family="Open Sans") +
  # put your y-axis down, flip it, and reverse it
  scale_y_reverse(breaks=NULL) +
  # manually fill scale colors to something you like...
  scale_fill_manual(values=c("Class Day"="steelblue", 
                             "Semester"="lightsteelblue",
                             "Breaks"="forestgreen",
                             "NA" = "grey80"),
                    #... but also suppress a label for a non-class semester day
                    breaks=c("Class Day","Breaks")) +
  scale_color_manual(
    values = c("MA Due" = "firebrick", 
               "GC Due" = "darkblue", 
               "NA" = "white"),
    breaks = c("MA Due", "GC Due")
  ) +
  labs(fill = "", color = "", x="", y="",
       caption = "Notable dates: APSA (9.02-9.04), Paternity (9.23-9.29), Fall Break (10.15-10.16), Thanksgiving (11.23-11.27), Final Day of Instruction (12.11)"
       )  -> class_cal

```

# Course Description

This class is called *War, Wealth, and World Politics*. Really, this is a course about international relations as subject matter and International Relations (IR) as a field of social science. War, wealth, and world politics all fall under this umbrella.  My goal (hope) for this semester isn't to teach you everything you need to know about international relations or IR as a discipline. My objective is more narrow. I want to show you how looking at the world like an IR researcher can help you make sense of international politics and call "BS" on dubious claims about what's going on in the world and the trajectory it's headed in. 

To equip you to do this, I've divided the class into three sections, each oriented around specific questions. In the first part of the course we'll ask: **who are the main actors in international relations, what issues are at stake in their interactions, and why is it so hard for them to cooperate with each other?** In the second part of the course we'll move away from specific issues and ask: **how do IR theories help us simplify the process of making sense of the world, what are their main moving pieces, and how do we verify when or if these theories are useful?** And in the final part of the course we'll home in on a specific but high-profile example of IR scholarship in action about the decline of war, asking: **how did this specific author blend theory and data to help readers make sense of a complex issue in international relations?**

By the end of this course you won't know how the world works --- it's too complex for any one person to comprehend --- but if you work hard and apply yourself, you might just know enough to get your bearings. To use an analogy, I'm not going to give you a map of the world's oceans, but I will give you a compass and sextant. 

\newpage

*Note*: Tentatively, the cadence for the class will be lecture/discussion readings on Mondays and Wednesdays. **In-person assessments** will be on Fridays. I'll flag when we deviate from this pattern.

# Getting Started

## `r advdate(mon, 1)`: Welcome!

- Readings: READ THE SYLLABUS

- Homework: READ THE SYLLABUS AGAIN!! 

- Assessment: NONE

\textit{Note: We are only meeting Friday this week.}

# Part I: The State, Actors, and Issues

## `r advdate(mon, 2)`: Setting the Stage

- Readings: Listen to Paul Poast, \href{https://open.spotify.com/episode/5FHM9YyzclsjYLUhtRLk70}{"So, what do IR scholars actually do?"} and, in Haass, read the intro to "Part IV: Order and Disorder" and the chapters "Sovereignty, Self-Determination, and Balance of Power," "Alliances and Coalitions," and "International Society" (pages 251-279). 

- Homework: NONE

- Assessment: NONE

\textit{Note: I'll be at the APSA Annual Meeting Sept. 2--4. So no class Wednesday or Friday.}

## `r advdate(mon, 3)`: Globalization, Terrorism, and Nukes

- Readings: In Haass, read "Globalization" (pages 159-165), "Terrorism and Counterterrorism" (pages 166-172), and "Nuclear Proliferation" (pages 173-182). 

- Homework: Grade-the-AI (due Friday)

- Assessment: IPA 1 (on Friday)

\hfill \textit{GAI due Fri.}

## `r advdate(mon, 4)`: Climate, Migration, and Cyber 

- Readings: In Haass, read "Climate Change" (pages 183-192), "Migration" (pages 193-200), and "The Internet, Cyberspace, and Cybersecurity" (pages 201-207). 

- Homework: Grade-the-AI (due Friday)

- Assessment: IPA 2 (on Friday)

## `r advdate(mon, 5)`: Health, Trade, and Money

- Readings: In Haass, read "Global Health" (pages 208-214), "Trade and Investment" (pages 2015-229), and "Currency and Monetary Policy" (pages 230-239). 

- Homework: Grade-the-AI (due Friday)

- Assessment: IPA 3 (on Friday)

\textit{Note: My new baby boy is due Wednesday this week. If he comes, class will be cancelled and IPA 3 will be combined with IPA 4 next week. We'll resume class the following Wednesday.}


## `r advdate(mon, 6)`: Development and AI

- Readings: In Haass, read "Development" (pages 240-250) and, from the *Council on Foreign Relations* read ["What Is Artificial Intelligence (AI)?"](https://www.cfr.org/backgrounder/what-artificial-intelligence-ai?_gl=1*fio61f*_gcl_au*MTYyMzUyNjMyNC4xNzU2MTQ5MzE5*_ga*NTM2NzcxMDAwLjE3NTYxNDkzMjA.*_ga_24W5E70YKH*czE3NTYxNDkzMTkkbzEkZzEkdDE3NTYxNDkzODEkajU5JGwwJGgw)

- Homework: Grade-the-AI (due Friday)

- Assessment: IPA 4 (on Friday)

## `r advdate(mon, 7)`: Work on MA1

- Readings: Review

- Homework: MA1 (due Monday 10.12)

- Assessment: NONE

# Part II: Theories

## `r advdate(mon, 8)`: The Supernatural and Theory

- Readings: In Drezner, read "Introduction...to the Undead," "The Zombie Literature," "Defining a Zombie," and "Distracting Debates about Flesh-eating Ghouls" (pages 1-29).

- Homework: NONE

- Assessment: NONE

\textit{Note: Fall break, so no class on Friday.}

## `r advdate(mon, 9)`: The Big 3

- Readings: In Drezner, read "The Realpolitik of the Living Dead" (pages 31-42), "Regulating the Undead in a Liberal World Order" (pages 43-54), and "The Social Construction of Zombies" (pages 55-62). 

- Homework: Grade-the-AI (due Friday)

- Assessment: IPA 5 (on Friday)

## `r advdate(mon, 10)`: Beyond the State as Actor

- Readings: In Drezner, read "Domestic Politics: Are All Zombie Politics Local?" (pages 81-92), "Bureaucratic Politics: The 'Pulling and Hauling' of Zombies" (pages 93-102), and "We're Only Human: Psychological Responses to the Undead" (pages 103-110). 

- Homework: Grade-the-AI (due Friday)

- Assessment: IPA 6 (on Friday)

## `r advdate(mon, 11)`: Critical Perspectives + MA2

- Readings: In Drezner, read "The Supergendered Politics of the Posthuman World" (pages 63-72) and "Subalternity and the Living Dead: Postcolonial Approaches to a Posthuman World" (pages 73-78).

- Homework: MA2 (due next Monday 11.09)

- Assessment: NONE

# Part III: IR Theory and Analysis in Action

## `r advdate(mon, 12)`: Intro to *Only the Dead*

- Readings: In Braumoeller, read "Preface" and "Introduction" (pages xi-17) and "Reasons for Skepticism, Part I: Data" (pages 21-46).

- Homework: Grade-the-AI (due Friday)

- Assessment: IPA 7 (on Friday)

## `r advdate(mon, 13)`: Is War Declining?

- Readings: In Braumoeller, read "Reasons for Skepticism, Part II: Explanation" (pages 47-72), "Is International Conflict Going out of Style?" (pages 75-99), and "Is International Conflict Getting Less Deadly?" (pages 100-130).

- Homework: Grade-the-AI (due Friday)

- Assessment: IPA 8 (on Friday)

## `r advdate(mon, 14)`: Thanksgiving Break

NO CLASS

## `r advdate(mon, 15)`: The Role of Order

- Readings: In Braumoeller, read "Are the Causes of International Conflict Becoming Less Potent?" (pages 131-139), "International Order" (pages 143-177), and "History and International Order" (pages 178-211).

- Homework: Grade-the-AI (due Friday)

- Assessment: IPA 9 (on Friday)

## `r advdate(mon, 16)`: Work on MA3

- Readings: In Braumoeller, read "Conclusion and Implications" (pages 215-224).

- Homework: MA3.

- Assessment: NONE.

## MA3 Due Date

TBD


\singlespacing
\setlength{\parskip}{5pt}

\clearpage

# Course Objectives

After taking this course, my hope is that you are able to answer the following questions:

1. *Who are the main actors in international relations, what issues are at stake in their interactions, and why is it so hard for them to cooperate with each other?* 
2. *How do IR theories help us simplify the process of making sense of the world, what are their main moving pieces, and how do we verify when or if these theories are useful?*
3. *What does it look like to blend theory and data to make sense of a complex issue in international relations?*

I also have some secondary learning objectives for you:

1. *Learn how to think critically about IR theories and data.*
2. *Develop the skill of making (and writing) a clear argument on the basis of theory and evidence*.

No doubt, elements of this course will probably fall short in helping you realize these objectives. All I ask is that you apply yourself and commit to learning, and I'll do my best to be your guide and come up with challenges that help you grow. I can ask for nothing more, or less.

# Required Readings

In the order you'll need them:

1.  Haass, Richard. *The World: A Brief Introduction.* Penguin, 2021.

2.  Drezner, Daniel W. *Theories of International Politics and Zombies: Apocalypse Edition.* Princeton University Press, 2022.

3.  Braumoeller, Bear F. *Only the Dead: The Persistence of War in the Modern Age.* Oxford University Press, 2019.

# Resources

## The Instructor (that's me!)

During my office hours, my door is always open! If you have any questions or concerns about the course, just drop by my office durring office hours (see top of syllabus). No appointment necessary. Outside of those times I'm available for a virtual meeting by appointment on Tuesdays and Thursdays between 1:30-3:00.

You can email me at [williamsmd\@denison.edu](mailto:williamsmd@denison.edu){.email} any time. I'll try to respond within 24 hours, but if you email me on a weekend be prepared for a longer response time.

## Artificial Intelligence

Increasingly students are turning to large language models (LLMs) for help. These tools are not perfect and may, for example, lead you astray in ways that you aren't well equipped to catch. And as you'll see in my AI-use policy (further down in the syllabus) I'm not too keen on you relying on AI (though you will be asked to use it for the Grade-the-AI assignments). While you may hear different views from other faculty (and I respect differing views), I firmly believe that AI-use is a poor scaffold for learning skills. To the contrary, I think learning skills is a necessary scaffold for using AI well. Avoiding turning to AI as much as you can. When you do use it, exercise self-control and treat it like myself or the TA: a resource when you need help, but not someone who will do your work for you.

## Accessibility

Students with a documented disability should complete a Semester Request for Accommodations through the MyAccommodations app on MyDenison. It is the student's responsibility to contact me privately as soon as possible to discuss specific needs and make arrangements well in advance of an evaluation. I rely on the Academic Resource Center (ARC) located in 020 Higley Hall, to verify the need for reasonable accommodations based on the documentation on file in that office. Reasonable accommodations cannot be applied retroactively and therefore ideally should be enacted early in the semester as they are not automatically carried forward from a previous term and must be requested every semester.

## Writing Center

Staffed by student Writing Consultants, the Writing Center is a free resource available to all Denison students. Writing Consultants from a range of majors work with writers one-on-one in all phases of the writing process, including (but not limited to): deciphering assignments, discussing ideas, developing an argument, integrating research and sources, working with faculty feedback, and/or polishing a draft. In addition, Consultants are happy to help with all types of writing, from lab reports, research papers, and informal writing assignments to cover letters, personal statements, and other application materials. The Center welcomes writers from all backgrounds and levels of college preparation, including multilingual writers. Should a multilingual writer need writing assistance that exceeds the abilities of consultants, the writer can be referred to the Coordinator for Multilingual Learning, Kaly Thayer ([thayerk\@denison.edu](mailto:thayerk@denison.edu){.email}). Writing Center consultations will take place in person in in the Atrium level of the Library; please visit the Writing Center's page (<https://my.denison.edu/campus-resources/writing-center>) on MyDenison for specific information regarding hours of availability and how to schedule an appointment. The Writing Center strongly recommends signing up for appointments in advance.

## Multilingual Support (L2)

Students who use English in addition to other languages are welcome to use the resources available at the Multilingual Learning Office. Kaly Thayer, the Assistant Director of Multilingual Learning, and Anna Adams, the English Language Support Specialist, as well the student consultants who work with them, are trained and experienced in helping students address the different issues that arise when working in more than one language. If English is not your first or only language, please consider utilizing this resource, which is available to ALL Denison students. Ms. Thayer, Ms. Adams, and the student consultants offer a variety of support for L2 students, including consulting with you about your written language (grammar, syntax, word-choices), strategies to manage your reading assignments, assistance with class conversation and presentations, and help devising ways to develop and effectively use all your skills in English. You can set up an appointment via MyDenison - Campus Resources - Multilingual Learning, or by emailing the Multilingual Learning Office directly at [englishhelp\@denison.edu](mailto:englishhelp@denison.edu){.email}.

## Reporting Sexual Assault

Essays, journals, and other coursework submitted for this class are generally considered confidential pursuant to the University's student record policies. However, students should be aware that University employees are required by University policy to report allegations of discrimination based on sex, gender, gender identity, gender expression, sexual orientation or pregnancy to the Title IX Coordinator or a Deputy Title IX Coordinator. This includes reporting all incidents of sexual misconduct, sexual assault and suspected abuse/neglect of a minor. Further, employees are to report these incidents that occur on campus and/or that involve students at Denison University whenever the employee becomes aware of a possible incident in the course of their employment, including via coursework or advising conversations. There are others on campus to whom you may speak in confidence, including clergy and medical staff and counselors at the Wellness Center. More information on Title IX and the University's Policy prohibiting sex discrimination, including sexual harassment, sexual misconduct, stalking and retaliation, including support resources, how to report, and prevention and education efforts, can be found at: <https://denison.edu/campus/title-ix> .

# Course Policy

The course policy and requirements are detailed below. It all basically boils down to: (1) show up to class, (2) learn some stuff, and (3) don't cheat or trick me into believing you've accomplished 2.

## Grading Policy

Grades at Denison are based on a standard 4.0 scale. You can read more about Denison's grading system [here](https://catalog.denison.edu/catalog/academic-program/grading-system-evaluation/). Generally, a 90 corresponds to an A$-$, an 80 to a B$-$, etc.

| **Grading Scale** |        |         |
|-------------------|:-------|:--------|
| A+: 98%+          | A: 92% | A-: 90% |
| B+: 88%           | B: 82% | B-: 80% |
| C+: 78%           | C: 72% | C-: 70% |
| D+: 68%           | D: 62% | D-: 60% |
| F: below 60       |        |         |

*Attendance and Participation* **10%**

> You should show up to class and participate! In terms of showing up, I'm generously offering 4 freebie absences (no questions asked). Life happens and my inbox is always full. So if you need a day because you're sick, you're traveling for sports or a performance, or you just want to go do something fun (I once cancelled class to go see Metallica) just take the time you need. However, once those 4 freebies are used up, you should communicate with me to justify why you need to miss class or you will lose all attendance points upon your 5th absence. As for participation, I expect you to come ready to learn and engage in class discussion and group work. One way I'll keep you accountable is by randomly calling on you to answer questions in class. You don't have to get the answers right, but if it's clear you didn't know the answer due to lack of preparedness for class that will cost you a small amount of participation points (1% of your total class grade). 

*3 Main Assignments* **15% Each**

> You'll have three team-based main assignments (MAs) to complete in this course. These will come at the end of each of our three units. Prompts for these will be provided on campus. Each MA will be different but will have the outline of a research paper.


*Grade the AI* **15% Total**

> Almost every week I'll give you a prompt asking a question relavent to the material we cover in class. You'll then give this prompt to an LLM-powered chat-bot (ChatGPT, Claude, Gemini, etc.) You'll copy its response to a Google or Word doc, and then give it a letter grade from "A"-"F" and write a short paragraph justifying your evaluation of the response. Your grade for the AI and justification will be graded by me on a pass/fail basis. If you clearly tried, you'll pass. This exercise (plus the lectures) is meant to prepare you for the main assignments (see above) and the weekly in-person assessments (see below). 

*In-person Assessments* **30% Total**

> Nearly every week (usually Fridays) there will be an in-person assessment. These will take the form of a pencil-and-paper assessment *with no devices or notes allowed*. They should not require longer than a class period to complete, but they will not be timed. If you require additional time to complete these, accomodations will be made. The goal with these assessments is to give me an unobsctructed view of your understanding of the theories/substantive issues related to the subject matter we're discussing in a given unit.

\bigskip



$$
\textbf{Tallying Your Grade:}\\
\begin{aligned}
10\%  & \text{ (Attendance and Participation)} \\
45\%  & \text{ (Main Assigments)} \\
15\%  & \text{ (Practice Sets)} \\
\underline{+ 30\%} &  \text{ (In-person Assessments)} \\
100\% & \text{ (Total Assignments)}
\end{aligned}
$$


## E-mail Policy

I will respond to your emails within 24 hours at the latest between Monday and Friday every week. Don't expect me to respond over the weekend or in the evenings --- those times are family time, so I'll be less consistent about checking my inbox.

## Make-Up Assignment Policy

There are **NO** make-ups for missed assignments. Don't bother asking. But, if you anticipate having troubles making a due-date and notify me ***in advance***, we can work out a solution. Otherwise, you have this syllabus which alerts you when assignments are due. Plan ahead!

## Second Chances Policy

You have exactly one chance to complete your assignments, and only one chance should be necessary. There are days in class specifically carved out to work on your MAs, and my door is always open during office hours. And there are so many other resources at your disposal as you complete your work in this class. Take advantage of them, and you won't need a second chance.

## Attendance Policy

Attendance is part of your grade. As I noted in the "Grading Policy" section, you have 4 freebie absences -- no questions asked or permission required. Beyond these 4 there are no more unexcused absences allowed. Missing class a fifth time will automatically trigger the loss of all your attendance and participation credit unless you communicate with me in advance to justify your absence. 

## Computer-based Excuses

Excuses for late or missed assignments based on CD, flash drive, or hard drive errors are **not acceptable**. The Denison network and server is reliable and accessible. If you use your Google Drive, all your work will be backed up and easy to access from any computer on campus.

## Late Assignments

Turn in your work on time. I want to give you timely feedback, which I can't do if you don't turn your work in. I try to start grading things within a week after due dates. If I don't see a submission from you, or if you haven't communicated with me about an extenuating circumstance requiring an extension, I'll give you a zero.

## Electronic Submission

You will submit all of your assignments electronically via Canvas.

## Academic Dishonesty Policy

Don't cheat. Just don't do it.

It should go without saying, but ***plagiarism*** is a form of cheating and it includes:

1.  Copying or paraphrasing the ideas of others without citation or attribution.
2.  Copying or paraphrasing the ideas of ***other students in the class***.

I've had to deal with students plagiarizing before. It's painful for me and it puts a blight on the record of the student. It's not only cheating, it's stealing.

When in doubt about whether something constitutes cheating, consult Denison's [Code of Academic Integrity](https://denison.edu/academics/curriculum/integrity).\footnote{Of course, if you have to ask yourself if something counts as cheating, then it probably is...} Be advised that this same Code of Academic Integrity requires that instructors notify the Associate Provost of cases of academic dishonesty. **Any incidence of academic dishonesty will result in failure of the course and referral to the Denison judicial process.**

## AI Policy (Or, Actually, a Plea)

```{r}
library(tidyverse)
library(googlesheets4)
library(coolorrr)
set_palette(
  binary = c("gray", "red3")
)
theme_set(theme_classic() + theme(plot.title.position = "plot"))
gs4_deauth()
m26 <- range_speedread("https://docs.google.com/spreadsheets/d/1kyWcrjy73xys1aSlart7njz37WL0BrdLX6tdVin-Nx0/edit?usp=sharing")
m26 <- m26 |> rename_all(tolower) |> 
  filter(finished==1)

m26 |>
  transmute(
    # familiar with AI == 1
    ai_fam = (q27 %in% 1:2) + 0,
    # used AI == 1
    ai_use = ifelse(
        if_all(
          q28_1:q28_6, is.na
      ), 0, 1
    ),
    ## used AI for coding, writing, or summarizing?
    ai_code = ifelse(is.na(q28_5), 0, 1),
    ai_write = ifelse(is.na(q28_3), 0, 1),
    ai_sum = ifelse(is.na(q28_2), 0, 1),
    ai_other = ifelse(
      ai_use == 1 & 
        (ai_code + ai_write + ai_sum) == 0,
      1, 0
    ),
    
    ## AI bad for education?
    ai_bad = ifelse(
      q29_2 %in% 1:2, 1, 0
    )
  ) -> dt
```

I think I've rewritten my AI policy every semester, both because the set of tools for which the "AI" label is used keeps evolving, and because my own views keep evolving, too. 

First, a note on terminology. I personally prefer to call these tools “pattern engines,” both because this is more descriptive of what they do and doesn’t incline me toward thinking these tools are alive in some mysterious way. They are models trained on patterns in data, and they spit out predicted patterns (often very good ones, and other times quite terrible ones) when given new data.

Next, I want to be transparent. I use these tools. I never would call myself a “high-end user,” but I use them all the same. 

Nonetheless, I'm a slow adopter compared to some. I toyed around with chat-bots now and then, like everyone else, after ChatGPT hit the scene, but I only began using the Claude chat-bot with some earnest in the Spring semester of 2026, and over the summer I started experimenting with Posit Assistant, which is an agentic coder that lives inside RStudio (the software I use to conduct research). I’ve always been skeptical of these tools. They can be helpful, but also I have no doubt that the interests of their creators are woefully misaligned with most of society’s, including higher education’s.

My reluctance aside, using these tools has rid me of all doubt that they are powerful and, when used well, can amplify the abilities of the user. I also have no doubt that these tools can undermine critical thinking, short-circuit the ability to learn new skills, and obscure mistakes by covering them with a veneer of sycophantic confidence. 

The fact that these tools can write passable essays and summaries of readings will be a powerful source of temptation for you this semester. You'll do yourself an equally powerful disservice if you succumb to that temptation. These models were trained on text. That means the text they produce and text produced by a human are mutually intelligible and, hence, compliments (things that enhance the value of one another rather than replace the value of one another). Put differently, because of mutual intelligibility, you can read and evaluate the output of AI, check for errors, and make corrections to what it produces. It can do the same for you. This means these tools can amplify your skills as a reader and writer. Seeing them as a replacement for these skills is a clear sign that someone is making a grave category error. The abilities to focus, to understand and evaluate what you read, and to write clearly and in a voice authentic to yourself are more important than ever. 

Of course, all this means that you need to have skills to amplify, and here, yet another irony arises. To develop the ability to read critically and write well, you need to do the hard work, learn to suffer through difficult bits, and repeat. You can pass off the hard work of drafting an essay to an LLM, but that only deprives you of a learning opportunity, and down the road you'll be less effective at using these tools to amplify your writing because you'll have only the poorest of writing skills to amplify. Learning to read and write well requires a patient determination that, I'm sorry to tell you, is a painful exercise in the moment, though it pays dividends in the future. 

You may, then, find it odd that you have a near-weekly assignment in this class called "Grade-the-AI" where I've specifically instructed you to use one of these tools to write an essay in response to a prompt. My goal is actually to help you cultivate the right instincts for LLM-generated text. Whenever you use this tool, your first move should be verification. Your next should be to don the hat of a heavy-handed editor if you plan to use this text for anything you plan to put your name on. That includes identifying and eliminating the vacuous fluff these tools are prone to generating. Both tasks require a user that is well-read and a practiced writer. You'll get plenty of experience with reading in this course, and the main assignments will give you practice writing. At the same time that I'm asking you to generate essays for the Grade-the-AI assignments, I'm asking you to abstain from using these tools to generate summaries of your readings before you've read them yourself, and from using these tools to generate first drafts of your main assignments. It should go without saying, but you should also grade the work of the AI in the Grade-the-AI assignments yourself rather than have the AI evaluate itself. 

This admonition to do the hard work applies to me, too. To use these tools well, I need to maintain and expand my own skills, and that requires continued practice. That's why I carve out spaces (like a weekly [data-driven newsletter](https://mileswilliams.substack.com/) that I write) where I do not use these tools at all. I write my own analysis code, and I write and edit my own words. I think of these sessions as workouts, much like my weightlifting sessions in my basement gym. Skills that go unused atrophy. If you are an athlete, artist, or musician, you know this all too well. 

In your case, the concern isn't atrophied skills as much as the absence of developing skills at all. It's no surprise that most of the empirical research on AI use in education shows mostly adverse effects. 

Most of your fellow students seem to agree with me. Over the last few semesters, DPR has fielded a survey of Denison students that includes questions about AI use and attitudes. As the figure below makes clear, most of your fellow Denisonians use AI in some way for help with coursework. In the Spring 2026 wave of the survey, 85% reported familiarity with AI, and over 81% said they used it for help with coursework in some form or fashion (how 15% were unfamiliar with AI is beyond me). 

```{r}
#| fig-height: 3
#| fig-width: 6

dt |>
  summarize(
    across(everything(), ~ mean(.x, na.rm = T))
  ) |>
  pivot_longer(1:6) |>
  mutate(
    ord = n():1,
    lab = c("Are familiar with AI", "Use AI for anything", "...for coding",
            "...for writing", "...for summaries",
            "...for something else")
  ) |>
  ggplot() +
  aes(
    x = value,
    y = reorder(lab, value)
  ) +
  geom_col(
    fill = "red3",
    color = "black"
  ) +
  geom_text(
    aes(label = scales::percent(value)),
    color = "white",
    fontface = "bold",
    hjust = 1.1
  ) +
  scale_x_continuous(
    breaks = NULL
  ) +
  labs(
    title = "Most Students Use AI",
    subtitle = "% of students that:",
    x = NULL,
    y = NULL,
    caption = "Source: DU Student Survey Spring 2026"
  ) +
  theme(
    panel.grid.major = element_blank()
  )
```

Here's the kicker. As you can see in the next figure, when you break responses down by whether and what students use AI for, and calculate the share who think AI undermines the value of a college education, those that use AI most, worry about its negative impact. The margins are substantial. 77% of students who use AI for writing think AI undermines the value of a college education, compared with 53% of those who don't. 71% who use it for coding think AI undermines a college education, compared to 52% who don't. The biggest gap lies between those who use AI for anything and those who don't use it at all. 63% of students who use AI for anything think AI undermines the value of a college education compared to 25% who don't use it at all.


```{r}
#| fig-height: 6
#| fig-width: 6
library(socsci)
dt |>
  pivot_longer(1:6) |>
  group_by(value, name) |>
  mean_ci(ai_bad, ci = .84) |>
  group_by(name) |>
  mutate(
    ord = max(mean)
  ) |>
  ungroup() |>
  mutate(
     lab = rep(
      c("Coding", "Familiar", "Other Use",
        "Summarizing", "Any Use", "Writing"),
      len = 12
    )
  ) |>
  ggplot() +
  aes(
    x = mean,
    y = reorder(lab, ord),
    color = ifelse(value == 0, "Non-users", "Users"),
    xmin = lower,
    xmax = upper
  ) +
  geom_pointrange() +
  geom_text(
    aes(label = scales::percent(mean, 1)),
    vjust = -1,
    fontface = "bold",
    show.legend = F
  ) +
  geom_vline(
    xintercept = .5,
    lty = 2,
    color = "black",
    linewidth = .5
  ) +
  labs(
    title = "AI Users Think AI Undermines Education",
    subtitle = "Share saying AI undermines education by AI use (84% CIs shown)",
    x = NULL,
    y = NULL,
    color = NULL,
    caption = "Source: DU Survey Spring 2026"
  ) +
  scale_x_continuous(
    labels = scales::percent,
    breaks = seq(0, 1, by = .1)
  ) +
  theme(
    legend.position = "bottom"
  ) +
  ggpal("binary")
```

Denisonians aren’t unique in holding these views. Just go online and search for other recent national surveys and you can find a similar pattern. Many people are worried about the negative impact these tools have on learning, and those that use these tools more frequently often harbor the strongest concerns. 

Many also worry (rightly) about the impact of these tools on credentials, like college degrees. But I have some good news for you. Most of this concern is focused on online, asynchronous programs and those at large state schools where it is comparatively easy to get away with the illusion of learning by outsourcing assignments to a chat-bot or coding agent. Now, because of these tools, is the best time to earn a degree at a place like Denison where smaller class sizes and face-to-face interactions with faculty send a more credible signal to future employers that you earned your degree. Partly this is because of the perception that it is easier to police unsanctioned AI use in a small, close-knit environment (which, I have to admit, isn’t always true). It’s also because a place like Denison gives you more opportunities to develop relationships with faculty and mentors who can vouch for your learning. 

Building on that last point, human connections are now more vital than ever, so don’t squander them by using AI when your professors ask you not to. At a place like Denison, you aren’t an anonymous face in the crowd. Faculty will remember you. In the future, when you need a faculty member to vouch for you, think about the impression you want to leave them with. Will they remember you as a student who blatantly used AI when told not to, or the student who put in the work? I’ll gladly vouch for a student who worked hard to earn a “B” or a “C” before I consider vouching for a student who did next to no work and got an “A”. 

So, carefully weigh the relational impact of AI use. Unsanctioned use could damage relationships with people you need to speak on your behalf down the road. These relationships will set you apart from the mass of people currently using Claude or ChatGPT to complete their online degrees who may, at the end of that experience, find that they don’t have an actual human being that can directly speak to their abilities when a future employer asks for references.

***Okay, Dr. Williams, wrap this thing up. How can and can’t students in this class use AI?*** My ask is simple. Treat AI like an additional resource, much like yours truly, ARC, the Writing Center, etc. All of these resources share two things in common: (1) they will support your learning by providing feedback, redirecting you to the right resources if they can’t answer your questions, or giving you helpful tips for completing assignments; and (2) they will not complete your assignments for you. When I ask you to do your own work, I expect you to do it. The learning process will teach you the habits of determination and patience required to solve arduous problems you may face in life and in work, and it will cultivate a habit of integrity that will pay off immensely for the human connections you form here at Denison. 

Thus ends my plea. 

