---
title: "Getting Started, Part I"
output: 
  github_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<center>

[Back to Notes Homepage](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/README.md) | [Getting Started, Part II -->](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/02_getting_started_cont.md)

</center>

## Goals

Learning objectives:

- Know how to access the RStudio server
- Get familiar with the elements of RStudio
- Know what the R console is
- Know how to open and use an RMarkdown file
- Start an R project and create a sensible file structure
- Know how to exit RStudio the right way

## The lay of the land

In this class we'll use R and RStudio for data visualization. Before we get started, it's essential that you familiarize yourself with the RStudio environment.

First, access the [RStudio server](r.denison.edu) and log-in with your Denison credentials. If you're off campus, you'll need a VPN connection to access the server. You can read more about how to do that via one of the following links:

-  [For Windows Users](https://docs.google.com/document/d/1z1eapXs8Lp0yA7TBXmLfViiz6U67ADvLgpFeWJoSLLA/edit#heading=h.48vu9pu55b3s)
-  [For Mac Users](https://docs.google.com/document/d/1FaJN4et9H9eKsJIVinwqC903NbKA5A-IYfvDPlFwwXk/edit#heading=h.48vu9pu55b3s)

Once you access the R-server, you may already be in a project space. If not, you'll need to start a new session.

The RStudio environment has a lot of components. It may look like a lot at first. If you feel overwhelmed, that's okay. In time, it'll all make more sense.

For now, I want to direct your attention to a few things:

1. **The R console:** This is the window in the lower-left side of the screen. This is where the code you write runs and where output from code appears. While you can work directly in the console, I almost never recommend doing so (but I'll let you know when there's an exception).
2. **Files**: On the lower-right you'll see a window with a few different tabs, one of which says "Files." This is where you'll save your work. In a bit, I'll give you some instructions for how I want you to set up your files.
3. **RMarkdown or .Rmd files**: Almost all of the coding you do in this class will take place in .Rmd or RMarkdown files. These let you work in plain text (taking notes, making comments, etc.) alongside your code. You can open a new RMarkdown file by clicking on the "+" button in the upper left. You'll see a drop-down menu open. Select "R Markdown" from the list of options. The file you see when you open this is an RMarkdown file. It has some boilerplate script in it that summarizes what it is and how you can use it.

At this point you should be able to tell that a lot is going on in RStudio. That's because RStudio is an "integrated development environment" (IDE). It actually is a separate application from R itself. R is pretty spartan all on its own, and it definitely falls short as far as user-friendly software goes. RStudio provides a better interface with R proper that lets you organize files, save your work in projects, and write reports within a single environment.

## Starting a project

The best way to work in RStudio is to save your work in projects. This lets you organize all your work in a tidy way. I highly recommend working in projects.

If you look at the upper right corner, you can see a cube with an "R" in it. If you select that you'll see a drop-down menu. Select "New Project" > "New Directory" >  "New Project" > enter a new directory name for your project > use "brows"  to find a place in your files you'd like to save your work > then create. For this class, I recommend creating a project called "DPR 101" that way it's obvious to your future self that this is where your work for this class is located.

After you create your project, create two new folders in your files. Call one "Data" and the other "Code." One will be where you save different datasets you use in this class. The other will be where you save the RMarkdown files you use in this class for notes and course assignments.

## Getting Started in RMarkdown

RMarkdown documents are a great place to 

1. Write code
2. Look at your output in-line with your code
3. Take notes alongside your code

You can also use these documents to write reports, but we'll talk more about that later.

These features of working in RMarkdown are great for learning to code with `ggplot()`, which is the way you'll create your data visualizations in this course. By being able to write your code together with your notes, it makes it really easy to move back and forth between tasks 1-3 listed above. You can make notes to yourself in plain text about what data you're working with and what your code is supposed to be doing. You can create your code. You can run it. Then you can adjust based on the output, make new notes, and repeat.

There are lots of helpful cheatsheets out there for working with RMarkdown. This is a [great one](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf?_ga=2.235103752.840764392.1674585715-719901239.1627613290).

When you work with Markdown, you can create section headers using `#` (where more hashtags = smaller headings). For example, here's text you would write to make a header:

`## Header`

Here's what it looks like after you "Knit" your document (I'll get to what "Knit" means in a second):

## Header

Headers don't only have value for creating reports. They provide you with a convenient way to navigate your notes and code. If you create a new section header, you can click on the orange hashtag at the bottom left of your RMarkdown file. You'll see a menu pop up that lists all the different headers you have in your document. If you click on one of these headers, you instantly are transported to that place in your file.

When you're done with a document, you can feed it to R using the "Knit" button at the top (the one that has the ball of yarn picture next to it). You can *knit* your document into various kinds of files like HTML, PDF, or Word. Try knitting to a Word doc and see what happens.

First, you'll notice that RStudio has created a Word document based on your RMarkdown file. If you open it, you'll notice, too, that all of your plain text along with your code chunks and their outputs appear in the knitted document. 

When you knit something, you can update a few things about how it renders. For example, go to the place in your file that says:

```
knitr::opts_chunk$set(echo = TRUE)
```

Change `echo = TRUE` to `echo = FALSE` then knit your document again. Do you see the difference? Now only your text appears and the output from your code, but not the code chunks. This is a great option when writing reports (or working on your main assignments in this class). You can write your report in plain text with headers and so on, and write your code alongside to produce various data visualizations (all in one document). 

## Code chunks

When you use RMarkdown, your notes/comments/writing in plain text will be interspersed with [**code chunks**](https://rmarkdown.rstudio.com/lesson-3.html).

A code chunk is created using three backticks ("```") followed by an "r" in brackets, and then it's closed with three more backticks.

Think of each code chunk as a self-contained space for writing and running a specific bit of code. 

After you make a code chunk and write some code in it, you have a bunch of different options for running it.

1. Click the green arrow to the right top side of the chunk, OR
2. On Windows, with your cursor somewhere inside the chunk, use "control + shift + enter" or in Mac use "command + shift + return".

In addition to making notes in plain text around your code chunks, you can make notes inside code chunks as well. Anything that follows a `#` in a bit of R code is "commented out." That means R knows not to run anything that follows the hashtag in the code. For example:

```{r}
# this is a comment that won't run any code
1 + 3 # this is some code that will run (but this comment won't)
# 1 + 3
```


## Working with R

There are few things to know about R. First and foremost, R is not just an application, it's a language. And just like learning any language, fluency in R takes time and a lot of practice.

R specifically is an "object oriented" and "functional" programming language. That means a few things.

First, **everything in R has a name**. You refer to the names of things to examine them or use them. These things can be variables or datasets that you manipulate, or functions that you use to perform operations.

Like any language, there are some grammatical rules in R that you should never break (and cannot break if you tried). For example, words like `TRUE` or `FALSE`, `Inf` or `else`, and several others have been reserved for core programming purposes and you couldn't name something in R one of these things if you tried.

Other words or letters, like `q`, `c`, or `mean` can technically be used to refer to other things, but avoid doing so! These are the names of basic functions in R, and if you give other things in R the same names, R will get confused and angry with you.

R is also case sensitive. So if something is named `This` R won't know what you're talking about if you try to call `This` by instead writing `this`.

Second, **everything in R is an object**.

Say we use the command `c()`, which is a function that stands for "concatenate." It takes a sequence of commands and returns a **vector** where each element is accessible:

```{r}
c(1, 2, 4, 8, 16, 32)
```

The output from the above is just all the elements in the vector we created using `c()`. If we didn't want this to just appear in the console but instead have it saved, we would need to **assign** the vector a name, which then saves it as an object:

```{r}
my_numbers <- c(1, 2, 4, 16, 32)
```

Now, every time we call the object `my_numbers`, the output will appear in the console (or as the output of a code chunk):

```{r}
my_numbers
```

Each of the numbers in this vector can be accessed directly, too. This is done using square brackets `[]` after the name of the object:

```{r}
my_numbers[3]
```

The above prints the 3rd element in the vector called `my_numbers`.

We created `my_numbers` using an **assignment operator** `<-`. When you want to save something as an object, you need to use an assignment operator. A cool feature of it, though, is that it can work not only from the right to the left (the usual way), but also from the left to the right. The latter is sometimes called "reverse assignment;" though, in reality it actually may be more intuitive for you.

```{r}
# norm assignment
x <- c(1, 2, 3)

# reverse assignment
c(4, 5, 6) -> y
```

You can technically use the `=` operator to assign things, too, but there are some things to note about this:

1. It only works in the "normal" direction for assignment (right to left)
2. It there are some objects that it won't play nice with

Generally, it's considered bad grammar to use `=` for assignment. Instead, we use `=` inside of functions (coming up next) to set commands or feed objects to functions when we want to perform an operation.

Speaking of functions, just about **everything you do in R with be with a function**. A function is a special kind of object that performs actions for you. You feed it some kind of input (like an object) and it provides an output (which you can assign to a new object for later use).

For example, there's a function called `mean()` which we can use on the object `my_numbers`:

```{r}
mean(x = my_numbers)
```

The function does exactly what its name suggests---it returns the mean or average of whatever numerical vector you feed it.

However, like all functions, `mean()` has some rules about what kinds of inputs it will accept.

If you feed it nothing, it'll give you an error that says `Error in mean.default() : argument "x" is missing, with no default`

If you feed it non-numerical data, it'll give you a warning and return `NA`:

```{r}
my_words <- c("Hello", "World!")
mean(x = my_words)
```

What's the average of "Hello" and "World!" I dunno, and neither does R.

If you ever want to learn more about a function, R is chock full of help files. All you need to do to access the help file for a particular function is write something like `help(function_name)` in the console. You could just write `?function_name` in the console as well.

An important thing to note about functions is that they come in **packages**. Some functions, like `mean()`, are in the base R package which is already open and ready to go the moment you open R.

Other functions can't be used until you **attach the package** using the `library()` function. 

In this class, we'll use the tidyverse of R packages. The tidyverse is actually a package of packages that have functions that are meant to be used together. Rather than attach each package in the tidyverse individually, if we write `library(tidyverse)` all these packages and the functions they contain are immediately accessible to us.

```{r}
library(tidyverse)
```


Some packages have already been pre-installed for you if you're using the Denison server. If you aren't, you'll need to install these using the `install.packages()` function. 

Also note that some packages have been produced by users and aren't yet accessible from the "CRAN." For these functions, you have to take some additional steps to install the package. 

One in particular is the coolorrr package that I'll have you use this semester. This is a package that I developed to help with color palettes. To install it, you'll need to run the following in your console:

```
devtools::install_github("milesdwilliams15/coolorrr")
```

Notice in the above that I used double colons `::` after `devtools`. If you ever only want to access a single function from a package (`devtools` is a package that helps with installing packages from sources like GitHub), but don't want to attach the full package in R, you can write the package name followed by `::` to call the function you want. The syntax will be something like `package_name::function_name()`

## More details about R

R is many things, including a glorified calculator. You can use a lot of different operations like `*` for multiplication, `/` for division, `+` for addition, and `-` for subtraction.

R also uses a number of **logical operators** like and `&`, or `|`, not `!`, equal to `==`, greater than `>`, greater than or equal to `>=`, less than `<`, less than or equal to `<=`, not equal to `!=`, and in `%in%`.

Remember the `x` and `y` objects I created earlier? Let's try out some of these operations on them and see what happens:

```{r}
# Mathematical operations
x + y # addition
x - y # subtraction
x * y # multiplication
x / y # division
# Logical operations
x == y # equivalence
x <= y # x less than or equal to y?
x %in% y # are x values in y?
```

Notice that mathematical operators return numerical outputs, while logical operators return logical outputs (`TRUE` or `FALSE`).

Note that R treats logical values as 0-1 values. For example, you can take the mean of a vector of `TRUE` and `FALSE` values:

```{r}
mean(x = c(0, 1))
mean(x = c(F, T)) # these are the same
```

You can also just write `T` for `TRUE` and `F` for false, as I did in the above chunk.

## Working with data tables

Most of the work we do in this class will involve working with datasets. Think of these as tables that store data in a central location for ease of access and use.

`mtcars` is a dataset that already comes pre-installed in R. We can check out the first 10 rows of the data using the `head()` function:

```{r}
head(mtcars, 10)
```

This is a "dataframe," or what we might also call a dataset.

To access a single column of a dataframe, just use the syntax `dataset$variable`:

```{r}
mtcars$mpg
```

A particularly useful way to save a dataframe in R is as a **tibble**. To convert a dataframe to a tibble (which is a special kind of dataframe) just write:

```{r}
mtcars_tb <- as_tibble(mtcars)
head(mtcars_tb, 10)
```

There are lots of ways for peering into a dataset/dataframe to get a sense for its structure (what it contains, how big it is, etc.).

You can use `summary()` to get some quick summary statistics that tell you about the variables in a dataset:

```{r}
summary(mtcars_tb)
```

You can use the `glimpse()` function to check the data's structure:

```{r}
glimpse(mtcars_tb)
```

## Closing RStudio

Every time you're done working, make sure you do a few things:

1. Close out the project session
2. DO NOT save your workspace image
3. DO save changes to files

You should always perform the above three steps when you close out (with special emphasis placed on step 2). This ensures R is always running smoothly and efficiently.

## Wrapping up

Be patient with yourself as you start working in R and RStudio. At the same time that you are familiarizing yourself with new software, you also are learning to speak a new language. If things don't make sense at first, that's okay. That's normal.

I can't possibly anticipate every possible issue you may run into as you use R, but I can give you a heads up about some common mistakes people make:

- Make sure that all parentheses and brackets are complete. If you start a phrase with an open parentheses "(", you need to have a corresponding ")" to close it.
- Sometimes you may not realize that you've only run part of your code and not all of it. You can see if this is the case if you check the console. If you see a `>` at the bottom of the console, all is good. If you see a `+` then something only partially ran.
- Give new objects that you create new and unique names. I can't tell you how many times I've seen someone's code break down because they created an object called `x`, then ran a function on it and saved the output as `x`, then tried to go back an run an old chunk on `x` only to find that it spits out an error. The old `x` that used to work with a function now no longer does because the new `x` isn't the same thing! 
- Don't forget to create new objects. If you don't, changes that you make to an object won't be saved in R's environment.

## Where to next?

<center>

[Back to Notes Homepage](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/README.md) | [Getting Started, Part II -->](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/02_getting_started_cont.md)

</center>