---
title: "Getting Started, Part II"
output: 
  github_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F,
                      out.width = "75%",
                      dpi = 500)
```

<center>

[<-- Getting Started, Part I](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/01_getting_started.md) | [Back to Notes Homepage](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/README.md) | [`ggplot()` Basics, Part I -->](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/03_ggplot_pt1.md)

</center>


## Goals

- Get data into R
- Make your first data viz
- Cover some helpful resources if you want to go deeper into the details of using R


## Reading data into R

Data viz is impossible without data. The only way we can use data for data viz is to get it into R. 

A lot of packages (including base R) provide tools for reading data into R. In this class, we'll primarily use the `read_csv()` function from the `{readr}` package. This is one of the many packages in the `{tidyverse}` ecosystem of packages.

There are three main ways that I will have you access data in this class. The first is via .csv files I have stored on my GitHub (the place that you're reading these notes!). The second is on Google Drive. The third is using R packages that have been developed to specifically pull data from different online databases.

Let's go over these different approaches.

### Way 1: From my GitHub

Let's start with the first way. I put together a dataset that cross-references instances of conflict onset for countries over time with their quality of governance score from the [Worldwide Governance Indicators](https://datacatalog.worldbank.org/search/dataset/0038026) database. Here's the url to the raw .csv file on my GitHub: [https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20101/Data/onset_and_wgi.csv](https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20101/Data/onset_and_wgi.csv)

To read the data into R, we can use the `read_csv()` function like so:

```{r}
library(tidyverse)
Data <- read_csv(
  "https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20101/Data/onset_and_wgi.csv"
)
```

Notice that I used quotation marks (`""`) around the url. Some things in R require the use of quotation marks (like data files or character strings) and other things don't (like the names of objects or functions).

You can give `read_csv()` a url like I did above, or the location of something in your local files. For example, I'm writing this note on my laptop, and the same dataset I have on my GitHub also lives in my files on my computer. To access it locally, I could use the `here()` function from the `{here}` package to tell R where to pull the data from:

```{r}
file_location <- here::here(
  "DPR 101", "Data", "onset_and_wgi.csv"
)
Data <- read_csv(file_location)
```

You don't have to use `here()`, but I like using it because it automatically adds all the pesky backtics I'd need to add to tell R where a file lives. If we look at the `file_location` object, we can see that `here()` added this for me and then some:

```{r}
file_location
```

Obviously, the above is unique to my computer. The input I gave `here()` and the output would look different for you.

Okay, back to the data. When `read_csv()` reads the data into R, it will spit out a few messages when it's done. These provide details about how `read_csv()` assigned variable classes to each of the columns in the data. In this case, it tells us that it assigned `country` to the character class (that means it's a non-ordered category) and it assigned `year`, `sumonset1`, and `wgi` the class double, which is R's way of saying real numbers.


### Way 2: From Google Drive

Some datasets for this class will come from Google Drive. Here's an example pulling from a Google Sheets document that has voter turnout data from a field experiment done in New Haven, CT in 1998:

```{r}
url <- "https://docs.google.com/spreadsheets/d/19RaIaVoJMChVsNGO45OrqZcSE3to-EwQ-vywngHLbks/edit#gid=817523709"
library(googlesheets4)
gs4_deauth()
turnout_data <- range_speedread(url)
```

To access Google Sheet files you need to open the `{googlesheets4}` package. It has a few functions for reading in data, but the best and fastest is `range_speedread()`.

The workflow for reading in data this way is very similar to the approach I took for reading in .csv files from GitHub. The main differences are:

1. You need to load the `{googlesheets4}` package.
2. You need to run the function `gs4_deauth()` before you try reading in the data.

The `gs4_deauth()` function removes the need to enter in some additional permissions for accessing files from Drive.


### Way 3: R Packages

Some datasets come pre-installed in R, and some are accessible with different R packages. For example, when you open the `socviz` package that Healy created to go along with our textbook you immediately have access to a number of datasets used in examples in the text.

There are tons of other packages that have been created to make it possible to access all kinds of political data directly in R. Some of these, like `{DemocracyData}` and `{peacesciencer}` we'll use later in this class.


## Making a figure

Speaking of accessing data using R packages, the first example of a data viz we see in Chapter 2 of Healy uses the Gapminder dataset from the `{gapminder}` package.

To make the data accessible in R, all we need to do is write:

```{r}
library(gapminder)
```

We now have an object called `gapminder` in R. This is a dataset that contains information for a bunch of countries over time about wealth and life expectancy. Below, I'm using the `sample_n()` function to look at 10 random rows from the data:

```{r}
gapminder %>%
  sample_n(10)
```

Using this data, we can make a simple scatter plot showing how per capita GDP predicts life expectancy:

```{r}
ggplot(gapminder) + 
  aes(x = gdpPercap, y = lifeExp) +
  geom_point()
```

You can notice a few things about ggplot from the code used to produce the above figure. First, ggplot works by building figures in steps. We call these **layers**. Second, we **add** layers (literally) by using the `+` operator. While normally we use this for addition (i.e., `2 + 2`) when we use ggplot `+` acts a lot like this thing called a pipe operator (`%>%` or for later versions of R `|>`), which we'll talk about more later. Basically, the `+` in ggplot just tells R that we want to add a new set of commands or instructions for creating a data viz.

More formally, the ggplot workflow looks like:

1. Feed ggplot data.
2. Map aesthetics (tell ggplot what relationships to show).
3. Draw geometry (tell ggplot how to show these relationships).

There technically is a fourth step---customization---but we'll save that for another day.

Back to the figure we made, as a first pass this isn't too bad. We could obviously add a few more flourishes to make our data viz publication ready, but this is enough to get a sense for the data. Take a look at the figure. What does it tell us about life expectancy and GDP per capita?

A cool thing to note about making figures with `ggplot()` is that we can save them as objects in R. Check it out:

```{r}
p <- ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp) + 
  geom_point()
```

The object `p` is our ggplot data viz. Now, every time we write `p`, it tells R to produce the figure:

```{r}
p
```

This feature of working with ggplot is great for reasons that will become more obvious later in the course. The biggest benefit is that it lets us build up a solid foundation for a data viz and then add new details later.

For example, say we want to compare the above figure to a version where the scale for GDP per capita is different (say using the log10 scale). All we need to do is add a new layer to `p` like so:

```{r}
p + scale_x_log10()
```

Because I saved the first plot as the object `p`, I didn't have to re-write the code to produce the old plot before adding a new layer.

Speaking of this new layer, does the log10 scale change any conclusions you previously drew about the relationship between per capita GDP and life expectancy?


## Helpful Resources for Learning More

This class is not about how to use R in its entirety. But, you may find it helpful to get more familiar with it. Here are some resources for you to check out on your own time:

- [swirlstats.com](https://swirlstats.com/)
- [Datacamp](https://www.datacamp.com/data-courses/r-courses)
- [Posit cheatsheets](https://posit.co/resources/cheatsheets/)

My personal favorite is swirlstats.com, because it lets you work at your own pace directly in R, and **for free**.


## Wrapping up

Working with code is hard, and I promise you that you **will** run into problems. When you do, just remember that everyone has problems with their code. It's normal, and it would be weird if you didn't have *any* issues.

We'll get more into the details of working with ggplot in the coming weeks. But for now, you should have at minimum some helpful examples for how to:

- Read data into R
- Make a scatter plot using ggplot
- Access other resources for working in R

## Where to next?

<center>

[<-- Getting Started, Part I](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/01_getting_started.md) | [Back to Notes Homepage](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/README.md) | [`ggplot()` Basics, Part I -->](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/03_ggplot_pt1.md)

</center>
