---
title: "Making ggplots Part 2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")
```

## Goals

- Understand the difference between mapping and setting aesthetics.
- Set aesthetics globally vs. by geom.
- Know how to save your data viz.


## When to map or set aesthetics

In the previous notes, we covered some of the basics of working with ggplot. [[Check them out here.](https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Notes/03_ggplot_pt1.md)]

In these notes, we're going to go a little deeper. Up to now we've talked about the basic ggplot workflow:

1. Feed `ggplot()` data.
2. Tell ggplot what relationships to show using `aes()`.
3. Tell ggplot how to show those relationships using `geom_*()` functions.

There are some important things to keep in mind when doing steps 2 and 3. Namely, it's really easy to confuse when you want to *map an aesthetic* for when you want to *set an aesthetic*.

What's the difference? 

- You map aesthetics inside `aes()`.
- You set aesthetics inside `geom_*()`.

Here's an example. Let's load the `gapminder` data and the `{tidyverse}` and make a plot with ggplot where we set some options for the color aesthetic. 

```{r}
library(tidyverse)
# open the tidyverse of packages

library(gapminder)
# access the gapminder data frame

ggplot(gapminder) + 
  aes(x = gdpPercap, 
      y = lifeExp,
      color = continent) +
  geom_point() +
  geom_smooth() + 
  scale_x_log10()
# produces a plot where line and point colors
# correspond with continent names
```

The above code contains a command in the `aes()` function that specifies `color = continent`. This option tells ggplot that it should use different colors to map to continent names. 

To do this, ggplot uses a default color palette for the mapping process (we'll talk about how to customize this later). What it doesn't do (at least using `aes()`) is let you set a specific color. Look at what happens if you try:

```{r}
ggplot(gapminder) + 
  aes(x = gdpPercap, y = lifeExp, color = "gold") +
  geom_point() +
  geom_smooth() +
  scale_x_log10()
```

Does the output look "gold" to you? It sure doesn't! Instead, it's read. What went wrong here? 

This is a classic example of mapping an aesthetic when what we actually need to do is set an aesthetic. To set the color of points and lines, we don't specify them by mapping. In the above example, setting a color in `aes()` leads ggplot to mistakenly conclude that we are adding a new column where each cell entry of that column is a category called "gold." It then picks a color to map to that category. 

To avoid this, we set aesthetic inside the geoms we tell ggplot to draw rather than inside `aes()`:

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp) +
  geom_point(
    color = "gold"
  ) +
  geom_smooth(
    color = "red"
  ) +
  scale_x_log10()
```

This terrible looking, McDonald's colored data viz looks just like it should. Instead of specifying color in `aes()` we specified it in the geom layers. On top of that, we gave each layer different instructions about the color to use. We told ggplot to color the points gold and the smoothed regression line red.

There are lots of different aesthetics that you can either map or set with ggplot, such as size, color, alpha (which indicates transparency), fill, and on and on.

For example, say we wanted to map point size to population in a scatter plot and set the transparency of points using alpha:

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha = 0.5) +
  scale_x_log10()
```

The above code maps point size to the `pop` variable in the `gapminder` data so that `geom_point()` draws larger points for observations that have a larger population. The command `alpha = 0.5` in `geom_point()` sets the transparency of points to 50\%.


## You can map aesthetics per geom layer or globally

Not only can we map and set different aesthetics at the same time, we can also tell ggplot just to map an aesthetic for one geom layer rather than all of them. Look at the difference between the plot this code produces:

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, color = continent) +
  geom_point() +
  geom_smooth() +
  scale_x_log10()
```

And compare it to what this code produces:

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp) +
  geom_point() +
  geom_smooth(aes(color = continent)) +
  scale_x_log10()
```

The first code maps color to continent name **globally**. That means that for each geom layer that we add, color will be mapped to continents. Conversely, the second code chunk maps color to continents only for the `geom_smooth()` layer. This is done by adding a new call to `aes()` inside the geom function.

Notice, too, the difference in the legends ggplot produces for these figures. Ggplot legends will always faithfully reflect the aesthetic mappings you add to your plots.


## Pick and choose your aesthetics wisely

Ggplot will dutifully map all kinds of aesthetics that you tell it to map. That doesn't mean that every possible mapping is a good idea.

Take the example from above where we mapped point size to population. This was a sensible choice because people have an easy time intuiting that point size corresponds with the ***size*** of the observation. We could have used color instead. Would this choice have been as effective?

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, color = pop) +
  geom_point() +
  scale_x_log10()
```

The answer is **no**. Here's another example, this time for a column plot where we've mapped the color of columns to continents:

```{r}
ggplot(gapminder) +
  aes(x = year, y = gdpPercap, color = continent) +
  geom_col()
```

There are a lot of empty dark grey spaces in the figure. That's because when you produce a column plot, the color aesthetic colors the borders of the columns, not the inside. We instead need to use the fill aesthetic:

```{r}
ggplot(gapminder) +
  aes(x = year, y = gdpPercap, fill = continent) +
  geom_col()
```


## How to save your plots

After you've produced your data viz, you probably will want to save it for later use for writing a report or to share on social media or some other outlet.

You have a lot of different options for saving your figures.


#### Copy and paste

You can copy and paste the data viz you produce directly from RStudio. Just right click on the figure that your code chunk produce and select copy and/or save to save it somewhere in your files.


#### Save using `ggsave()`

You can also using the `ggsave()` function to save a plot directly to somewhere in your files. 

Say you created a folder in your project called `Figures`. Here's how you would use a combo of `here()` and `ggsave()` to save your work:

```
my_plot <- ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha = 0.5) +
  scale_x_log10()
# save your ggplot as an object

library(here)
# open the here package

ggsave(here("Figures", "my_first_figure.png"), 
       plot = my_plot)
# save it to the Figures folder and name it
# "my_first_figure.png" to save it as a .png file
```

This is my favorite and preferred way of saving a figure. The biggest pros of using this approach are:

1. The ability to control dimensions.

2. The ability to control the sharpness of the figure.

You can check out these options and a lot more by writing `?ggsave()` in the console.


## Things to try

Play around with the `gapminder` data and try out some different aesthetics to see what happens.

You can also try out some different plot scales. I used `scale_x_log10()` to put the x-axis on the log-10 scale in the above figures. There are [lots of other scaling options](https://ggplot2.tidyverse.org/reference/scale_continuous.html), too.

Try to save some of you figures and play around with `ggsave()` options to see what happens.
