---
title: "Making sure you show the right numbers, Part 1"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")
```

## Goals

- Understand where different errors in plotting come from.
- Understand how you can use ggplot commands to transform data.


## Making a data viz is an iterative process

The first time you write some code to produce a data viz, it likely won't look pretty. Sometimes you will try to tell R to do one thing, but the way you say it won't be sensible or meaningful to R. You'll have to write and rewrite your code until you get it right. I still have to do the same thing all the time. 

If we open the `{socviz}` package we can access some data called `county_data`. This provides information at the county level in the US about different population demographics and election outcomes in 2018.

Let's look at the relationship between population size and the share of the population that's black. And let's also try using `geom_line()`, which draws lines by connecting observations in order of the variable on the x-axis. This choices may seem sensible. We want to see how population relates to the share of the population that's black. But, when we tell ggplot to show the relationship using `geom_line()` we get some wonky output.

```{r}
library(tidyverse)
# open tools in the tidyverse

library(socviz)
# open socviz to access country_data

Data <- na.omit(county_data)

# here's the plot:
ggplot(Data) +
  aes(x = pop, y = black) +
  geom_line() + 
  scale_x_log10()
```

What went wrong? We told R what to do, but our instructions weren't terribly sensible. This is a case where a different geometry makes better sense. For example, `geom_point()`:

```{r}
ggplot(Data) +
  aes(x = pop, y = black) +
  geom_point() +
  scale_x_log10()
```

We might also want to add a smoother to show the average trend.

```{r}
ggplot(Data) +
  aes(x = pop, y = black) +
  geom_point() +
  geom_smooth() +
  scale_x_log10()
```

There are a lot of points in this data and there's a lot of noise, too. We might wonder how reliable this overall sample trend is. Maybe it would be better to do the trend by state.

But, but, but...

Before we pull the trigger, we have a thought: there are 50 states! If we map something like color to state how will we even make sense of the plot, let alone have a legend that will fit in the data viz?

It turns out, we have another option for mapping aesthetics: grouping. Rather than map a color or something like that to state, we can just tell ggplot to draw a different smoothed trend per state but not to add different colors or a legend. Observe:

```{r}
ggplot(Data) +
  aes(x = pop, y = black) +
  geom_point() +
  geom_smooth(
    aes(group = state)
  ) +
  scale_x_log10()
```

Whoa! That was supposed to be better. What happened? Let's try a few more updates to the code:

```{r}
ggplot(Data) +
  aes(x = pop, y = black) +
  geom_point() +
  stat_smooth(
    aes(group = state),
    se = F,
    size = 0.75,
    alpha = 0.5
  ) +
  scale_x_log10()
```

Alright, that's getting us *somewhere*, but we probably can do better still. Another option to try is to add a couple of different smoothed layers. One where we map groups to states and another that just shows the overall trend.

```{r}
ggplot(Data) +
  aes(x = pop, y = black) +
  geom_smooth(
    aes(group = state),
    se = F,
    method = "lm",
    color = "gray"
  ) +
  geom_smooth(
    se = F,
    method = "lm"
  ) +
  scale_x_log10()
```

Eh...still not great, but getting better. At the very least we can see that the positive trend between these two variables doesn't always hold up at the state level. In some states the relationship is actually negative and strongly so. Maybe we should add another tool to our toolkit to better interrogate the data. Enter `facet_wrap()`.


## Small multiples

Among the various ways that we can tell ggplot to show our data is to **facet** it by a variable, creating a "small multiple" plot.

There's a variable in our data called `census_region` that has four values indicating whether a state is in the Northeast, Midwest, South, or West. Let's see what happens when we facet by this variable:

```{r}
ggplot(Data) +
  aes(x = pop, y = black) +
  geom_point(
    alpha = 0.5,
    color = "grey"
  ) +
  geom_smooth(
    aes(group = state),
    se = F,
    size = 0.5,
    color = "gray30"
  ) +
  geom_smooth(
    se = F
  ) +
  scale_x_log10() +
  facet_wrap(~ census_region)
```

Interesting! What `facet_wrap()` has done is tell ggplot to make four different sub-plots, one for each census region in the data. We used the phrase `~ census_region` inside the function to give it instructions to facet by the values in this column in the data.

The output shows that the South is really weird. Just about everywhere else, as county population goes up, so does the share of the population that's black. But in the South the relationship goes in the opposite direction in several states.

We can add a few different customizations to our faceted plot. We might, for instance, want to give ggplot the freedom to fit the range of values shown in the different sub-plots to the data in a given region. Right now, it uses a range for the x-axis and y-axis that works for the lowest and highest values for the whole data. We can tell it `scales = "free"` to change this:

```{r}
ggplot(Data) +
  aes(x = pop, y = black) +
  geom_point(
    alpha = 0.5,
    color = "grey"
  ) +
  geom_smooth(
    aes(group = state),
    se = F,
    size = 0.5,
    color = "gray30"
  ) +
  geom_smooth(
    se = F
  ) +
  scale_x_log10() +
  facet_wrap(~ census_region,
             scales = "free")
```

We can also use `facet_grid()` similarly to `facet_wrap()`. This function let's you set easily facet by multiple categories at once.

There's actually a very helpful version of `facet_grid()` that we can access by installing and then loading the `{ggh4x}` package:

```
install.packages("ggh4x")
```

The function we want to use is called `facet_grid2()`. What I like about this function is that it permits freeing up the axis scales, an option that the original `facet_grid()` doesn't allow. 

Using this function, let's facet both by census region and whether or not the majority of people in a country voted for Trump or Clinton in the 2018 U.S. Presidential election:

```{r}
library(ggh4x)
ggplot(Data) +
  aes(x = pop, y = black) +
  geom_point(
    alpha = 0.5,
    color = "grey"
  ) +
  geom_smooth(
    se = F
  ) +
  scale_x_log10() +
  facet_grid2(winner ~ census_region,
             scales = "free_y",
             independent = "y")
```

Okay, so the South continues to look weird, but in particular counties in the South that voted disproportionately in favor of Clinton in 2018 are the weirdest. Why do you think this would be?

## Transforming data with geoms

A final thing to note about working with ggplot is that we can use it to transform data for us before it's plotted.

We've already done something like that when using `scale_x_log10()`. This transforms the x variable to the log-10 scale. Some geoms do something like this, too. For example, `geom_bar()`.

```{r}
ggplot(Data) +
  aes(x = census_region) +
  geom_bar()
```

This function returns a count of the number of observations in the data per region. In this case, an observation is a county, so it tells us the number of counties in different regions.

When working with geoms like `geom_bar()`, we can give it extra instructions for how to summarize the data. For example, say we want proportions rather than counts:

```{r}
ggplot(Data) +
  aes(x = census_region) +
  geom_bar(aes(y = ..prop..))
```

Oops! We forgot to do something. By default, `geom_bar()` takes a proportion by the x variable. What happens when you take a number then divide by that number? You get 1 of course. We need to be more explicit with `geom_bar()`:

```{r}
ggplot(Data) +
  aes(x = census_region) +
  geom_bar(aes(y = ..prop.., group = 1))
```

By adding `group = 1` we've "tricked" `geom_bar()` into grouping the data by a variable where each value is the same (in this case, 1). This makes the function return a proportion where the count per region is divided by the total count since all regions are in our new "group" where each value is the same.

Say we also were feeling artistic and wanted to add some color to the above plot. We could map colors to regions, but remember what we talked about in the first week of class about redundancy. We're going to produce a legend for our colors. That's just silly, since the legend labels are going to be identical to our x-axis labels.

It turns out we can fix that:

```{r}
ggplot(Data) +
  aes(x = census_region, fill = census_region) +
  geom_bar(show.legend = F)
```

We'll introduce some new functions later on that let us do many of these transformations and more before we even give the data to ggplot. In my opinion, writing some code to prep the data before giving it to ggplot is "better," but it's worthwhile to know that you can do some transformations of your data "under the hood" with ggplot, too.