---
title: "Showing the right numbers, part 2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Goals

- Use ggplot to summarize data.
- Use ggplot to show distributions.
- Show why transforming data *before* plotting is more efficient.


## Frequency plots for showing summaries

Some geoms transform or summarize the data for us when we plot it. We talked about using `geom_bar()` last time. This function really shines when we want to break the data down by multiple categories at once---imagine the data viz equivalent of a cross-tab.

```{r}
library(tidyverse)
library(socviz)

ggplot(gss_sm) + 
  aes(x = bigregion, fill = religion) +
  geom_bar()
```

The above summarizes the number of observations by census region in the GSS 2016 dataset from the `{socviz}` package. By mapping fit to religion, it further breakds down the count by self-identified region among survey respondents.

By updating just one option we can convert this output from a raw count to a proportion. 

```{r}
ggplot(gss_sm) +
  aes(x = bigregion, fill = religion) +
  geom_bar(position = "fill")
```

By setting `position = "fill"` the output now shows proportions by our x variable. We can now see the share of respondents by region that self-identy with a particular religion.

We could show the proportions side-by-side to make comparisons even clearer. But we have to take a few more steps than you might think. For example, say we tried using `position = "dodge`:

```{r}
ggplot(gss_sm) +
  aes(x = bigregion, fill = religion) +
  geom_bar(position = "dodge")
```

The bars appear side-by-side, but we're back to showing counts. Maybe we should try mapping `y` to `..prop..`?

```{r}
ggplot(gss_sm) +
  aes(x = bigregion, y = ..prop.., fill = religion) +
  geom_bar(
    position = "dodge"
  )
```

Shoot! That didn't work either. The final step we need to take is to map groups to `religion`, too.

```{r}
ggplot(gss_sm) +
  aes(x = bigregion, 
      y = ..prop.., 
      fill = religion,
      group = religion) +
  geom_bar(
    position = "dodge"
  )
```

Instead of mapping fill to religion, we could facet by region to avoid the need to map fill.

```{r}
ggplot(gss_sm) +
  aes(
    x = religion,
    y = ..prop..,
    group = bigregion
  ) +
  geom_bar(
    position = "dodge"
  ) +
  facet_wrap(
    ~ bigregion
  )
```


## Show distributions

Bar plots provide a useful way to show the distribution of observations across discrete categories. We can use other geoms to summarize numerical variables. 

Here's some data on fatalities from political violence events, including battles, protests, etc., in Ukraine. The data is at the month level from 2018 to 2022.

```{r}
path <- here::here("DPR 101", "Data", "ukraine_political_violence.csv")
ukraine_data <- read_csv(path)
str(ukraine_data)
```

Just like `geom_bar()` summarizes discrete data, `geom_histogram()` summarizes continuous data. For example, we can use it to summarize the distribution of fatalities in political violence events in a given month:

```{r}
ggplot(ukraine_data) +
  aes(x = Fatalities) +
  geom_histogram()
```

Histograms work by putting continuous variables into "bins" and then counting up the number of observations that fall into those bins. We can directly adjust the number of bins in a ggplot histogram. The below code sets `bins = 10`.

```{r}
ggplot(ukraine_data) +
  aes(x = Fatalities) +
  geom_histogram(
    bins = 10
  )
```
We can also group data to show its distribution by different categories. For example, by year:

```{r}
ggplot(ukraine_data) +
  aes(x = Fatalities, fill = as.factor(Year)) +
  geom_histogram(
    position = "identity",
    alpha = 0.5
  ) 
```

An alternative to binning with a histogram is a density plot. This uses a smooth function to represent the distribution of a continuous variable:

```{r}
ggplot(ukraine_data) +
  aes(x = Fatalities) +
  geom_density()
```

We can also group by different categories in the data:

```{r}
ggplot(ukraine_data) +
  aes(x = Fatalities, color = as.factor(Year)) +
  geom_density()
```

We can also modify the scales when showing continuous distributions. For example, when showing fatality data from political conflicts, applying log scales to both the x and y axis might be a good idea. Conflict fatalities are thought to follow a [power-law](https://en.wikipedia.org/wiki/Power_law) distribution which has the distinct characteristic of being linear in log-log space.

```{r}
ggplot(ukraine_data) +
  aes(x = Fatalities) +
  geom_density() +
  scale_x_log10() +
  scale_y_log10() 
```

Kinda' bumpy, but could be close to linear.

## Don't transform if you don't have to

You often can and should transform data before you give it to ggplot. This can help you avoid some of the weird quirks with using functions like `geom_bar()`

Say for example I wanted to summarize the distribution of fatalities with the Ukraine data using a method often used with power-law distributed variables.

Rather than use a histogram, I can add a new variable to `ukraine_data` that gives me the probability of a violence event deadlier than a given recorded value. I then can make a log-log plot mapping observed fatalities to x and the probability of higher fatality events to y:

```{r}
ukraine_data <- ukraine_data %>%
  mutate(prob = rank(-Fatalities) / max(rank(-Fatalities)))

ggplot(ukraine_data) +
  aes(x = Fatalities, y = prob) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = F
  ) +
  scale_x_log10() +
  scale_y_log10()
``` 

The data has the characteristic linear relationship in log-log space typical of power-law distributed outcomes.

Using already transformed data, we can also use a geom similar to `geom_bar()` called `geom_col()`. Following our text, we can replicate the column plot showing how life expectancy in the US compares to other OECD countries over time.

```{r}
ggplot(oecd_sum) +
  aes(x = year, y = diff, fill = hi_lo) +
  geom_col() +
  guides(
    fill = F
  ) +
  labs(
    x = NULL,
    y = "Difference in Years",
    title = "The US Life Expectancy Gap",
    subtitle = "Difference between US and OECD average life expectancies, 1960-2015",
    caption = "Data: OECD"
  )
```

The data `oecd_sum` already is summarized in the appropriate way to make mapping the data to aesthetics straightforward.

