---
title: "Dealing with multi-column responses"
output: github_document
---

## Intro

Sometimes with survey data you have to deal with multi-part questions. You may find these annoying from a recoding perspective, but they're essential when people can possibly fall into more than one category. This is true for variables like major or race, for example. 

Because these variables can be tricky to deal with, I thought I'd show a quick example for how you can both recode these variables and use them for data viz. The below example walks through recodes for DU student declared majors in the DU Feb. 2023 survey and how these can be used to look at variation in vaping frequency.

## Here we go!

Alright, here's an example "setup" code chunk. Check out my recodes for major(s):

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%",
                      fig.height = 4,
                      fig.width = 7)
library(tidyverse)
library(coolorrr)
library(estimatr)
set_theme()
set_palette()
library(socsci)
library(patchwork)
library(googlesheets4)
gs4_deauth()
d23 <- range_speedread("https://docs.google.com/spreadsheets/d/1tjXE5es_6kXC4HNFGlQL_xMJ80DH14nS7wNP-rd3nME/edit?usp=share_link")

## Do some recodes

d23 <- d23 %>%
  
  ## Recode columns for major(s):
  mutate(
    Q23_1 = ifelse(Q23_1 == 1, "Social Sciences", ""),
    Q23_2 = ifelse(Q23_2 == 1, "Sciences", ""),
    Q23_3 = ifelse(Q23_3 == 1, "Humanities", ""),
    Q23_4 = ifelse(Q23_4 == 1, "Arts", ""),
    Q23_5 = ifelse(Q23_5 == 1, "Interdisciplinary", ""),
    Q23_6 = ifelse(Q23_6 == 1, "Undecided", "")
  ) %>%
  
  ## Unite responses into a single column called "major"
  unite(
    major,
    Q23_1:Q23_6,
    remove = F,
    na.rm = T,
    sep = ", "
  ) %>%
  
  ## We can do the same thing with race:
  mutate(
    Q50_1 = ifelse(Q50_1 == 1, "White", ""),
    Q50_2 = ifelse(Q50_2 == 1, "Hispanic", ""),
    Q50_3 = ifelse(Q50_3 == 1, "Black", ""),
    Q50_4 = ifelse(Q50_4 == 1, "Asian", ""),
    Q50_5 = ifelse(Q50_5 == 1, "other", "")
  ) %>%
  unite(
    race,
    Q50_1:Q50_5,
    remove = F,
    na.rm = T,
    sep = ", "
  ) %>%
  
  ## Let's add some other variables:
  mutate(
    pub_sector = frcode(Q5 == 1 ~ "very unfulfilling",
                        Q5 == 2 ~ "unfulfilling",
                        Q5 == 3 ~ "neither fulfulling nor unfulfiling",
                        Q5 == 4 ~ "fulfilling",
                        Q5 == 5 ~ "very fulfilling"),
     econ_class = frcode(Q51 == 1 ~ "Lower class",
                      Q51 == 2 ~ "Lower Middle class",
                      Q51 == 3 ~ "Middle class",
                      Q51 == 4 ~ "Upper Middle class",
                      Q51 == 5 ~ "Upper class")
  ) %>%
  
  ## Let's show how to break things down by vaping:
  mutate(
    vaping = frcode(
      Q6 == 1 ~ "Never",
      Q6 == 2 ~ "Seldom",
      Q6 == 3 ~ "Sometimes",
      Q6 == 4 ~ "Often",
      Q6 == 5 ~ "Very Often"
    )
  )
```


I can use the major(s) columns to summarize vaping in a couple different ways. Way 1 gives you a really busy but precise data viz. Way 2 gives you a more concise but also slightly less precise data viz. 

### Way 1: By the `major` column

```{r fig.height=8, fig.width=12}
d23 %>%
  group_by(major) %>%
  ct(vaping, show_na = F) %>%
  drop_na() %>%
  filter(major != "") %>%
  ggplot() +
  aes(
    x = vaping,
    y = pct
  ) +
  geom_col() +
  facet_wrap(~ major) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    x = NULL,
    y = "Share of Students",
    title = "Figure 1: Vaping by declared major(s)"
  ) +
  scale_y_continuous(
    labels = scales::percent
  )
```

## Way 2: Being less specific

```{r}
d23 %>%
  select(Q23_1:Q23_6, vaping) %>%
  pivot_longer(
    cols = - vaping
  ) %>%
  group_by(value) %>%
  ct(vaping, show_na = F) %>%
  drop_na()%>%
  ggplot() +
  aes(
    x = pct,
    y = value,
    fill = vaping
  ) +
  geom_col(
    position = position_fill(reverse = T)
  ) +
  ggpal(aes = "fill") +
  labs(
    x = "Share of students",
    y = NULL,
    title = "Figure 2: Vaping by area of study"
  ) +
  scale_x_continuous(
    labels = scales::percent
  )
```

