---
title: "Data Challenges"
author: "Miles D. Williams"
date: '2023-01-31'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## DC 1

Make a simple scatter plot using this data.

Data: https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20101/Data/DC1_da
ta.csv

Complete your code in a .Rmd file. Knit your code to a Word document and submit the document.


## DC 2

**What I did**

For this challenge, I'm having students summarize the distribution of cabinet turnover for different presidents. First, I want to condense the data to the presidential level. Then I'll save the new data and have them use that:

```{r}
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/cabinet-turnover/cabinet-turnover.csv"
Data <- read_csv(url)
```

```{r}
Data %>%
  filter(
    president %in% c("Reagan","Bush 41", "Bush 43", "Trump"),
    position %in% c(
      "Secretary of the Treasury",
      "Secretary of State",
      "Secretary of Defense",
      "Attorney General"
    )
  ) %>%
  group_by(president, position) %>%
  summarize(
    mean_tenure = mean(days, na.rm=T)
  ) -> new_data

library(here)
write_csv(new_data, here("DPR 101", "Data", "cabinet_turnover.csv"))
```


**What they'll do**

Here's some data on the average tenure of members of U.S. Presidential cabinets from Carter to the first couple years of Trump's presidency. 

Data: https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20101/Data/cabinet_turnover.csv

Make a data viz showing how mean tenure differed across presidents. In your figure, map an aesthetic to the cabinet position and set at least one other aesthetic. 

Knit your work to a Word document and submit it.


**Example winning answer**

```{r}
# open the tidyverse
library(tidyverse)

# read in the data
url <- "https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20101/Data/cabinet_turnover.csv"
the_data <- read_csv(url)

# make a column plot
ggplot(new_data) +
  aes(x = president,
      y = mean_tenure,
      fill = position) +
  geom_col(
    color = "darkblue"
  )
```


## DC 3

For this data challenge, use the Ukraine event-fatality data to show how the number of events has changed or not changed from 2018 to 2022.

Do you think a histogram that groups the data by year, or a line plot is a more effective way to show the data? Make both kinds of plots and write a short paragraph explaining which you prefer.

Here's a link to the data: https://github.com/milesdwilliams15/Teaching/blob/main/DPR%20101/Data/ukraine_political_violence.csv


## DC 4

Use the county_map_data we made in the notes here [link here].

1. Filter the data down to a state of your choice.
2. Make a map that shows county level detail about voter turnout in 2016. [You'll need to set the mapping fill = 100 * (vote_total_2016 / pop).]
3. Knit your work to an HTML or word and submit to Canvas.

Read in the data:
```{r}
library(tidyverse)
library(socviz)
county_map_data <- left_join(county_map, county_data)
glimpse(county_map_data)
```

Filter to look at Ohio:
```{r}
oh_map_data <- county_map_data %>%
  filter(state == "OH")
```

Draw an Ohio map:
```{r}
ggplot(oh_map_data) +
  aes(x = long, y = lat, 
      group = group,
      fill = 100 * (total_votes_2016 / pop)) +
  geom_polygon(
    color = "gray",
    size = 0.1
  ) +
  coord_fixed() +
  theme_void() +
  scale_fill_gradient(
    low = "white", high = muted("blue")
  ) +
  labs(
    fill = "% Turnout",
    title = "Ohio voter turnout, 2016",
    caption = "By Miles D. Williams"
  ) +
  theme(legend.position = c(0.3, 0.9),
        legend.direction = "horizontal")
```

