---
title: "Data Challenges"
author: "Miles D. Williams"
date: '2023-01-31'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## DC 1

Make a simple scatter plot using this data.

Data: https://raw.githubusercontent.com/milesdwilliams15/Teaching/main/DPR%20101/Data/DC1_da
ta.csv

Complete your code in a .Rmd file. Knit your code to a Word document and submit the document.


## DC 2

**What I did**

For this challenge, I'm having students summarize the distribution of cabinet turnover for different presidents. First, I want to condense the data to the presidential level. Then I'll save the new data and have them use that:

```{r}
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/cabinet-turnover/cabinet-turnover.csv"
Data <- read_csv(url)
```

```{r}
Data %>%
  group_by(president, position) %>%
  summarize(
    mean_tenure = mean(as.numeric(length), na.rm=T)
  ) -> new_data

library(here)
write_csv(new_data, here("DPR 101", "Data", "cabinet_turnover.csv"))
```


**What they'll do**

Here's some data on the average tenure of members of U.S. Presidential cabinets from Carter to the first couple years of Trump's presidency. Make a data viz where you map an aesthetic to the president and set at least one other aesthetic. 

Knit your work to a Word document and submit it.


**Example winning answer**

```{r}

```






