---
title: "GC3"
author: "Miles D. Williams"
date: "2023-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F)
```

## Prompt

In this graph challenge, your objective is to make a heat map for any one of the 50 states (except Alaska) in the US showing county-level detail of the 2016 GOP candidate's (Trump) electoral margin relative to the Democratic candidate's (Clinton). This is not a variable in the data, so you will need to construct it with two other variables, specifically: "per_gop_2016" and "per_dem_2016". The first is the proportion of votes in a county that Trump received. The second is the proportion of votes in a county that Clinton received. Make sure that your map has an informative title and that you update your plot with the "theme_void()" function.

After you produce your figure, discuss in a short paragraph why you think some areas of your chosen state are bluer (Trump's margin was less than Clinton's) and some areas are redder (Trump's margin was greater than Clinton's).

Here's some code to get your started. It will open the packages you need and create a county-level dataset that is merged with the appropriate county boundary coordinates. It then filters the data down to a state. You just need to edit the part that says "State acronym here" to tell R which state you want to look at. Once you have the data, it's up to you to add a new column of Trump's vote margin relative to Clinton's before plotting.

```
## open packages
library(tidyverse)
library(socviz)

## merge county data to county boundary coordinates
cty_map_data <- left_join(county_map, county_data)
st_map_data <- cty_map_data |>
  filter(state == "State acronym here")
```

## read in the packages

```{r}
library(tidyverse)
library(socviz)
```

## merge and filter the data

```{r}
cty_map_data <- left_join(county_map, county_data)
st_map_data <- cty_map_data |>
  filter(state == "AK")
```

## plot

show the change in the GOP or Dem vote share between 2016 and 2012:

```{r}
st_map_data |>
  mutate(
    gop_margin = 100 * (per_gop_2016 - per_dem_2016)
  ) |>
  ggplot() +
  aes(
    x = long,
    y = lat,
    group = group,
    fill = gop_margin
  ) +
  geom_polygon(
    color = "black",
    size = 0.05
  ) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_void() +
  coord_fixed() +
  labs(
    fill = "% point change",
    title = "Trumps 2016 vote margin in 2016"
  )
```

