---
title: "Data and Data Bases"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = F,
                      message = F,
                      dpi = 500,
                      out.width = "75%")
```

## Learning Objectives

- Describe what a relational database is, and why you might need to store your data in one.
- Identify the different parts of a relational database (tables, key, query).
- Know that Structured Query Language (SQL) is the way to "talk" to relational DBs.

Example database, soccer from Kaggle: https://www.kaggle.com/hugomathien/soccer 

## What is a relational database?

In class, we've been working with relatively small datasets that fit easily into our computer's memory. But what about *really* big datasets? These are often better stored outside of R and inside databases. Using R, we can connect to a database, and retrieve only the parts of the data that we need for a given analysis. But there are many other reasons why you might want to store your data in a relational DB.

A relational DB is a set of linked tables. Each of the tables are linked by at least one column between each one that can help them "relate" or bring the data back together. These links are defined relationships, that explicitly state how each table relates to which other table. A specialized command language called (Structured Query Language, SQL [see-kwel]), is usually used to manipulate data held in relational databases.

Benefits of a relational database are:

- Can break a large complex dataset into digestible pieces (each table is a cohesive "unit" of the data), for better and less repetitive storage.
- Can use (and store) a "query" (like asking the database a question) to find exactly which parts of the data you need to use (e.g. only the basketball games that are home, and not away).
- You must specify a data type for each field (e.g. you can’t mix numbers and text in a single column).
- It is not possible to sort a column independently of other columns (this is possible in excel and can cause problems!).
- Is often a better way to store very large datafiles, especially ones you plan to continue to grow and maintain long-term.
- Widely used to store and maintain lots of existing data.
- Fast processing of large amounts of data.
- Improve quality control of data entry.
- Set valid data type and value constraints.
- Often possible to set up data entry forms (e.g. Access, Excel, Filemaker).
- Can use quality control scripts to test entered data.
- Being able to conceptualize a relational database can aid in understanding data manipulation in other languages (e.g. SQL is similar to dplyr).

Take a real-life example of something that is stored in a relational database (Amazon orders, names/addresses, payment methods, etc). When you order something from a company online, they need to store a lot of different information about you to process that order. This is a simplified version of how the data might be organized (Note: this is how the database schema would look in MS Access).  

```{r, echo=FALSE}
knitr::include_graphics(
  here::here("DA 101", "Notes", "amazon_relational_db.JPG")
)
```

What do the "keys" mean? What about the lines? The numbers and infinity symbols?

Each table must have at least one column that only contains *unique* values, nothing is allowed to be repeated. So this is often some sort of ID number, or code. Or it could be someone’s name, as long as those names are not repeated. This is known as the primary key and is represented by the yellow key symbol next to a column name. You can see that the lines connect each primary key to another field in another table. This is the information that can be used to connect tables, and bring them back together. So information could be a primary key in one table, but not in another. 

So, while CustomerID is the primary key in the Customer’s table, it is the foreign key in the Orders table. A variable could be both the primary key in it’s table, and the foreign key that matches to another table. (e.g., the nycflights13 example, where origin is primary key in weather table, but foreign key for the airport table.)

Tables in a Relational Data Base (RDB) *should* really have a primary key. If it doesn’t you could make one (sometimes called a surrogate key), by simply adding a column that is “rowID”. This could just number each of the rows 1 to whatever the total number of rows is, so you have a set of unique values to match to and identify each row. 

Typically, relations are one-to-many (1 to $\infty$), but they can be many-to-many ($\infty$ to $\infty$), or many-to-1 ($\infty$ to 1). 

What are the relations in the example table shown? Are they all 1-to-many relations?

- Q: Why would CustomerID not be a primary key in the Orders table? 
- A: Because a customer could place more than one order, so it would necessarily be repeated in the orders table.

Here’s another way to think through the logic of how these tables are structured, and why they are stored in a relational set of data tables.

Our basic business logic can be described in the following points:

- All Customers can place 1 or more orders.
- Each Order can only be attached to 1 Customer.
- Each Order will have 1 or more Order Detail records associated with it.
- Each Order Detail will deal with only 1 Product.
- A Product can of course be found on multiple Order Details because many customers may want to order the same products.

Does this seem confusing? What would this database look like if we put all the information in a single table? How many columns would we need? Take some time to make sure most students are following along here. Ask for student volunteers to help me draw up what the data table would look like.

And what we’ve been looking at is actually a pretty simplified version of what this kind of database could contain. Here’s a more realistic version.

```{r}
knitr::include_graphics(
  here::here("DA 101", "Notes", "amazon_big_relational_db.JPG")
)
```

(example from https://bohr.wlu.ca/rhenderson/cp102/labs/about_databases2.htm)

Does this help make it more clear why you *really* wouldn’t want to enter, sort, or load this data as a single large flat file?

Relational databases have much in common with the "rules" for flat datasets (e.g. an excel file). They’re just a little bigger. Kind of like the difference between writing a paragraph, and an essay. (There is a lot similar between data and coding and writing). 

Relational database rules:

- Data is stored in tables.
- One table per type of data.
- Tables can be linked together to combine information.
- Each row contains a single record.
- A single observation or data point.
- Each column or field contains a single attribute.
- A single type of information.
- The concepts of a relational database management are core to understanding how to do similar things using programming languages like R or Python.


## Try it out

For an extra challenge, install one of these data packages, and explore how the database is set up.

Here's what the code would look like:

```
install.packages("Lahman")
library(Lahman)
```

Here are your options:

- Lahman
- babynames
- nasaweather
- fueleconomy
- nycflights13

Hint: If you type databasename:: then the tab complete will show you what the tables are. You could also google the package to see a description of the tables.

- For your group’s database, what are the tables? 
- Choose one or two tables and identify the keys in those tables.
- What are the relationships among the tables? Can you draw a diagram?
