---
title: "Data Visualization"
output: 
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")
```

## Goals

- Identify different kinds of graphs and the data inputs to produce them.
- List rules of thumb for deciding what kind of graph is the most useful.
- Solve common problems, like making color graphs readable for colorblind people.
- Practice using ggplot to make figures.
- Practice layering different aspects of data in ggplots.
- Learn about some add-on packages that work with ggplot.

For some additional reading, check out:

> [Data Visualization: A Practical Introduction](http://socviz.co/index.html#what-you-will-learn)

I use this book to teach my intro to data visualization course. 


## Why data viz?

Data visualization plays an integral part of data analysis. It is generally a bad idea to jump right into analysis before **looking at the data**.

There's a famous example of why looking at the data is so important---it's called [Anscombe's Quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet). The below figure shows the relationship between two variables, x and y, but with four different versions of the data. Each panel shows a scatter plot along with a linear regression line showing the estimated trend. Even though the data are very different, the trend is identical. In fact, in all four versions of the data x and y have the same estimated correlation of 0.82 (that's a really strong positive correlation!).

```{r, echo=FALSE}
library(tidyverse)
library(coolorrr)
set_theme()
anscombe %>%
  pivot_longer(
    cols = everything()
  ) %>%
  mutate(
    var = str_remove(name, "[1-4]"),
    group = str_remove(name, "[x-y]")
  ) %>%
  select(-name) %>%
  pivot_wider(
    names_from = var,
    values_from = value
  ) %>%
  unnest() %>%
  ggplot() +
  aes(x = x, y = y) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = F
  ) +
  facet_wrap(~ group, ncol = 2) +
  labs(
    title = "Anscombe's Quartet",
    caption = "All variables have the same correlation of 0.82"
  )
```

This example highlights why looking at data is so essential to drawing good inferences. Each version of the data tells a different story, but if we had jumped straight to analysis we may have missed this.


## Fundamentals of Good Data Viz

The difference between good and bad data visualization comes down to three things:

1. **Good taste:** We want to avoid a lot of "chartjunk" or unnecessary plotting elements.
2. **Good data:** Bad data can give misleading results.
3. **Good perception:** Subjective human perception can get in the way of objective insights.


### Chartjunk

Chartjunk is any kind of design element or elements that don't communicate anything new or useful. In fact, chartjunk often misleads an audience about the data. Think about things like 3-D bar plots.


### Bad data

Data itself can lead people astray. Someone might convert a continuous scale into categories to make a trend look worse than it actually is. Or they might use variables where one or both have not been appropriately measured, either due to faulty instruments or human error.


### Bad perception

In addition to some 3-D plots that make it hard to gauge values, some 2-D plots can create problems of perception, too. Pie charts are a notorious example, but even a simple matter of stretching or condensing a plot can distort the data.


## Make good choices

Use these three principles to help you make good choices about how to show a relationship, or distribution, or difference using data viz. Getting all the elements right requires thinking about a few things.


### 1. What's your point?

Be clear about your intent. Do you want to explore the data to better understand it? Or, do you want to explain a relationship or make a point for an audience? You need to think about:

- What you want people to understand.
- What you want people to walk away with.

Never will you visualize **all** of the data. You need to consider: what is the least I can show to maximize the impact of my message?


### 2. What's the right plot?

Different kinds of plots are meant to describe different kinds of things. Which you should use will depend on some details about your data and what you want to show.

- Is your data numerical or categorical?
- Do you want to show a distribution? A relationship? A comparison?
- Do you want to show all the data or break it down to smaller chunks?

You've already gotten some experience using a few different plots in class:

- Histograms.
- Scatterplots.
- Boxplots.
- Barplots.

Some other options include:

- Heatmaps
- Layered or stacked bar graphs.
- Maps

With each kind of plot, you need to follow some guidelines based on the way humans perceive information:

- Position data on a common scale.
- Bars are better than circles or squares to communicate size.
- People discern color better than shapes in scatter plots.
- Avoid pie charts.
- Avoid 3-D plots.
- Use small multiples to show changes in data.

Here are some basic rules of thumb when deciding what kind of plot to use:

- Line charts track changes or trends over time and/or show the relationship between two or more variables.
- Bar charts compare quantities (like counts or proportions) of different categories.
- Scatter plots show the relationship (joint variation) of two numerical variables.
- Pie charts...don't use pie charts. Some people think pie charts are good for showing proportions. Just use a bar chart instead.
- Use color, shape, or size to show additional information in a 2-D plot based on other categories or variables in the data.


### 3. Less is more (but avoid extreme minimalism)

Don't overcomplicate your data viz. Be creative, but make sure an style elements you add don't obscure the data.

This is the principle of "minimizing the ink-to-data ratio."

Don't use redundant elements, like a color, labels, and a legend to highlight the same categories.

But don't go too extreme in an effort to ruthlessly eliminate ink from your data viz. Research shows that going to minimalist can also interfere with human perception of results.


### 4. Use color intentionally

Use color to draw your audience's attention to certain points.

- If you don't have a good reason to color your histogram blue, don't do it.
- If you don't need to have each bar in your bar chart be a different color of the rainbow, don't do it.
- Do use color to differentiate groups in the data.
- Do use color to make a scatter plot with a regression line easier to see.


### 5. Use good titles and text

Use sensible and informative titles, axis labels, and annotations. Tell people what the data is, tell them about the relationship it shows, tell them what the variables are or what the categories mean, etc. Use active voice as much as possible.


### 6. Get feedback and try again

Producing a data visualization is never a one and done exercise. It requires multiple attempts to get it right. It usually helps to have others look at your data viz along the way for a more objective take on how an audience might perceive the data. You know what relationships or distributions you want to show, so your figure's insights seem obvious to you. But do they look obvious to someone else?


### 7. Look at other data viz and copy

R is open source, and lots of people use it for data viz. Look online to find examples that people have posted or created for ideas. Often these examples come with code that you can copy and then modify for your own data.


## Common mistakes

Mark Twain popularized the phrase:

> There are three kinds of lies: lies, damned lies, and statistics.

He attributed this to Benjamin Disraeli, a former British Prime Minister.

Not to put too fine a point on it, data viz provides all the tools necessary to lie quite effectively.

Scaling matters. Take the below figures. Both panels show the same data over time: \% growth from 2010 to 2013. However, the left panel shows this relationship using a y-axis that only extends across the range of values growth takes over this time period. The right panel uses a y-axis that instead starts at zero. The first version obviously creates a the impression of more dramatic growth.

```{r, echo=FALSE}
Data <- tibble(
  x = 2010:2013,
  y = (47:50)/100
)
library(patchwork)
p1 <- ggplot(Data) +
  aes(x = x, 
      y = y) +
  geom_line(size = 1) +
  scale_y_continuous(
    labels = scales::percent
  )
p2 <- ggplot(Data) +
  aes(x = x,
      y = y) +
  geom_line(size = 1) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, .5)
  )
p1 + p2 +
  plot_annotation(
    title = "Same Data, Different Scale"
  ) &
  labs(x = NULL,
       y = NULL,
       subtitle = "Growth over Time") 
```


How about the below figure? Does the use of color do anything for us? It shows the average life expectancy of individuals from different continents from 1952-2007. Each bar in the chart has a different color.

```{r, echo=FALSE}
Data <- gapminder::gapminder %>%
  group_by(continent) %>%
  summarize(
    lifeExp = mean(lifeExp)
  )
ggplot(Data) +
  aes(x = lifeExp,
      y = continent,
      fill = continent) +
  geom_col(show.legend = F) +
  labs(
    x = "Average life expectancy in years",
    y = NULL,
    title = "Life Expectancy by Continents, 1952-2007"
  )
```

Does the next figure use color more effectively? This data viz uses data at the country level for 2007 to show the relationship between GDP per capita (a common measure of average income in a country) and average life expectancy. It uses color to show groups of countries by region. It also uses point size to show population size.

```{r, echo=FALSE}
Data <- gapminder::gapminder %>%
  filter(year == 2007)
fig <- ggplot(Data) +
  aes(x = gdpPercap,
      y = lifeExp,
      color = continent,
      size = pop) +
  geom_point(
    alpha = 0.5
  ) +
  scale_x_log10(
    labels = scales::dollar
  ) +
  scale_size(guide = "none") +
  labs(
    x = "GDP per capita (log-10 scale)",
    y = "Average life expectancy (years)",
    title = "Higher GDP per Capita = Higher Life Expectancy, 2007",
    color = NULL
  ) +
  annotate(
    geom = "text",
    x = 10000,
    y = 65,
    label = "Size = Population"
  )
fig
```

## A note on colorblindness

Use of color is good only in so far as your audience can make sense of it. But not everyone can. 

There's a great package called `{colorblindr}` that you can use to test how your color choices would look to someone who is colorblind.

Here's how to install it:

```
remotes::install_github("clauswilke/colorblindr")
```

Here's an example of how it works using the last scatter plot I produced where I used color to group points by regions. Clearly, some kinds of colorblindness prevent people from distinguishing all five categories when using ggplot's default color options.

```{r}
library(colorblindr)
cvd_grid(fig + labs(x = NULL, y = NULL, title = NULL) +
           theme(legend.position = "none") +
           scale_x_log10(
             breaks = c(1000, 10000)
           ))
```

Using contrast in addition to color can be an effective way to deal with some kinds of color blindness. Even if people can't tell apart colors, they can see that some things are lighter and others darker.