---
title: "Using tools in `{dplyr}`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")
```

## Learning Objectives

- Apply the main dplyr functions to manipulate a dataset
- Layer complexity by chaining together multiple dplyr functions to solve a problem.


## Using dplyr for data manipulation/wrangling

Dplyr is a powerful tool for letting us transform a dataframe into what we need for analysis.

There are 6 key dplyr functions that we need to know (and much more that it can do!)

- `filter()`: Choose observations based on their values (==, !=, >, <, >=, <=)
- `arrange()`: Reorder the rows (sort based on a variable)
- `select()`: Choose variables (columns) based on their names
- `mutate()`: Use existing variables to create new variables
- `summarize()`: Collapse many variables into a single summary 
- `group_by()`: Specify the scope of the function that you want to perform (for example, calculate the average height of plants, grouped by each region). ORDERING MATTERS!

To use these functions, don't forget to tell R to attach the dplyr package:

```{r, warning=FALSE, message=FALSE}
library(dplyr)
```


These are all dplyr functions, sometimes we call them “verbs”, and they all take input in a similar structure.

`filter(data, colname == "value" & colname >= 1000)`

We can chain together multiple dplyr “layers”, like ggplot, but we use the %>% pipe instead of +.

1. First, we start with the data frame
2. Then, we add arguments that describe what to do with the data frame, using the column (variable) names. Note that we don’t need to put the column names in quotes, because we’ve already specified the data frame that they are contained in, R will be able to find them as a “known” object. 
3. The output is a new data frame

Do you recall using `filter()` in our Garlic Mustard lab? What do we need to add if we want to save the output so we can do things with it later? 

```
filtered_data <- data %>%
	filter(column_name == "some text")
```

To filter, we need to remember and know how to use the comparison operators: 

- Greater than `>`
- Less than `<`
- Greater than or equal to `>=`
- Less than or equal to `<=`
- Equal to `==`
- Not equal to `!=`
- In `%in%`
- Is NA? `is.na()`

Aside: Recall some common mathematical operators (should follow order of operations).

- addition `+`
- subtraction `-`
- multiplication `*`
- division `/`
- square `^2` 

We also need to know the logical operators

- and `&`
- or `|`
- not `!`


So let’s try using filter with the NHANES dataset, which is another datapackage we can install and look at. This is an excerpt from the Center for Disease Control’s (CDC) National Health and Nutrition Examination Survey (NHANES), a national study conducted every two years that gathers data on American’s health, nutrition, and exposure to environmental chemicals. Data from NHANES is used widely by environmental health scientists to study the U.S. public’s exposure to chemicals.

First install the NHANES package:

```
install.packages("NHANES")
```

Then attach it. We can use dim() to check the size of the data (number of rows and columns).

```{r, message=FALSE, warning=FALSE}
library(NHANES)
dim(NHANES)
```

Let’s take a quick look at this dataset - each row is the record for a person who responded to the survey, and each column is the variables that they responded to. You can see that there are a lot of variables we can sort or filter on!

### filter()

Let’s just practice filter first, which we’ve already seen in lab. Let’s say we want all the data from men.

```{r}
NHANES %>%
  filter(Gender == "male")
```

OK but what if we want to filter on two variables, males who are also married?

```{r}
NHANES %>%
  filter(Gender == "male" & MaritalStatus == "Married")
```

OK but what if we want to filter on two variables, college graduates or homeowners?

```{r}
NHANES %>%
  filter(Education == "College Grad" | HomeOwn == "Own")
```

What does this give us? Rows are college grad & homeowners, and college graduates only, and homeowners only. 

Note: `filter()` only includes rows where the condition is `TRUE`. So if we want to keep in NA (missing) values, then we need to ask for them explicitly, for example:

What would we do if we wanted only rows where values in a column were blank (indicated by NA on the computer). Would this work?

```
NHANES %>%
  filter(Education == NA)
```

Nope! Since NA is a special value that indicates missing data, it cannot be equal to anything, or not equal to anything. This means we can only include or exclude NA using functions designed to do that. For example, is.na() or na.rm=TRUE. The example below would work to find all rows where Education was either “College Grad” OR was missing.

```{r}
NHANES %>%
  filter(Education == "College Grad" | is.na(Education))
```


### arrange()

We haven’t seen this yet, but it works similarly to `filter()`, except instead of selecting rows, it changes their order - Kind of like using "sort" in Excel, except you know that you won’t accidentally sort just the one row and scramble all your data. 

Let’s sort the NHANES dataset by chronological age:

```{r}
NHANES %>%
  arrange(Age)
```

OK, cool, but what if we want to see the oldest people first? We can specify the order, and change the default:

```{r}
NHANES %>%
  arrange(desc(Age))
```

But what if we really want to use the age category? Do you think we can sort on a factor or character? Let’s try! We can try anything in R...

```{r}
NHANES %>%
  arrange(AgeDecade)
```

Awesome. Where do you think the NAs went? How would we check? 
Hint: NAs are always sorted to the end, whether you use ascending or descending order.

### select()

Here’s another one that we’ve seen briefly - remember, we used select to choose only the columns that we were interested in. It’s another way to “slice” our dataframe. Usually, we don’t need ALL the variables for a given data analysis, so it is more efficient to just look at the ones that we need. 

How many columns does NHANES have? Use ncol()

```{r}
ncol(NHANES)
```

Let's just get the ones we want.

```{r}
NHANES %>%
  select(Gender, Age, Weight, Height, Diabetes)
```

What if we wanted most of the columns, but just not the ones related to sexual health? Wouldn’t it be terrible to have to type ALL those names? Even just typing all the sexual health columns (7) would be a pain! Luckily, we don’t have to!

```{r}
NHANES %>%
  select(-(SexEver:PregnantNow))
```

We can even use some “helper functions” inside of select() to get even more specific. A few examples are:

- `starts_with("")` 
- `end_with("")`
- `contains("ijk")`

So let’s say we know that we have a bunch of columns for sexual health, but we don’t want to write them all, or look them up, even though we need them for analysis. We can be lazy and just say we want all the columns that contain “Sex” in the name!

```{r}
NHANES %>%
  select(contains("Sex"))
```


### mutate()

Sometimes we need to create new variables. This might be a function of an existing column. For example, in last week’s lab, some people recorded the data twice, in two columns. They could have used mutate() to take the average of those two columns, and add it to their dataframe. 

This is our fourth dplyr command - let’s start chaining some of these together to see how they work! Here’s one that we did before, to select just 5 columns:

```
NHANES %>%
  select(Gender, Age, Weight, Height, Diabetes)
```

To chain on another command, we use the pipe `%>%`, just like in ggplot where we added layers to our graphic using `+`. Let's make our own calculation of BMI. First we need to convert our weight and height to kilograms and centimeters.


Here's an example making a column called `Height_m`:

```
NHANES %>%
  select(Gender, Age, Weight, Height, Diabetes) %>%
  mutate(Height_m = Height / 100)
```

Here's one showing how you can build on new columns you make in `mutate()` within the same call to the function:

```
NHANES %>%
  select(Gender, Age, Weight, Height, Diabetes) %>%
  mutate(Height_m = Height / 100,
         BMI = Weight / Height_m ^ 2)
```

See above, we can even do this all at once, by referring to columns that we have just created!

Or we can do it all in one equation, if we don’t need to save out a column for height in meters.

```
NHANES %>%
  select(Gender, Age, Weight, Height, Diabetes) %>%
  mutate(BMI = Weight / (Height / 100) ^ 2)
```

Note that as things are getting more complex, we just did one step at a time, and then went back to edit what we just did, making sure each step along the way worked before making it more complicated. This is a good recipe for success, and to cut down on frustration!

You could use any mathematical operators, here, including `sum(x)`, `y - mean(y)`, `log()`, `log2()`, `log10()`, `lead()`


### summarize() and group_by()

This can be used to collapse a dataframe to a single row, or a set of rows (if we use group_by()). For example, here's the mean height:

```{r}
NHANES %>%
  summarize(
    mean = mean(Height, na.rm=T)
  )
```

And here's mean height when grouping by gender:

```{r}
NHANES %>%
  group_by(Gender) %>%
  summarize(mean(Height, na.rm=T))
```

OK, but this includes all the ages, which probably doesn’t make sense. All those babies are bringing the heights down. Can you calculate the average heights of adult men and women in feet?

```{r}
NHANES %>%
  filter(Age > 18) %>%
  mutate(Height_ft = (Height * 0.39 )/12) %>%
  group_by(Gender) %>%
  summarize(mean(Height_ft, na.rm=TRUE))
```

Finally, we also have data on diabetes, can we include that too? How many men and women have diabetes?

```{r}
NHANES %>%
  filter(Age > 18 & Diabetes == 'Yes') %>%
  mutate(Height_ft = (Height * 0.39 )/12) %>%
  group_by(Gender) %>%
  summarize(AvgHeight = mean(Height_ft, na.rm=TRUE),
            NumDiabetic = n())
```

Adding a column for count to go along with your summary can also be helpful, because it can help you see how many things were averaged (or whatever). You don’t want to draw conclusions from a very small number of samples!

Note: We can also calculate median (midpoint of the data), which tells us a little different information from mean (average)

```{r}
NHANES %>%
  filter(Age > 18 & Diabetes == 'Yes') %>%
  mutate(Height_ft = (Height * 0.39 )/12) %>%
  group_by(Gender) %>%
  summarize(MedianHeight = median(Height_ft, na.rm=TRUE),
             	NumDiabetic = n())
```

There a lot of different functions you can use:

- `mean()`: average
- `median()`: midpoint
- `sd()`: standard deviation
- `IQR()`: interquartile range
- `mad()`: median absolute deviation
- `min()`: minimum
- `max()`: maximum
- `quantile(x, 0.25)`: value in the data that is > 25% of the values and < 75% of the values
- `n()`: count
- `sum(!is.na(x))`: count all non-missing values (don’t count NAs)
- `n_distinct()`: count all distinct (unique) values


## From summarizing to plotting

Sometimes it will make sense to do a combination of `summarize()` and `group_by()` with your data before using `ggplot()` to plot relationships.

Say we wanted to create a column plot comparing mean height for adults by reported gender. We would first summarize heights by gender and save the output as a new object:

```{r}
adult_heights <- NHANES %>%
  filter(Age > 18) %>%
  mutate(Height_ft = (Height * 0.39 )/12) %>%
  group_by(Gender) %>%
  summarize(mean_height = mean(Height_ft, na.rm=TRUE))
```

And then we would use this data to make a figure:

```{r}
library(ggplot2)
ggplot(adult_heights) +
  aes(x = Gender,
      y = mean_height) +
  geom_col(width = 0.5) +
  labs(
    x = "Gender",
    y = "Average height (ft.)",
    title = "Difference in mean height by gender"
  )
```

