---
title: "Functional Programming in R"
output: 
  github_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F,
                      dpi = 500,
                      out.width = "75%")
```


## Goals

- We can make functions in R
- Making functions makes it easier to do repeated tasks or routines.
- When we make functions, make sure you document what you're doing and why. Your future self and other collaborators will thank you.


## Defining a function

The basic syntax for creating a function in R looks like this:
```
fun_name <- function(x) {
  newx <- operation1(x)
  newx <- operation2(newx)
  newx # returned object
}
```

- You'll assign your function a name
- Specify inputs and other options for the function using `function()`
- You'll write the code that the function will apply given your inputs
- You'll specify what object or objects your function will return

Here's an example of a function that "demeans" a numeric variable. To demean means or mean center a variable means that we update a variable so that it has a mean of zero.

```{r}
demean <- function(x) {
  cx <- x - mean(x, na.rm=T)
  cx # return the results
}
```

Say we have the following numerical vector:
```{r}
my_nums <- 1:10
```

We can demean it by using our `demean()` function:
```{r}
demean(my_nums)
```

Notice that the output is now such that the mean of the variable is zero and its min. and max. values have been adjusted accordingly.
```{r}
dm_nums <- demean(my_nums)
mean(my_nums)
mean(dm_nums)
```


## Composing functions in succession

We can compose functions in succession. This is a nice strategy when we want to create more complicated functions/routines.

For example, let's create a function that converts a numeric vector into z-scores:
```{r}
zscore <- function(x) {
  zx <- demean(x) / sd(x, na.rm=T)
  zx # return the results
}
```

z-scores are values that have been re-scaled to have mean of zero and variance of one. To create them, we simple demean a variable and then divide it by its standard deviation. That's just what the above function does. But notice that to demean it uses the `demean()` function that we already created.

See it in action:
```{r}
cbind(
  x = my_nums,
  cx = demean(my_nums),
  zx = zscore(my_nums)
)
```

To check that the `zscore()` function worked, we can write:
```{r}
mean(zscore(my_nums)) # should = 0
sd(zscore(my_nums))   # should = 1
```


## Setting defaults

As we start to compose more complex functions, we can begin to think about setting defaults. 

Often when we want to re-scale a variable, we typically either want to convert it to z-scores or to the unit interval (0-1). The function `rescale()` created below will do either transformation, depending on whether `zscore = TRUE`. Check it:
```{r}
rescale <- function(x, zscore = TRUE) {
  if(zscore) {
    newx <- zscore(x)
  } else {
    newx <- (x - min(x, na.rm=T)) /
      (max(x, na.rm=T) - min(x, na.rm=T))
  }
  newx # return
}
```

The default option is `zscore = TRUE`, meaning that unless we specify otherwise, the function will convert a numeric variable to z-scores. But if we set it to `FALSE`, it'll scale the variable to the unit interval:
```{r}
rescale(my_nums)
rescale(my_nums, zscore = F)
```


## Documenting your work

As you compose new functions, remember that while your programs are written in R, you want them to be legible to other humans. That's why you should always make sure that you document your new functions to tell your future self or others what the function does and why.

Here are the above examples in a single chunk with documentation:

```{r}
## Creates a function that "demeans"
## a variable (ie - sets the mean to zero)
demean <- function(x) {
  # take the input x and subtract the mean
  # since na.rm=T, you don't have to wory about NAs
  cx <- x - mean(x, na.rm=T)
  cx # return the results
}

## Creates a function that converts
## a variable to z-scores with mean
## of zero and variance of one.
zscore <- function(x) {
  # demean x in the numerator and take the
  # SD in the denominator
  zx <- demean(x) / sd(x, na.rm=T)
  zx # return the results
}


## Creates a function that rescales
## a variable either to z-scores or to
## the 0-1 scale.
rescale <- function(x, zscore = TRUE) {
  ## Use simple logic to return either
  ## the z-score or scaled to [0, 1]
  if(zscore) { 
    ## If zscore = TRUE
    newx <- zscore(x)
  } else {
    ## if zscore = FALSE
    newx <- (x - min(x, na.rm=T)) /
      (max(x, na.rm=T) - min(x, na.rm=T))
  }
  newx # return
}
```

There is no formal set of rules for how to document your code, but some good rules of thumb are:

- Explain the overall purpose of the function
- Explain the sequence of steps the function takes "under the hood" --- i.e., inside the `{ }`s

## Challenge

Check out the following routine. Can you create a function called `lm_summary()` that will implement the following routine for any linear model and dataset, returning the output in a data table with a column for the variable name, the coefficient, and the lower and upper bounds of the 95% confidence intervals?
```{r}
form <- mpg ~ hp + wt + cyl
fit <- lm(form, data = mtcars)
out <- data.frame(names(coef(fit)),
                  coef(fit),
                  lmtest::coefci(fit))
names(out) <- c("term", "coef", "lower", "upper")
out <- tibble::as_tibble(out)
out
```

