---
title: "Regular Expressions"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Examples

You'll need to install the `{babynames}` package.

```{r}
library(tidyverse)
library(babynames)
head(babynames)
```

`babynames` is a dataset from the Social Security Administration in the US, and it documents all the names given to people registering for new social security numbers (so mostly babies) in a given year, from 1880 to 2015. To make it a little easier to work with (>1M rows), let’s use a summary version of this data: 

```{r}
NameList <- babynames %>%
	group_by(name, sex) %>%
	summarize(total=sum(n)) %>%
  arrange(desc(total))
```

In the past, we’ve used `filter(column=="string")` to find the values that we want to keep, but this gets more complicated if we want things that match only part of the string. For example, all names that have an “A” in them, or all names with “m” as the second letter.

Recall, if we use grepl alone, it will look at the strings that we give it, in this case, a column containing character strings, and assign each index [1] TRUE or FALSE logical values:

```
grepl("Margaret", babynames$name)
```

So to return the values that we want, we can use it together with filter(). 

Say the name contains “shine”, as in “sunshine”, or “moonshine” or “Shinelle”. 

```{r}
NameList %>%
	filter(grepl("shine", name, ignore.case = TRUE))
```

Why do you think ignore.case=TRUE might be an important argument here?

Say the name contains 3 or more vowels in a row.
```{r}
NameList %>%
	filter(grepl("[aeiou]{3,}", name, ignore.case = TRUE))
```

Some names have multiple different spellings: Sara vs Sarah
```{r}
NameList %>%
	filter(grepl("^[s][a][r][a]", name, ignore.case = TRUE))
```

Or 

```{r}
NameList %>%
	filter(grepl('^sara', name, ignore.case = TRUE))
```
Or 

```{r}
NameList %>%
	filter(grepl('^sara.$', name, ignore.case = TRUE))
```

Or 

```{r}
NameList %>%
	filter(grepl('^sar[ah]$', name, ignore.case = TRUE))
```

Or 

```{r}
NameList %>%
	filter(grepl('^sar[ah]{1,2}$', name, ignore.case = TRUE))
```

Or 

```{r}
NameList %>%
 	filter(grepl('^sar(ah|a)$', name, ignore.case = TRUE))
```

**Challenge: you try…!**

- The name contains 3 or more consonants in a row.
- The name contains “mn”.
- The first, third, and fifth letters are consonants.

```{r}
NameList %>%
	filter(grepl("[^aeiou]{3,}", name, ignore.case = TRUE))
```
```{r}
NameList %>%
	filter(grepl("mn", name, ignore.case = TRUE))
```

```{r}
NameList %>%
	filter(grepl("^[aeiou].[^aeiou].[^aeiou]", name, ignore.case = TRUE))
```

How often do boys versus girls names end in vowels?

```{r}
NameList %>%
	filter(grepl("[aeiou]$", name, ignore.case = TRUE)) %>%
	group_by(sex) %>%
	summarise(total=sum(total))
```

## Uing stringr

The `{stringr}` package can be a great help, if you need to manipulate, filter, or clean messy data. 

For example, take this text…

```{r}
Text <- "I am  terrible  at  getting my spacing correct       on my sentences   .      I really   need some    help."
```

Try this:

```{r}
text <- str_replace_all(Text, "\\s+", " ")
text
```

So much better! Something like this could be a handy way to get rid of, or control extra spaces that appear within text or in columns of data, that can cause inconsistencies in how the data is read.

Or imagine that you had some data that included special characters: 

```{r}
GDP <- c("$18.036 Trillion", "$16.832 Trillion", "$11.158 trillion")
```
But if we want to do math/conversions on this, we can’t have all the other info in there with the numbers. How would we get rid of it?

Either of will leave us with just the numbers:
```{r}
str_extract_all(GDP, "\\d+")
str_extract_all(GDP, "[:digit:]+")
```

But the results aren't so good.

Instead, clean up the vector to remove the $ and Trillion
```{r}
gdp <- str_replace(GDP, "\\s(T|t)rillion", "")
gdp <- str_replace(gdp, "\\$", "")
gdp
```

or

```{r}
str_replace_all(GDP, "\\$|%|\\s|(T|t)rillion|,", "")
```


