---
title: "Data Wrangling"
format: html
---

## Setup

```{r}
# install.packages("peacesciencer")
# install.packages("devtools")
# devtools::install_github("ryanburge/socsci")
library(tidyverse)
library(peacesciencer)
library(socsci)
```

## Make a dataset with `{peacesciencer}`

How it works:

1. Create a base dataset using a `create_*()` function.
2. Populate the dataset with variables using `add_*()` functions.

```{r}
create_stateyears() |>
  add_gml_mids() |>
  add_democracy() -> dt
```

## Start cleaning

First, let's make sure we drop any years from the data that don't have valid conflict observations. The MID dataset only runs up to 2010.

```{r}
dt |>
  filter(
    year <= 2010
  ) -> dt
```

Second, let's use the `polity2` column, which provides a measure of quality of democracy on a -10 to 10 scale, to create a categorical measure of regime type based on the Polity Project's recommendations. That is, we'll code countries with a score < -5 as autocracies, those with a score between -5 and 5 as anocracies, and those with a score > 5 as democracies. The `frcode()` function will let us do this.

```{r}
dt |>
  mutate(
    regime = frcode( 
      polity2 < -5 ~ "Autocracy",
        # if polity2 < 5, then return "Autocracy"
      polity2 %in% c(-5:5) | is.na(polity2) ~ "Anocracy",
        # if polity2 is bt -5 to 5, then return "Anocracy"
      polity2 > 5 ~ "Democracy"
        # if polity2 > 5, then return "Democracy"
    )
  ) -> dt
```


## Poke around

```{r}
dt |>
  select(statenme, year, regime, polity2) |>
  sample_n(10)
```


```{r}
dt |>
  filter(
    regime == "Democracy"
  ) |>
  count(statenme) |>
  arrange(-n)
```

```{r}
dt |>
  filter(regime == "Democracy") |>
  group_by(statenme) |>
  summarize(
    mid_init = sum(gmlmidonset_init)
  ) |>
  arrange(-mid_init)
```


## Make some data viz

First, let's just summarize the number of MIDs initiated over time.

```{r}
dt |>
  group_by(year) |>
  summarize(
    n_mids = sum(gmlmidonset_init)
  ) |>
  ggplot() +
  aes(
    x = year,
    y = n_mids
  ) +
  geom_line() +
  labs(
    x = NULL,
    y = NULL,
    title = "MIDs initiated over time",
    subtitle = "MIDs have become more frequent"
  )
```


What about the MID rate?

```{r}
dt |>
  group_by(year) |>
  summarize(
    mean_mids = mean(gmlmidonset_init)
  ) |>
  ggplot() +
  aes(
    x = year,
    y = mean_mids
  ) +
  geom_line() +
  labs(
    x = NULL,
    y = NULL,
    title = "MIDs initiated over time",
    subtitle = "MIDs have not really become more or less likely"
  )
```



Let's show the overall rate of starting MIDs by regime type.

```{r}
dt |>
  group_by(regime) |>
  summarize(
    mid_init_rate = mean(gmlmidonset_init)
  ) |>
  ggplot() +
  aes(
    x = regime,
    y = mid_init_rate
  ) +
  geom_col() +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "MID initiation rate among countries by regime",
    subtitle = "Anocracies are the least prone to starting fights"
  )
```

Let's show the rate over time by regime.

First, by count.

```{r}
dt |>
  group_by(year, regime) |>
  summarize(
    mid_init_rate = sum(gmlmidonset_init)
  ) |>
  ggplot() +
  aes(
    x = year,
    y = mid_init_rate,
    color = regime
  ) +
  geom_point(
    alpha = 0.4
  ) + 
  geom_smooth(
    method = "gam", 
    se = F
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "MID initiation count over time by regime",
    subtitle = "Democracies weren't always more peaceful",
    color = NULL
  ) +
  theme(
    legend.position = "bottom"
  )
```


Next, by rate.

```{r}
dt |>
  group_by(year, regime) |>
  summarize(
    mid_init_rate = mean(gmlmidonset_init)
  ) |>
  ggplot() +
  aes(
    x = year,
    y = mid_init_rate,
    color = regime
  ) +
  geom_point(
    alpha = 0.4
  ) + 
  geom_smooth(
    method = "gam", 
    se = F
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "MID initiation rate over time by regime",
    subtitle = "Democracies weren't always more peaceful",
    color = NULL
  ) +
  theme(
    legend.position = "bottom"
  )
```

Use `pivot_longer()` to show the count and rate in a small multiple:

```{r}
dt |>
  group_by(year) |>
  summarize(
    "MID Frequency" = sum(gmlmidonset_init),
    "MID Rate" = mean(gmlmidonset_init)
  ) |>
  pivot_longer(starts_with("MID")) |>
  ggplot() +
  aes(
    x = year,
    y = value
  ) +
  geom_point(color = "gray") + 
  geom_smooth(
    method = "gam", 
    se = F
  ) +
  facet_wrap(~ name, scales = "free_y") +
  labs(
    x = NULL,
    y = NULL,
    title = "MID initiation count and rate over time",
    subtitle = "Conflict is becoming more frequent but not more likely"
  ) 
```

