---
title: "Working with survey data"
format: html
---


## Setup

```{r packages}
library(tidyverse)
library(socsci)
theme_set(theme_bw())
```

```{r data}
wthh <- read_csv(
  "http://filesforprogress.org/datasets/wthh/DFP_WTHH_release.csv"
)
```


## Explore

```{r look at data}
wthh
```

link to codebook: https://www.filesforprogress.org/datasets/wthh/WTHH_Core_and_DFP_modules.pdf

## Recodes

Use `case_when()` to make recodes:

race What racial or ethnic group best describes you?
<1> White
<2> Black or African-American
<3> Hispanic or Latino
<4> Asian or Asian-American
<5> Native American
<8> Middle Eastern
<6> Mixed Race
<7> Other

```{r case_when}
wthh |>
  mutate(
    race_cat = case_when(
      race == 1 ~ "White",
      race == 2 ~ "Black",
      race == 3 ~ "Latino",
      race == 4 ~ "Asian",
      TRUE ~ "Other"
    )
  ) -> wthh
```

Use `frcode()` to make recodes:

```{r frcode}
wthh |>
  mutate(
    race_frcat = frcode(
      race == 1 ~ "White",
      race == 2 ~ "Black",
      race == 3 ~ "Latino",
      race == 4 ~ "Asian",
      TRUE ~ "Other"
    )
  ) -> wthh
```


```{r ct with race}
wthh |>
  ct(race_frcat)
```


Use `frcode()` with `pid7`:

```{r party lean}
wthh |>
  mutate(
    pid7_cat = frcode(
      pid7 == 1 ~ "Strong Dem",
      pid7 == 2 ~ "Dem",
      pid7 == 3 ~ "Lean Dem",
      pid7 == 4 ~ "Ind",
      pid7 == 5 ~ "Lean Rep",
      pid7 == 6 ~ "Rep",
      pid7 == 7 ~ "Strong Rep",
      TRUE ~ "Other"
    )
  ) -> wthh
```

Use `ct()` to summarize distribution of party lean:

```{r ct party lean}
wthh |>
  ct(pid7_cat)
```

Use `ifelse()` for binary categories:

```{r recode gender}
wthh |>
  mutate(
    gender_cat = ifelse(gender == 1, "Male", "Female"),
    gender_frcat = frcode(
      gender == 1 ~ "Male",
      TRUE ~ "Female"
    )
  ) -> wthh

wthh |>
  select(gender, gender_cat)
```

Use mathematical operations to recode age:

```{r recode age}
wthh |>
  mutate(
    age = 2018 - birthyr
  ) -> wthh
```


Making age a category:

```{r age category}
wthh <- wthh |>
  mutate(
    agecat = frcode(
      age < 25 ~ "18-24",
      between(age, 25, 29) ~ "25-29",
      between(age, 30, 39) ~ "30s",
      between(age, 40, 49) ~ "40s",
      between(age, 50, 59) ~ "50s",
      between(age, 60, 69) ~ "60s",
      age > 69 ~ "70+"
    )
  ) 
```

## Data viz

Make a bar plot of party lean:

```{r party lean viz}
wthh |>
  filter(pid7_cat != "Other") |>
  ct(pid7_cat) |>
  ggplot() +
  aes(x = pid7_cat, y = pct) +
  geom_col()
```

Make a bar plot of party lean weighted:

```{r weighted party lean viz}
wthh |>
  filter(pid7_cat != "Other") |>
  ct(pid7_cat, wt = weight_DFP) |>
  ggplot() +
  aes(x = pid7_cat, y = pct) +
  geom_col()
```

Make a bar plot that breaks party lean down by age category:

```{r party lean by age category}
wthh |>
  group_by(pid7_cat) |>
  mean_ci(age, wt = weight_DFP) |>
  ggplot() +
  aes(
    x = pid7_cat,
    y = mean,
    ymin = lower,
    ymax = upper
  ) +
  geom_pointrange()
```

