---
output: 
  stevetemplates::beamer:
    keep_tex: TRUE
    latex_engine: pdflatex # use xelatex here instead! I recommend it, but this is minimal reprex
    dev: cairo_pdf # I typically comment this out  if latex_engine: pdflatex
    slide_level: 3 # I prefer this, but I won't force it on you.
title: 'The Dangers of a Growing Anarchical Society (Revisited)'
subtitle: "DPR 190: War, Wealth, & World Politics"
author: Miles Williams
institute: Data for Political Research
# titlegraphic: "`r paste0(Sys.getenv('HOME'), '/Dropbox/clemson/watermarks/clemson-university-wordmark-orange-purple.png')`"
# scaletg: .9 # Optional, proportional (0, 1) to paper width. Defaults to .3
fontsize: 12pt
make149: TRUE
# mainfont: "Open Sans" # Try out some font options if xelatex
# titlefont: "Titillium Web" # Try out some font options if xelatex
# specify color as six-digit RGB (no pound sign)
# primarycolor: "990000" 
# secondarycolor: "000099"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=F)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')
knitr::opts_chunk$set(echo = F, message = F, warning = F,
                      out.width = "95%",
                      dpi = 500,
                      fig.height = 4, 
                      fig.width = 6)

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
library(tidyverse)
library(coolorrr)
set_palette()
set_theme()
library(stevemisc)
library(peacesciencer)
library(here)
library(socsci)
library(priceR)

## attach the new and shiny MIE data!
mie_data <- read_csv(
  here("DPR 190", "Data", "events", "mie-1.0.csv")
)

## aggregate to the country-year level:
mie_data |>
  group_by(ccode1, styear) |>
  summarize(
    endyear = max(endyear),
    event = 1,
    num_events = n(),
    mindeaths = sum(
      fatalmin1 + fatalmin2
    ),
    maxdeaths = sum(
      fatalmax1 + fatalmax2
    ),
    .groups = "drop"
  ) -> mie_country_year

## add some correlates from {peacesciencer}
create_stateyears(subset_years = 1816:2014) |>
  add_democracy() |>
  add_sdp_gdp() |>
  add_nmc() |>
  add_cow_majors() |>
  add_igos() -> cov_data

## make some new variables:
## - n: num. of countries in the world in a given year
## - country_age: indicator of how many years a country is in the data
## - new_country: indicator of a new country
cov_data |>
  arrange(year) |>
  group_by(year) |>
  mutate(
    n = n()
  ) |>
  group_by(ccode) |>
  mutate(
    country_age = 1:n(),
    new_country = (country_age==1)+0
  ) |>
  ungroup() -> cov_data

## merge the datasets
left_join(
  cov_data, mie_country_year |> 
    rename(ccode = ccode1, year = styear),
  by = c("ccode", "year")
) -> Data

## deal with missingness
Data |>
  mutate(
    across(
      endyear:maxdeaths,
      ~ replace_na(.x, 0)
    )
  ) -> Data
```

# Why revisit the growing pains hypothesis?

### Why revisit the growing pains hypothesis?

- Confounding factors.
- Are these new countries even directly responsible for the spike in conflicts?
- Are there other mechanisms?

# Looking at the data

### Looking at the data

```{r}
Data |>
  group_by(year) |>
  summarise(
    n = sum(num_events)
  ) |>
  ggplot() +
  aes(year, n) +
  geom_line(size = 1) +
  scale_x_continuous(
    n.breaks = 20
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    x = "Year",
    y = "Number of Militarized Events",
    title = "Militarized Interstate Disputes, 1816-2014"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1
    ),
    panel.grid.major.x = element_blank()
  )
```

### Looking at the data

```{r}
Data |>
  group_by(year) |>
  summarize(
    events = sum(num_events),
    n = n(),
    "No. of New Countries" = sum(new_country),
    .groups = "drop"
  ) |>
  ungroup() |>
  mutate(
    "Change in No. of Countries" = c(0, diff(n)),
    "No. of New Countries" = 
      ifelse(year==1816, 0, `No. of New Countries`)
  ) |>
  select(-n) |>
  pivot_longer(
    3:4
  ) |>
  ggplot() +
  aes(value, events) +
  facet_wrap(~ name, scales = "free_x",
             strip.position = "bottom") +
  geom_point(
    alpha = 0.5
  ) +
  stat_smooth(
    method = "glm",
    method.args = list(family = quasipoisson),
    se = F
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  theme(
    strip.placement = "outside",
    strip.background = element_blank()
  ) +
  labs(
    x = NULL,
    y = "No. of Militarized Events",
    title = "MIEs per Changes in the No. of Countries, 1816-2014"
  )
```

### Looking at the data

```{r}
Data |>
  group_by(event) |>
  ct(new_country) |>
  filter(event == 1) |>
  mutate(
    new_country = ifelse(new_country, "New Country", "Not New")
  ) |>
  ggplot() +
  aes(new_country, pct) +
  geom_col() +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = NULL,
    y = "% of Militarized Events",
    title = "Are new countries involved in MIEs?"
  )
```

### Looking at the data

```{r}
Data |>
  ggplot() +
  aes(country_age, event) +
  stat_smooth(
    aes(group = ccode),
    method = "glm",
    method.args = list(family = binomial),
    se = F,
    color = "lightgray",
    linewidth = 0.05
  ) +
  stat_smooth(
    color = "black",
    method = "glm",
    method.args = list(family = binomial),
    se = F
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = "Country Age",
    y = "(%) Likelihood of MIE",
    title = "Are young countries more violent?"
  ) -> p
p
```

### Looking at the data

```{r}
newData <- Data |>
      group_split(ccode) |>
      map_dfr(
        ~ {
          d <- .x
          fit <- glm(event ~ country_age, 
                     data = d,
                     family = binomial)
          d$pos <- ifelse(
            coef(fit)[2] > 0,
            1, 0
          )
          d
        }
      ) 
frac_neg <- newData |>
  group_by(ccode) |>
  count(pos) |>
  ungroup() |>
  summarize(neg = 1-mean(pos))
p + 
  stat_smooth(
    data = newData |> filter(pos == 0),
    aes(group = ccode),
    method = "glm",
    method.args = list(family = binomial),
    se = F,
    color = "royalblue",
    linewidth = 0.1
  ) +
  annotate(
    "text",
    x = 175,
    y = 0.8,
    label = paste0(
      round(frac_neg$neg*100, 2), "%"
    ),
    color = "royalblue"
  )
```

### Looking at the data

```{r}
Data |>
  group_by(new_country, year) |>
  summarize(
    "Surplus Domestic Product (log)" = mean(sdpest, na.rm=T),
    "IGOs" = mean(sum_igo_anytype, na.rm=T),
    "Democracy" = mean(v2x_polyarchy, na.rm=T)
  ) |>
  pivot_longer(
    3:5
  ) |>
  ggplot() +
  aes(year, value, color = ifelse(new_country == 1, "Yes", "No")) +
  geom_point(
    alpha = 0.4
  ) +
  geom_smooth(
    linesize = 0.75,
    se = F
  ) +
  facet_wrap(~ name, scales = "free_y",
             strip.position = "left") +
  theme(
    strip.background = element_blank(),
    strip.placement = "outside"
  ) +
  labs(
    x = "Year",
    y = NULL,
    color = "New Country? ",
    title = "What makes new countries unique?"
  ) +
  ggpal() +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1
    )
  )
  
```

### Looking at the data

```{r}
bop <- function(x, y) {
  # x <- replace_na(x, 0)
  # y <- replace_na(y, 0)
  x <- x / y
  p <- x / sum(x, na.rm=T)
  n <- sum(!is.na(x))
  ssp <-  sum(p^2)
  con <- sqrt((ssp - 1/n)/(1 - 1/n))
  1 - con
}
Data |>
  group_by(year) |>
  summarize(
    n = n(),
    bop = bop(cinc, 1)
  ) |>
  ungroup() |>
  mutate(
    change_n = c(0, diff(n)),
    change_bop = c(0, diff(bop))
  ) |>
  ggplot() +
  aes(change_n, bop) +
  geom_point() +
  stat_smooth(
    method = "glm",
    method.args = list(family = quasibinomial),
    se = F
  ) +
  labs(
    x = expression(Delta*" No. of Countries"),
    y = "Balance of Power",
    title = "Change in the system and the balance of power"
  ) +
  scale_y_continuous(
    labels = scales::percent
  )
```

### Looking at the data

```{r}
Data |>
  group_by(year) |>
  summarize(
    n = n(),
    bop = bop(cinc[cowmaj==1], 1)
  ) |>
  ungroup() |>
  mutate(
    change_n = c(0, diff(n)),
    change_bop = c(0, diff(bop))
  ) |>
  ggplot() +
  aes(change_n, bop) +
  geom_point() +
  stat_smooth(
    method = "glm",
    method.args = list(family = quasibinomial),
    se = F
  ) +
  labs(
    x = expression(Delta*" No. of Countries"),
    y = "Balance of Power",
    title = "Change in the system and the balance of power"
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  annotate(
    "text",
    x = 12.5, y = .55,
    label = "(BoP of Major Powers)",
    color = "blue",
    size = 3
  )
```

### Looking at the data

```{r}
Data |>
  group_by(year, cowmaj) |>
  ct(event) |>
  filter(event == 1) |>
  ggplot() +
  aes(year, pct, color = ifelse(cowmaj==1, "Major Power", "Minor Power")) +
  geom_point(
    alpha = 0.4
  ) +
  geom_smooth(
    se = F
  ) +
  ggpal() +
  labs(
    x = "Year",
    y = "Proportion of Events",
    color = NULL,
    title = "Who drives militarized events?"
  ) +
  scale_x_continuous(
    n.breaks = 20
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1
    )
  )
```

### Looking at the data

```{r}
wm <- map_data("world")
wm <- wm |>
  mutate(
    cname = countrycode::countrycode(
      region, "country.name", "country.name"
    )
  )
mp <- Data |>
  group_by(statenme) |>
  summarize(
    cowmaj = any(cowmaj==1)+0
  ) |>
  ungroup() |>
  mutate(
    cname = countrycode::countrycode(
      statenme, "country.name", "country.name"
    )
  )

wm_dt <- left_join(wm, mp)
ggplot(wm_dt) +
  aes(long, lat, group = group) +
  geom_polygon(
    fill = "white",
    color = "black", 
    linewidth = 0.05
  ) +
  geom_polygon(
    data = wm_dt |> filter(cowmaj == 1),
    fill = "royalblue",
    color = "black",
    linewidth = .1
  ) +
  coord_fixed() +
  scale_x_continuous(
    breaks = NULL
  ) +
  scale_y_continuous(
    breaks = NULL
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Who are the major powers?"
  ) +
  coord_fixed()
```

# Conclusions

### Conclusions

\begin{center}
\textit{What do you all think?}
\end{center}

### Conclusions

- Maybe it's the change in the balance of power?
- As major powers try to manage the "Anarchical Society" new countries upset the balance.
- This drives conflict, not involving the new countries that enter the world, but ***among major powers*** themselves.

### Conclusions

\begin{center}
\textit{Maybe. Maybe not.}
\end{center}

# Elements of order

### Elements of order

- Respect for sovereignty
- Balance of power
- \textcolor{gray}{Alliances}
- \textcolor{gray}{Democracy}
- \textcolor{gray}{Economic interdependence}
- \textcolor{gray}{Global governance}
- \textcolor{gray}{International law}

### Elements of order

- Respect for sovereignty
- Balance of power
- \textcolor{blue}{Alliances}
- \textcolor{blue}{Democracy}
- \textcolor{blue}{Economic interdependence}
- \textcolor{blue}{Global governance}
- \textcolor{blue}{International law}

# Next time

### Next time

-  David Lake, "Escape from the State
of Nature"
