---
output: 
  stevetemplates::beamer:
    keep_tex: TRUE
    latex_engine: pdflatex # use xelatex here instead! I recommend it, but this is minimal reprex
    dev: cairo_pdf # I typically comment this out  if latex_engine: pdflatex
    slide_level: 3 # I prefer this, but I won't force it on you.
title: Globalization
subtitle: DPR 190
author: Miles D. Williams
institute: Data for Political Research
# titlegraphic: "`r paste0(Sys.getenv('HOME'), '/Dropbox/clemson/watermarks/clemson-university-wordmark-orange-purple.png')`"
# scaletg: .9 # Optional, proportional (0, 1) to paper width. Defaults to .3
fontsize: 12pt
make149: TRUE
# mainfont: "Open Sans" # Try out some font options if xelatex
# titlefont: "Titillium Web" # Try out some font options if xelatex
# specify color as six-digit RGB (no pound sign)
# primarycolor: "990000" 
# secondarycolor: "000099"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=F,
                      message = F,
                      warning = F,
                      echo = F,
                      #dpi = 500,
                      out.width = "90%",
                      fig.height = 4.5,
                      fig.width = 7.5)
library(tidyverse)
library(stevemisc)
library(peacesciencer)
library(priceR)
library(coolorrr)
set_theme()
set_palette()

data_exists <- file.exists(
  here::here("DPR 190", "Data", "global_trade.csv")
)
if(data_exists) {
  Data <- read_csv(
    here::here("DPR 190", "Data", "global_trade.csv")
  )
} else {
  Data <- create_dyadyears() |>
    add_cow_trade() |>
    add_sdp_gdp()
  
  Data |> 
  mutate(
    ## add regional indicators
    region1 = countrycode::countrycode(
      ccode1, "cown", "region"
    ),
    region2 = countrycode::countrycode(
      ccode2, "cown", "region"
    ),
    same_region = ifelse(
      region1 == region2,
      yes = "Yes", no = "No"
    )
  ) -> Data

  ## save it
  write_csv(
    Data,
    here::here("DPR 190", "Data", "global_trade.csv")
  )
}

## set up for adjusting prices
country <- "US"
inflation_dataframe <- retrieve_inflation_data(country)
countries_dataframe <- show_countries()
  
```


# Globalization

### Globalization

\begin{centering}

What is \textit{globalization}?

\end{centering}

### Globalization

\begin{centering}

\textbf{Globalization} usually is a term used to describe increasing \textit{cross-border} interconnections across the globe.

\end{centering}

### Globalization

\begin{centering}

How can we \textit{measure} globalization?

\end{centering}

### Globalization

\begin{centering}

A common way to measure globalization is \textit{trade}.

\end{centering}

# When did globalization begin?

### When did globalization begin?

```{r}
Data |>
  group_by(year) |>
  summarize(
    trade = sum(flow1 + flow2, na.rm = T) * 1e06,
    gdp   = sum(exp(wbgdp2011est1) + exp(wbgdp2011est2), na.rm = T),
    .groups = "drop"
  ) |>
  filter(trade > 0) |>
  # mutate(
  #   ## adjust the price of trade to 2011 USD
  #   trade = afi(
  #     trade, year, country, to_date = 2011,
  #     inflation_dataframe, countries_dataframe
  #   ),
  #   trade_per_gdp = trade / gdp
  # ) |>
  ggplot() +
  aes(x = year, y = trade / 1e09) +
  geom_point(color = "gray") +
  geom_line(
    aes(y = zoo::rollmean(trade / 1e09, 5, na.pad = T)),
    size = 1,
    color = "navy"
  ) +
  scale_y_continuous(
    labels = scales::dollar
  ) +
  scale_x_continuous(
    breaks = seq(1850, 2020, by = 5)
  ) +
  labs(
    x = "Year",
    y = "Global Trade\n(Billions)",
    title = "When did globalization begin?",
    subtitle = "Total trade per year in current prices"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1
    )
  )
```

### When did globalization begin?

```{r}
Data |>
  group_by(year) |>
  summarize(
    trade = sum(flow1 + flow2, na.rm = T) * 1e06,
    gdp   = sum(exp(wbgdp2011est1) + exp(wbgdp2011est2), na.rm = T),
    .groups = "drop"
  ) |>
  filter(trade > 0) |>
  mutate(
    ## adjust the price of trade to 2011 USD
    trade = afi(
      trade, year, country, to_date = 2011,
      inflation_dataframe, countries_dataframe
    ),
    trade_per_gdp = trade / gdp
  ) |>
  ggplot() +
  aes(x = year, y = trade / 1e09) +
  geom_point(color = "gray") +
  geom_line(
    aes(y = zoo::rollmean(trade / 1e09, 5, na.pad = T)),
    size = 1,
    color = "navy"
  ) +
  scale_y_continuous(
    labels = scales::dollar
  ) +
  scale_x_continuous(
    breaks = seq(1850, 2020, by = 5)
  ) +
  labs(
    x = "Year",
    y = "Global Trade\n(Billions)",
    title = "When did globalization begin?",
    subtitle = "Total trade per year in 2011 prices"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1
    )
  )
```


### When did globalization begin?

```{r}
Data |>
  group_by(year) |>
  summarize(
    trade = sum(flow1 + flow2, na.rm = T) * 1e06,
    gdp   = sum(exp(wbgdp2011est1) + exp(wbgdp2011est2), na.rm = T),
    .groups = "drop"
  ) |>
  filter(trade > 0) |>
  mutate(
    ## adjust the price of trade to 2011 USD
    trade = afi(
      trade, year, country, to_date = 2011,
      inflation_dataframe, countries_dataframe
    ),
    trade_per_gdp = trade / gdp
  ) |>
  ggplot() +
  aes(x = year, y = trade_per_gdp) +
  geom_point(color = "gray") +
  geom_line(
    aes(y = zoo::rollmean(trade_per_gdp, 5, na.pad = T)),
    size = 1,
    color = "navy"
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  scale_x_continuous(
    breaks = seq(1850, 2020, by = 5)
  ) +
  labs(
    x = "Year",
    y = "Global Trade per GDP (%)",
    title = "When did globalization begin?",
    subtitle = "Total trade per real GDP in 2011 USD"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1
    )
  )
```

# How global is globalization?

### How global is globalization?

- A key part of globalization is the idea that growing cross-border interconnections are *global*.
- Does this seem right to you?
- How could we check?

### How global is globalization?

```{r}
Data |>
  group_by(year, same_region) |>
  summarize(
    trade = sum(flow1 + flow2, na.rm = T) * 1e06,
    gdp   = sum(exp(wbgdp2011est1) + exp(wbgdp2011est2), na.rm = T),
    .groups = "drop"
  ) |>
  filter(trade > 0) |>
  drop_na() |>
  group_by(year) |>
  mutate(
    gdp = sum(gdp)
  ) |>
  ungroup() |>
  mutate(
    ## adjust the price of trade to 2011 USD
    trade = afi(
      trade, year, country, to_date = 2011,
      inflation_dataframe, countries_dataframe
    ),
    trade_per_gdp = trade / gdp
  ) |>
  group_by(same_region) |>
  mutate(
    roll_avg = zoo::rollmean(
      trade_per_gdp, 5, na.pad = T
    )
  ) |>
  ungroup() |>
  ggplot() +
  aes(x = year, y = trade_per_gdp, color = same_region) +
  geom_point(alpha = 0.3) +
  geom_line(
    aes(y = roll_avg),
    size = 0.75
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  scale_x_continuous(
    breaks = seq(1850, 2020, by = 5)
  ) +
  labs(
    x = "Year",
    y = "Global Trade per GDP (%)",
    title = "How global is globalization?",
    subtitle = "Total trade per real GDP in 2011 USD",
    color = "Same Region?\n"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1
    )
  ) +
  ggpal()
```

### How global is globalization?

```{r}
Data |>
  group_by(year, same_region) |>
  summarize(
    trade = sum(flow1 + flow2, na.rm = T) * 1e06,
    gdp   = sum(exp(wbgdp2011est1) + exp(wbgdp2011est2), na.rm = T),
    .groups = "drop"
  ) |>
  filter(trade > 0) |>
  drop_na() |>
  group_by(year) |>
  mutate(
    trade_per_region = trade / sum(trade)
  ) |>
  group_by(same_region) |>
  mutate(
    roll_avg = zoo::rollmean(
      trade_per_region, 5, na.pad = T
    )
  ) |>
  ungroup() |>
  ggplot() +
  aes(x = year, y = trade_per_region, fill = same_region) +
  geom_area(
    show.legend = F
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  scale_x_continuous(
    breaks = seq(1870, 2015, by = 5)
  ) +
  labs(
    x = "Year",
    y = "(%) of Global Trade",
    title = "How global is globalization?",
    subtitle = "Share of trade within versus across regions"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1
    ),
    panel.grid.major.x = element_blank()
  ) +
  ggpal(aes = "fill") +
  annotate(
    "text",
    x = 1885, y = .3,
    label = "Within Region\nTrade"
  ) +
  annotate(
    "text",
    x = 1980, y = .7,
    label = "Bewteen Region\nTrade",
    color = "white"
  ) +
  geom_hline(
    yintercept = c(.5, .5),
    color = c("black", "white"),
    linewidth = c(1, 0.75)
  ) 
```


### How global is globalization?

```{r}
Data |>
  group_by(year, same_region) |>
  summarize(
    trade = sum(flow1 + flow2, na.rm = T) * 1e06,
    gdp   = sum(exp(wbgdp2011est1) + exp(wbgdp2011est2), na.rm = T),
    n     = n(),
    .groups = "drop"
  ) |>
  filter(trade > 0) |>
  drop_na() |>
  group_by(year) |>
  mutate(
    gdp = sum(gdp)
  ) |>
  ungroup() |>
  mutate(
    ## adjust the price of trade to 2011 USD
    trade = afi(
      trade, year, country, to_date = 2011,
      inflation_dataframe, countries_dataframe
    ),
    trade_per_gdp = (trade) / n
  ) |>
  group_by(same_region) |>
  mutate(
    roll_avg = zoo::rollmean(
      trade_per_gdp / 1e09, 5, na.pad = T
    )
  ) |>
  ungroup() |>
  ggplot() +
  aes(x = year, y = trade_per_gdp / 1e09, color = same_region) +
  geom_point(alpha = 0.3) +
  geom_line(
    aes(y = roll_avg),
    size = 0.75
  ) +
  scale_y_continuous(
    labels = scales::dollar
  ) +
  scale_x_continuous(
    breaks = seq(1850, 2020, by = 5)
  ) +
  labs(
    x = "Year",
    y = "Average Trade\n(Billions)",
    title = "How global is globalization?",
    subtitle = "Average trade within and between regions in 2011 USD",
    color = "Same Region?\n"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1
    )
  ) +
  ggpal()
```

### How global is globalization?

- Cross-border connections have been growing.
- But these cross-border connections remain disproportionately *regional* rather than *global*.
- Do you think this conclusion would be the same if we tried to measure globalization a different way?

# Next Time

### Next Time

Henry Farrell and Abraham Newman, "Chained to Globalization" and "Will the Coronavirus End Globalization as we Know It?"
